---
title: "eDNA Course - Diversity analysis"
author: "Patrick Nichols"
date: "2026-07-02"
output: html_document
---

## 1. Setup

Session -> Set working directory -> To source file location...

Carefully update the user-defined settings for each run, making sure column names and index positions match.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(iNEXT)
library(vegan)
library(ggpubr)
library(indicspecies)
library(microDecon)
library(purrr)
library(tibble)
library(readxl)
set.seed(13)
numperm=9999 #Permutation tests

paths=list(
  data_raw      ="data/raw",
  data_processed="data/processed",
  results_rds   ="results/rds",
  results_tables="results/tables",
  results_figs  ="results/figures"
)

walk(paths, dir.create, showWarnings = FALSE, recursive = TRUE)

cols.method=c("#031d44", "#1a659e", "#d5896f", "#dab785", "#70a288", "snow3", "black")

#### RUN SETTINGS #######################################################
num.blanks=14 # how many columns of blanks/negative control samples?
SAMPLE="Sample_ID" # name of metadata column with sample IDs
OTU_ID="OTU_ID" # name of OTU ID column in OTU table
GROUPING="TREATMENT" # name of grouping variable column in metadata
TAXONOMICRANK= "Genus" # choose taxonomic rank for indicator taxon analysis
distance.method="bray" #choose distance method for community dissimilarity ordination
#########################################################################

#### READ IN OTU TABLE ##################################################
otutab.raw=read_excel(file.choose())  # raw OTU table, with Unique ID/OTU_ID as first column, apscale taxonomic information
otutab.predecon=otutab.raw[,c(1:129)] # select only OTU ID and numerical read count columns
otutab.predecon[, -1] <- lapply(otutab.predecon[, -1], function(x) {
  if(is.character(x)) as.numeric(trimws(x)) else x
})
# Remove zero-sum rows (keep ID column)
otutab.predecon <- otutab.predecon[rowSums(otutab.predecon[, -1]) != 0, ]
# Remove zero-sum columns (keep ID column)
otutab.predecon <- otutab.predecon[, c(1, which(colSums(otutab.predecon[, -1]) != 0) + 1)]
apscale.info=otutab.raw[,c(1,130:ncol(otutab.raw))] # select taxonomic info and other columns from apscale
#########################################################################

#### READ IN METADATA FILE #############################################
metadata=read_excel(file.choose())
metadata=subset(metadata, Run != "1")
metadata=subset(metadata, SITE=="Station2" | SITE=="Station3")
metadata=subset(metadata, SITE != "Control")
metadata=subset(metadata, Sample_ID != "Teleo2_run4_barcode68") # failed samples?
metadata=subset(metadata, Sample_ID != "Teleo2_run4_barcode65") # failed samples?
metadata$QUBIT <- as.numeric(gsub(",", ".", trimws(metadata$QUBIT)))
#######################################################################
```

### Decontamination

#### MicroDecon

```{r}
otutab.predecon <- as.data.frame(otutab.predecon)

# Run decontamination
microdecon.decontaminated <- decon(otutab.predecon,numb.blanks = num.blanks, numb.ind = (ncol(otutab.predecon)-num.blanks-1),taxa = FALSE,thresh = 1,prop.thresh = 0, runs = 1)

reads.removed=microdecon.decontaminated$reads.removed
decontaminated.table <- microdecon.decontaminated$decon.table
decontaminated.table$Mean.blank <- NULL

reads.removed=microdecon.decontaminated$reads.removed
reads.removed=merge(reads.removed, apscale.info, by=OTU_ID)

decontaminated.table.metadata=merge(decontaminated.table, apscale.info, by=OTU_ID) # merge taxonomic info back into decontaminated file

write.csv(decontaminated.table.metadata, file=file.path(paths$data_processed, "decontaminated_data_teleo.csv"), row.names = F) ## save to PC
```


### Data cleaning and prepping

This block removes human-related taxa (genus Homo) from the dataset and renames the rows using organism IDs. It then isolates the taxonomic classification columns into a separate table called 'taxa'.

```{r}
otutab=decontaminated.table.metadata

# Filter unwanted taxa
data=subset(otutab, Order !="Primates") # if taxonomy information is present in OTU table
data=subset(data, Kingdom !="NoMatch") # if taxonomy information is present in OTU table
data=subset(data, Genus !="Cranioleuca") # if taxonomy information is present in OTU table


row.names(data)=data[[OTU_ID]]
final=data[,2:115] # Update me
rownames(final)=data[["OTU_ID"]]

final.t=as.data.frame(t(final))
final.t[[SAMPLE]]=rownames(final.t)
final.t.meta=merge(final.t, metadata, by=SAMPLE)

final.t=final.t.meta[,2:(nrow(data)+1)]
rownames(final.t)=final.t.meta[[SAMPLE]]

final=as.data.frame(t(final.t))
final <- final[, colSums(final) != 0] # remove zero sum columns
final <- final[rowSums(final) != 0, ] # remove zero sum rows
#final <- final[, colSums(final != 0) > 1] # remove samples with only 1 OTU
ntax=nrow(final)
nsample=ncol(final)
final[[OTU_ID]]=rownames(final)

final=merge(final, apscale.info, by=OTU_ID)
rownames(final)=final[[OTU_ID]]

# Taxonomic metadata only
taxa=final[, (nsample+2):ncol(final)]
rownames(taxa)=final[[OTU_ID]]
taxa[[OTU_ID]]=rownames(taxa)
```


### Transformations

This code prepares three versions of your OTU table (raw counts, Hellinger-transformed, and eDNA index), then lets you choose one for downstream analysis. The final steps transpose the table so taxa become columns and samples become rows, then merges it with site metadata for statistical testing.

#### Raw Reads (no transformation)

```{r}
# Raw reads OTU table only, no metadata
table.raw=final[, 2:(nsample+1)]
final_reads=final # make a copy for raw reads (will not undergo transformations)
table.reads=final_reads[,2:(nsample+1)]

sum(table.reads)

table.reads.t <- as.data.frame(t(table.reads))  # Flip OTU table so taxa as columns

table.reads.t[[SAMPLE]] <- row.names(table.reads.t) # add sample column to merge with metadata
metatable.reps.reads=merge(table.reads.t, metadata, by=SAMPLE) ##OTU table with site metadata correctly formatted for downstream analyses

table.reads.t[[SAMPLE]]=NULL #remove non-numeric columns for downstream analyses
```

#### Presence-Absence (Binary) 

```{r}
table_pa=decostand(table.reads, method = "pa")
table_pa[[OTU_ID]]=rownames(table_pa)
final_pa=merge(table_pa, taxa, by=OTU_ID)
rownames(final_pa)=final_pa[[OTU_ID]]
table_pa[[OTU_ID]]=NULL
table_pa.t=t(table_pa)
```

#### Hellinger 

```{r}
table_hel=decostand(table.reads, method = "hellinger")
table_hel[[OTU_ID]]=rownames(table_hel)
final_hel=merge(table_hel, taxa, by=OTU_ID)
rownames(final_hel)=final_hel[[OTU_ID]]
table_hel[[OTU_ID]]=NULL
table_hel.t=t(table_hel)
```

#### eDNA Index

```{r}
table.raw <- as.matrix(table.reads)  # Compute relative abundance per sample (column-wise)
relative_abundance <- sweep(table.raw, 2, colSums(table.raw), "/")  # Find the maximum observed abundance for each OTU across all samples (row-wise)
max_relative_abundance <- apply(relative_abundance, 1, max, na.rm = TRUE)  # Compute the eDNA index by dividing each OTU's abundance by its max
table_index <- as.data.frame(sweep(relative_abundance, 1, max_relative_abundance, "/"))  # Add OTU IDs back to the table

table_index[[OTU_ID]] <- rownames(table_index) # Merge with taxa metadata (assuming 'taxa' has OTU_ID as key)
final_index <- merge(table_index, taxa, by = OTU_ID) # Ensure row names match OTU_ID
rownames(final_index) <- final_index[[OTU_ID]] 
table_index[[OTU_ID]] <- NULL
table_index.t=t(table_index)
```

#### Prepare community data

```{r}
####CHOOSE MATRIX###################################################
community.table.t=table_index.t #transposed otu table version (OPTIONS: table_index.t, table_hel.t, table.reads.t. table_pa.t)
community.table=table_index #otu table only (OPTIONS: table_index, table_hel, table.reads, table_pa, etc.)
community.data=final_index #version with metadata (OPTIONS: final_index, final_hel, final_reads, final_pa, etc.)
###################################################################
#community.table.t=table.reads.t #transposed otu table version (OPTIONS: table_index.t, table_hel.t, table.reads.t. table_pa.t)
#community.table=table.reads #otu table only (OPTIONS: table_index, table_hel, table.reads, table_pa, etc.)
#community.data=final_reads #version with metadata (OPTIONS: final_index, final_hel, final_reads, final_pa, etc.)
```

#### Merge filtered metadata

```{r}
table.t <- as.data.frame(t(as.matrix(community.table)))

ntax=ncol(table.t)
nsample=nrow(table.t)

table.t[[SAMPLE]]=row.names(table.t) ##OTU table only

# Remove leading zeros from SAMPLE column
metatable.reps=merge(table.t, metadata, by=SAMPLE) ##OTU table with site metadata
table.t[[SAMPLE]]=NULL
```



####Melt data long format

```{r}
data_long <- as.data.frame(community.table.t) %>%
  rownames_to_column(var = SAMPLE) %>%
  pivot_longer(
    cols = -SAMPLE,  # Keep SampleID as identifier
    names_to = OTU_ID,
    values_to = "abundance"
  )

# Remove zero abundances (optional)
data_long <- data_long %>%
  filter(abundance > 0)

data_long=merge(data_long, metadata, by=SAMPLE)
data_long=merge(data_long, apscale.info, by=OTU_ID)
```

## 2. Treatment Summaries

### Total DNA Concentration

```{r}
# Create a distinct sample-level dataset for QUBIT values
qubit_by_sample <- data_long %>%
  distinct(!!sym(SAMPLE), .keep_all = TRUE) 

# OR, if you want to be more explicit:
qubit_by_sample <- data_long %>%
  dplyr::select(SAMPLE, Run, TREATMENT, FILTRATION, EXTRACTION, QUBIT, SITE) %>%
  distinct()

qubit_by_sample %>% 
  group_by(Run) %>% 
  summarise(total_qubit = sum(as.numeric(trimws(QUBIT)), na.rm = TRUE))

# By Treatment - QUBIT summary (corrected)
qubit_by_sample %>% 
  group_by(TREATMENT) %>% 
  summarise(total_qubit = sum(QUBIT, na.rm = TRUE))

# By Filtration - QUBIT summary (corrected)
qubit_by_sample %>% 
  group_by(FILTRATION) %>% 
  summarise(total_qubit = sum(QUBIT, na.rm = TRUE))

# By Extraction - QUBIT summary (corrected)
qubit_by_sample %>% 
  group_by(EXTRACTION) %>% 
  summarise(total_qubit = sum(QUBIT, na.rm = TRUE))

# By all four factors - QUBIT summary (corrected)
qubit_by_sample %>% 
  group_by(TREATMENT, FILTRATION, EXTRACTION, Run) %>% 
  summarise(
    total_qubit = sum(QUBIT, na.rm = TRUE),
    mean_qubit = mean(QUBIT, na.rm = TRUE),
    n = n()
  ) %>% 
  arrange(desc(total_qubit))

# Sum QUBIT across treatments and runs (corrected)
qubit_by_sample %>%
  group_by(TREATMENT, Run) %>%
  summarise(total_qubit = sum(QUBIT, na.rm = TRUE)) %>%
  pivot_wider(names_from = Run, values_from = total_qubit, values_fill = 0)

# Overall summary by Treatment - QUBIT (corrected)
qubit_by_sample %>%
  group_by(TREATMENT) %>%
  summarise(
    total = sum(QUBIT, na.rm = TRUE),
    mean = mean(QUBIT, na.rm = TRUE),
    median = median(QUBIT, na.rm = TRUE),
    min = min(QUBIT, na.rm = TRUE),
    max = max(QUBIT, na.rm = TRUE),
    sd = sd(QUBIT, na.rm = TRUE),
    n = n()
  )

# 1. By TREATMENT - Boxplot with jitter (QUBIT) - CORRECTED
ggplot(qubit_by_sample, aes(x = TREATMENT, y = QUBIT, fill = TREATMENT)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.8) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 2, fill = "white") +
  labs(
    title = "QUBIT Distribution by Treatment",
    x = "Treatment",
    y = "QUBIT (ng/uL)"
  ) +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) + 
  scale_color_manual(values = cols.method) + 
  scale_fill_manual(values = cols.method)

# 2. By Run - Boxplot (QUBIT) - CORRECTED
ggplot(qubit_by_sample, aes(x = Run, y = QUBIT, fill = Run)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.8) +
  labs(
    title = "QUBIT Distribution by Run",
    x = "Run",
    y = "QUBIT (ng/uL)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# 5. Treatment × Run interaction plot (QUBIT) - CORRECTED
treatment_run_summary <- qubit_by_sample %>%
  group_by(TREATMENT, Run) %>%
  summarise(
    mean_qubit = mean(QUBIT, na.rm = TRUE),
    sd_qubit = sd(QUBIT, na.rm = TRUE),
    n = n()
  ) %>%
  mutate(se_qubit = sd_qubit / sqrt(n)) 

ggplot(treatment_run_summary, aes(x = Run, y = mean_qubit, color = TREATMENT, group = TREATMENT)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_qubit - se_qubit, ymax = mean_qubit + se_qubit),
                width = 0.2, alpha = 0.5) +
  labs(
    title = "Treatment × Run Interaction (QUBIT)",
    x = "Run",
    y = "Mean QUBIT (ng/uL)",
    color = "Treatment"
  ) +
  theme_minimal() + 
  scale_color_manual(values = cols.method) + 
  scale_fill_manual(values = cols.method)

# 7. Faceted by Treatment: QUBIT by Run - CORRECTED
ggplot(qubit_by_sample, aes(x = Run, y = QUBIT, fill = Run)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~TREATMENT, scales = "free_y") +
  labs(
    title = "QUBIT by Run within Each Treatment",
    x = "Run",
    y = "QUBIT (ng/uL)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# 8. Faceted by Run: QUBIT by Treatment - CORRECTED
ggplot(qubit_by_sample, aes(x = TREATMENT, y = QUBIT, fill = TREATMENT)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~Run) +
  labs(
    title = "QUBIT by Treatment within Each Run",
    x = "Treatment",
    y = "QUBIT (ng/uL)"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) + 
  scale_color_manual(values = cols.method) + 
  scale_fill_manual(values = cols.method) + scale_y_sqrt()

# 10. Summary bar plot with error bars (QUBIT) - CORRECTED
treatment_summary <- qubit_by_sample %>%
  group_by(TREATMENT) %>%
  summarise(
    mean_qubit = mean(QUBIT, na.rm = TRUE),
    se_qubit = sd(QUBIT, na.rm = TRUE) / sqrt(n()),
    n = n()
  )


meanqubitplot <- ggplot(treatment_summary, aes(x = TREATMENT, y = mean_qubit, fill = TREATMENT)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  geom_errorbar(aes(ymin = mean_qubit - se_qubit, ymax = mean_qubit + se_qubit),
                width = 0.2) +
  labs(x = "", y = "Total DNA Concentraton (ng/uL)") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_blank()) + 
  scale_color_manual(values = cols.method) + 
  scale_fill_manual(values = cols.method) + scale_y_sqrt()

meanqubitplot
```


### Read abundance

```{r}
# By Run
aggregate(abundance ~ Run, data = data_long, FUN = sum)
# or with dplyr
library(dplyr)
data_long %>% group_by(Run) %>% summarise(total_abund = sum(abundance))

# By Treatment
data_long %>% group_by(TREATMENT) %>% summarise(total_abund = sum(abundance))

# By Filtration
data_long %>% group_by(FILTRATION) %>% summarise(total_abund = sum(abundance))

# By Extraction
data_long %>% group_by(EXTRACTION) %>% summarise(total_abund = sum(abundance))

# By all four factors
data_long %>% 
  group_by(TREATMENT, FILTRATION, EXTRACTION, Run) %>% 
  summarise(
    total_abund = sum(abundance),
    mean_abund = mean(abundance),
    n = n()
  ) %>% 
  arrange(desc(total_abund))

# Sum abundance across treatments and runs
library(tidyr)
data_long %>%
  group_by(TREATMENT, Run) %>%
  summarise(total_abund = sum(abundance)) %>%
  pivot_wider(names_from = Run, values_from = total_abund, values_fill = 0)

# Overall summary by Treatment
data_long %>%
  group_by(TREATMENT) %>%
  summarise(
    total = sum(abundance),
    mean = mean(abundance),
    median = median(abundance),
    min = min(abundance),
    max = max(abundance),
    sd = sd(abundance),
    n = n()
  )

data_long <- data_long %>%
  mutate(Run = as.factor(Run))

# 1. By TREATMENT - Boxplot with jitter
ggplot(data_long, aes(x = TREATMENT, y = abundance, fill = TREATMENT)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.8) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 2, fill = "white") +
  scale_y_log10() +  # Log scale if data spans many orders of magnitude
  labs(
    title = "Abundance Distribution by Treatment",
    x = "Treatment",
    y = "Abundance (log scale)"
  ) +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) + scale_color_manual(values = cols.method) + scale_fill_manual(values = cols.method)

# 2. By Run - Boxplot
ggplot(data_long, aes(x = Run, y = abundance, fill = Run)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.8) +
  scale_y_log10() +
  labs(
    title = "Abundance Distribution by Run",
    x = "Run",
    y = "Abundance (log scale)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")


# 5. Treatment × Run interaction plot
treatment_run_summary <- data_long %>%
  group_by(TREATMENT, Run) %>%
  summarise(
    mean_abund = mean(abundance, na.rm = TRUE),
    sd_abund = sd(abundance, na.rm = TRUE),
    n = n()
  ) %>%
  mutate(se_abund = sd_abund / sqrt(n)) 

ggplot(treatment_run_summary, aes(x = Run, y = mean_abund, color = TREATMENT, group = TREATMENT)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_abund - se_abund, ymax = mean_abund + se_abund),
                width = 0.2, alpha = 0.5) +
  labs(
    title = "Treatment × Run Interaction",
    x = "Run",
    y = "Mean Abundance",
    color = "Treatment"
  ) +
  theme_minimal()+ scale_color_manual(values = cols.method) + scale_fill_manual(values = cols.method)


# 7. Faceted by Treatment: Abundance by Run
ggplot(data_long, aes(x = Run, y = abundance, fill = Run)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~TREATMENT, scales = "free_y") +
  scale_y_log10() +
  labs(
    title = "Abundance by Run within Each Treatment",
    x = "Run",
    y = "Abundance (log scale)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# 8. Faceted by Run: Abundance by Treatment
ggplot(data_long, aes(x = TREATMENT, y = abundance, fill = TREATMENT)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~Run, scales = "free_y") +
  scale_y_log10() +
  labs(
    title = "Abundance by Treatment within Each Run",
    x = "Treatment",
    y = "Abundance (log scale)"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))+ scale_color_manual(values = cols.method) + scale_fill_manual(values = cols.method)

# 10. Summary bar plot with error bars
treatment_summary <- data_long %>%
  group_by(TREATMENT) %>%
  summarise(
    mean_abund = mean(abundance, na.rm = TRUE),
    se_abund = sd(abundance, na.rm = TRUE) / sqrt(n()),
    n = n()
  )

meanreadabund=ggplot(treatment_summary, aes(x = TREATMENT, y = mean_abund, fill = TREATMENT)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  geom_errorbar(aes(ymin = mean_abund - se_abund, ymax = mean_abund + se_abund),
                width = 0.2) +
  labs(
    x = "Treatment",
    y = "Read Abundance"
  ) +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))+ scale_color_manual(values = cols.method) + scale_fill_manual(values = cols.method) +scale_y_sqrt()
meanreadabund

dnaabundplot=ggarrange(meanqubitplot, meanreadabund, nrow = 2, labels = "AUTO")

ggsave(
  filename=file.path(paths$results_figs,paste0("DNA_abundance_", GROUPING,"_teleo2.pdf")),
  plot    =dnaabundplot,
  width   =8,
  height  =6,
  units   ="in",
  device  =cairo_pdf,
  family  ="Arial"
)
```

### Contamination

#### Load

```{r}
rownames(reads.removed)=reads.removed$ID
reads.removed$Mean.blank=NULL
reads.removed$ID=NULL
reads.removed.t=as.data.frame(t(reads.removed))
reads.removed.t[[SAMPLE]]=rownames(reads.removed.t)

contaminatedreads=merge(reads.removed.t, metadata, by=SAMPLE)


otu.cols <- !(names(contaminatedreads) %in%
    c(SAMPLE,"Index","Run","Marker","PrimerF","PrimerR",
      "FIELD_ID","FILTRATION","EXTRACTION","TREATMENT",
      "Unique_SAMPLE_ID","LAT","LON","SITE"))

removed <- contaminatedreads

removed[otu.cols] <- lapply(removed[otu.cols], function(x)
    as.numeric(replace(x, is.na(x), 0)))

removed$contam_reads <- rowSums(removed[, otu.cols])

contamreads=ggplot(removed, aes(x = TREATMENT, y = contam_reads, fill = TREATMENT)) +
  geom_violin(alpha = 0.5, scale = "width") +
  geom_boxplot(width = 0.2, alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.8) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 2, fill = "white") +
  scale_y_log10() +
  #scale_y_sqrt() +
  labs(y = "Contaminant Reads", x="") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_blank()) + scale_fill_manual(values = cols.method)
contamreads

library(glmmTMB)
contamload.mod=glmmTMB(contam_reads ~ TREATMENT + (1|SITE),
        family=nbinom2,
        data=removed)
contamload.mod
summary(contamload.mod)
```

#### Frequency

```{r}
removed$contam_otus <-
    rowSums(removed[, otu.cols] > 0, na.rm = TRUE)

contamfreq.mod=glmmTMB(contam_otus ~ TREATMENT,
        family = nbinom2,
        data = removed)

summary(contamfreq.mod)

library(emmeans)
emmeans(contamfreq.mod, pairwise ~ TREATMENT, type = "response")

# Visualize contaminant OTUs by treatment
contamotus=ggplot(removed, aes(x = TREATMENT, y = contam_otus, fill = TREATMENT)) +
  geom_violin(alpha = 0.5, scale = "width") +
  geom_boxplot(width = 0.2, alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.8) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 2, fill = "white") +
  scale_y_log10() +
  labs(y = "Contaminant OTUs") +
  theme_minimal() +
  theme(
    legend.position = "none", 
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  ) + 
  scale_fill_manual(values = cols.method) + scale_y_sqrt()
contamotus

# Model contaminant OTUs with GLMM (including SITE as random effect)
contamfreq.mod <- glmmTMB(
  contam_otus ~ TREATMENT + (1|SITE),
  family = nbinom2,
  data = removed
)

# Model summary
summary(contamfreq.mod)

# Check model diagnostics
# Residuals
simulationOutput <- DHARMa::simulateResiduals(contamfreq.mod)
plot(simulationOutput)

# Get treatment means and pairwise comparisons
library(emmeans)

# Estimated marginal means
emmeans(contamfreq.mod, ~ TREATMENT, type = "response")

# Pairwise comparisons
pairwise_comparisons <- emmeans(contamfreq.mod, pairwise ~ TREATMENT, type = "response")
print(pairwise_comparisons)

# Extract just the pairwise comparisons table
pairwise_comparisons$contrasts

# If you want compact letter display
library(multcomp)
cld_contamfreq <- cld(emmeans(contamfreq.mod, ~ TREATMENT, type = "response"),
                      Letters = letters,
                      adjust = "tukey")
print(cld_contamfreq)

# Create a summary table of contaminant OTUs by treatment
contam_otus_summary <- removed %>%
  group_by(TREATMENT) %>%
  summarise(
    n_samples = n(),
    mean_otus = mean(contam_otus, na.rm = TRUE),
    median_otus = median(contam_otus, na.rm = TRUE),
    sd_otus = sd(contam_otus, na.rm = TRUE),
    se_otus = sd_otus / sqrt(n()),
    min_otus = min(contam_otus, na.rm = TRUE),
    max_otus = max(contam_otus, na.rm = TRUE)
  )

# Export results
# Save pairwise comparisons
pairwise_df <- as.data.frame(pairwise_comparisons$contrasts)


contamplot=ggarrange(contamreads, contamotus, nrow = 2, labels = "AUTO")
contamplot

ggsave(
  filename=file.path(paths$results_figs,paste0("Contaminated_", GROUPING,"_teleo2.pdf")),
  plot    =contamplot,
  width   =8,
  height  =8,
  units   ="in",
  device  =cairo_pdf,
  family  ="Arial"
)
```


#### Proportion (Not working NA treatments)

```{r}
#### Proportion of Reads Removed from Original (Pre-Decontamination) Reads

# Extract the data frames
reads_removed <- microdecon.decontaminated$reads.removed
otutab_predecon <- otutab.predecon  # Your original raw OTU table

library(tidyverse)

# DATA STRUCTURES:
# reads_removed: 
#   - Rows = OTUs (contaminants, identified by ID)
#   - Columns = Samples (named with run_barcode names) + ID + Mean.blank
#   - Values = number of reads of that contaminant OTU removed from that sample

# otutab_predecon:
#   - Rows = ALL OTUs (identified by ID)  
#   - Columns = Samples (named with run_barcode names) + ID
#   - Values = number of raw reads for that OTU in that sample

# Get sample names from pre-decontamination table (exclude ID column)
sample_names <- colnames(otutab_predecon)[!(colnames(otutab_predecon) %in% c("ID"))]

# Verify that reads_removed has the same sample columns
sample_names_removed <- colnames(reads_removed)[!(colnames(reads_removed) %in% c("ID", "Mean.blank"))]

cat("Number of samples in pre-decon table:", length(sample_names), "\n")
cat("Number of samples in reads_removed:", length(sample_names_removed), "\n")

# Use intersection to be safe
common_samples <- intersect(sample_names, sample_names_removed)
cat("Number of common samples:", length(common_samples), "\n")

# CORRECT METHOD: Calculate total reads removed per sample from reads_removed
total_reads_removed <- colSums(reads_removed[, common_samples, drop = FALSE], na.rm = TRUE)

# Calculate total original reads per sample from pre-decontamination table
total_reads_original <- colSums(otutab_predecon[, common_samples, drop = FALSE], na.rm = TRUE)

# Calculate number of contaminant OTUs per sample (OTUs with >0 reads removed)
n_contaminant_otus <- colSums(reads_removed[, common_samples, drop = FALSE] > 0, na.rm = TRUE)

# Calculate total OTUs in original sample
n_total_otus <- colSums(otutab_predecon[, common_samples, drop = FALSE] > 0, na.rm = TRUE)

# Create summary data frame
sample_stats <- data.frame(
  SAMPLE = common_samples,
  total_reads_original = total_reads_original[common_samples],
  total_reads_removed = total_reads_removed[common_samples],
  n_total_otus = n_total_otus[common_samples],
  n_contaminant_otus = n_contaminant_otus[common_samples],
  stringsAsFactors = FALSE
) %>%
  mutate(
    # Calculate proportion removed from original reads
    proportion_removed = ifelse(total_reads_original == 0, 0, total_reads_removed / total_reads_original),
    percent_removed = proportion_removed * 100,
    
    # Optional: Calculate remaining reads (for verification)
    total_reads_remaining = total_reads_original - total_reads_removed,
    
    # Proportion of OTUs that were contaminants
    proportion_contaminant_otus = ifelse(n_total_otus == 0, 0, n_contaminant_otus / n_total_otus),
    percent_contaminant_otus = proportion_contaminant_otus * 100
  ) %>%
  arrange(desc(percent_removed))

# View results
print(head(sample_stats))

# Summary statistics
cat("\nSummary of percent removed from original reads:\n")
print(summary(sample_stats$percent_removed))

cat("\nSummary of percent contaminant OTUs:\n")
print(summary(sample_stats$percent_contaminant_otus))

# Check if any samples have zero original reads
zero_samples <- sample_stats %>% filter(total_reads_original == 0)
if(nrow(zero_samples) > 0) {
  cat("\nSamples with zero total reads:", nrow(zero_samples), "\n")
  print(zero_samples$SAMPLE)
}

# Verify that removed reads don't exceed original reads
sample_stats <- sample_stats %>%
  mutate(
    check_removed_exceeds_original = total_reads_removed > total_reads_original
  )

if(any(sample_stats$check_removed_exceeds_original)) {
  cat("\nWarning: Some samples have more removed reads than original reads!\n")
  print(sample_stats %>% filter(check_removed_exceeds_original) %>% 
          dplyr::select(SAMPLE, total_reads_original, total_reads_removed))
}

# Add Mean.blank from reads_removed
sample_stats$Mean.blank <- unique(reads_removed$Mean.blank)[1]

# Join metadata
sample_stats_with_treatment <- sample_stats %>%
  left_join(metadata, by = c("SAMPLE" = "SAMPLE_ID"))

# Check for missing metadata
missing_treatment <- sample_stats_with_treatment %>%
  filter(is.na(TREATMENT))

if(nrow(missing_treatment) > 0) {
  cat("\nWarning:", nrow(missing_treatment), "samples have no treatment information:\n")
  print(missing_treatment$SAMPLE)
} else {
  cat("\nAll samples matched with treatment information!\n")
}

# Visualizations
# Make sure cols.method is defined
if(!exists("cols.method")) {
  # Define some default colors
  treatments <- unique(sample_stats_with_treatment$TREATMENT)
  if(requireNamespace("RColorBrewer", quietly = TRUE)) {
    cols.method <- setNames(RColorBrewer::brewer.pal(min(8, length(treatments)), "Set2"), treatments)
  } else {
    cols.method <- setNames(rainbow(length(treatments)), treatments)
  }
}

# 1. Violin + Boxplot of percent removed by treatment
ggplot(sample_stats_with_treatment, aes(x = TREATMENT, y = percent_removed, fill = TREATMENT)) +
  geom_violin(alpha = 0.5, scale = "width") +
  geom_boxplot(width = 0.2, alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.8) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 2, fill = "white") +
  labs(
    title = "Percent of Original Reads Removed as Contaminants by Treatment",
    y = "Percent Reads Removed (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none", 
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  ) + 
  scale_fill_manual(values = cols.method) + scale_y_sqrt()

# 3. Scatter plot: Original reads vs Removed reads
ggplot(sample_stats_with_treatment, aes(x = total_reads_original, 
                                         y = total_reads_removed,
                                         color = TREATMENT)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    title = "Original Reads vs Removed Reads by Treatment",
    x = "Total Original Reads (log scale)",
    y = "Total Reads Removed (log scale)",
    color = "Treatment"
  ) +
  theme_minimal() +
  scale_color_manual(values = cols.method)

# Statistical Modeling (Beta regression)
library(glmmTMB)
library(emmeans)
library(DHARMa)

# Convert percent to proportion (0-1 scale) and adjust for beta regression
sample_stats_with_treatment <- sample_stats_with_treatment %>%
  mutate(proportion_removed = percent_removed / 100)

# Adjust proportions to avoid 0 and 1 for beta regression
n <- nrow(sample_stats_with_treatment)
sample_stats_with_treatment <- sample_stats_with_treatment %>%
  mutate(
    proportion_removed_adj = (proportion_removed * (n - 1) + 0.5) / n,
    # Check if any are still 0 or 1
    check_zero = proportion_removed_adj == 0,
    check_one = proportion_removed_adj == 1
  )

# Beta regression with random effect for site
prop_removed.mod <- glmmTMB(
  proportion_removed_adj ~ TREATMENT + (1|SITE),
  family = beta_family(link = "logit"),
  data = sample_stats_with_treatment
)

summary(prop_removed.mod)

# Check model diagnostics
simulationOutput <- simulateResiduals(prop_removed.mod)
plot(simulationOutput)

# Get treatment means and pairwise comparisons
emmeans(prop_removed.mod, ~ TREATMENT, type = "response")

# Pairwise comparisons
pairwise_prop <- emmeans(prop_removed.mod, pairwise ~ TREATMENT, type = "response")
print(pairwise_prop)

# Extract just the pairwise comparisons table
pairwise_df <- as.data.frame(pairwise_prop$contrasts)
print(pairwise_df)

# Compact letter display
library(multcomp)
cld_prop <- cld(emmeans(prop_removed.mod, ~ TREATMENT, type = "response"),
                Letters = letters,
                adjust = "tukey")
print(cld_prop)

# Create summary table by treatment
prop_summary <- sample_stats_with_treatment %>%
  group_by(TREATMENT) %>%
  summarise(
    n_samples = n(),
    mean_percent_removed = mean(percent_removed, na.rm = TRUE),
    median_percent_removed = median(percent_removed, na.rm = TRUE),
    sd_percent_removed = sd(percent_removed, na.rm = TRUE),
    se_percent_removed = sd_percent_removed / sqrt(n()),
    min_percent_removed = min(percent_removed, na.rm = TRUE),
    max_percent_removed = max(percent_removed, na.rm = TRUE),
    mean_total_reads_original = mean(total_reads_original, na.rm = TRUE),
    mean_total_reads_removed = mean(total_reads_removed, na.rm = TRUE),
    mean_percent_contaminant_otus = mean(percent_contaminant_otus, na.rm = TRUE)
  )

print(prop_summary)

# Optional: Export results
# write.csv(prop_summary, "proportion_removed_by_treatment_summary.csv", row.names = FALSE)
# write.csv(pairwise_df, "proportion_removed_pairwise_comparisons.csv", row.names = FALSE)
# write.csv(sample_stats_with_treatment, "sample_stats_with_metadata.csv", row.names = FALSE)
```

#### Negatives

```{r}
total_reads_removed_per_sample <- colSums(reads_removed[, sample_names_removed, drop = FALSE], na.rm = TRUE)

# Create a data frame with sample-level removal summary
sample_removal_summary <- data.frame(
  SAMPLE = sample_names_removed,
  total_reads_removed = total_reads_removed_per_sample,
  n_contaminant_otus = colSums(reads_removed[, sample_names_removed, drop = FALSE] > 0, na.rm = TRUE),
  max_contaminant_reads = apply(reads_removed[, sample_names_removed, drop = FALSE], 2, max, na.rm = TRUE),
  mean_contaminant_reads = apply(reads_removed[, sample_names_removed, drop = FALSE], 2, mean, na.rm = TRUE),
  median_contaminant_reads = apply(reads_removed[, sample_names_removed, drop = FALSE], 2, median, na.rm = TRUE),
  stringsAsFactors = FALSE
) %>%
  arrange(desc(total_reads_removed))

# View top samples with most reads removed
cat("\n========== TOP 20 SAMPLES WITH MOST READS REMOVED ==========\n")
print(head(sample_removal_summary, 20))


# Get all column names from otutab.predecon
all_cols <- colnames(otutab.predecon)

# Find control columns (ending with _EB or _PB)
control_cols <- all_cols[grepl("_EB$|_PB$", all_cols)]

cat("Found", length(control_cols), "control columns\n")
cat("First 10 controls:\n")
print(head(control_cols, 10))

# Calculate total reads in each control column
control_reads <- data.frame(
  SAMPLE = control_cols,
  total_reads = colSums(otutab.predecon[, control_cols, drop = FALSE], na.rm = TRUE),
  n_otus = colSums(otutab.predecon[, control_cols, drop = FALSE] > 0, na.rm = TRUE),
  stringsAsFactors = FALSE
) %>%
  mutate(
    # Extract run, barcode, and type (EB or PB)
    Run = ifelse(grepl("^run1", SAMPLE), 1,
                ifelse(grepl("^run2", SAMPLE), 2,
                ifelse(grepl("^run4", SAMPLE), 4, NA))),
    Type = ifelse(grepl("_EB", SAMPLE), "Extraction Blank",
                 ifelse(grepl("_PB", SAMPLE), "PCR Blank", NA)),
    Marker = ifelse(grepl("Teleo2", SAMPLE), "Teleo2",
                   ifelse(grepl("MiBird", SAMPLE), "MiBird", "Unknown"))
  ) %>%
  arrange(desc(total_reads))

# View all controls ranked by contamination
cat("\nCONTROL SAMPLES RANKED BY CONTAMINATION (highest first):\n")
print(control_reads)

# Top 10 most contaminated controls
top_controls <- control_reads %>%
  head(10)

cat("\nTOP 10 MOST CONTAMINATED CONTROLS:\n")
print(top_controls[, c("SAMPLE", "total_reads", "n_otus", "Run", "Type", "Marker")])

# Identify the single most contaminated control
most_contaminated <- control_reads %>%
  slice_max(total_reads, n = 1)

cat("\n★ MOST CONTAMINATED CONTROL SAMPLE ★\n")
print(most_contaminated[, c("SAMPLE", "total_reads", "n_otus", "Run", "Type", "Marker")])

# Visualizations

# 1. Bar plot of top 10 controls
ggplot(top_controls, aes(x = reorder(SAMPLE, total_reads), y = total_reads, fill = Type)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Top 10 Most Contaminated Control Samples (Pre-Decontamination)",
    x = "Control Sample",
    y = "Total Reads",
    fill = "Control Type"
  ) +
  theme_minimal() + scale_y_log10()


# Summary statistics
summary_stats <- control_reads %>%
  group_by(Type) %>%
  summarise(
    n = n(),
    mean_reads = mean(total_reads, na.rm = TRUE),
    median_reads = median(total_reads, na.rm = TRUE),
    sd_reads = sd(total_reads, na.rm = TRUE),
    max_reads = max(total_reads, na.rm = TRUE),
    mean_otus = mean(n_otus, na.rm = TRUE),
    max_otus = max(n_otus, na.rm = TRUE)
  )

cat("\nSUMMARY STATISTICS BY CONTROL TYPE:\n")
print(summary_stats)

# Summary by Run
run_stats <- control_reads %>%
  group_by(Run) %>%
  summarise(
    n = n(),
    mean_reads = mean(total_reads, na.rm = TRUE),
    median_reads = median(total_reads, na.rm = TRUE),
    max_reads = max(total_reads, na.rm = TRUE)
  )

cat("\nSUMMARY STATISTICS BY RUN:\n")
print(run_stats)

# Summary by Marker
marker_stats <- control_reads %>%
  group_by(Marker) %>%
  summarise(
    n = n(),
    mean_reads = mean(total_reads, na.rm = TRUE),
    median_reads = median(total_reads, na.rm = TRUE),
    max_reads = max(total_reads, na.rm = TRUE)
  )

cat("\nSUMMARY STATISTICS BY MARKER:\n")
print(marker_stats)

# Export results
#write.csv(control_reads, "control_contamination_predecon.csv", row.names = FALSE)

# Find controls with reads above a threshold (e.g., > 1000 reads)
high_contamination <- control_reads %>%
  filter(total_reads > 1000) %>%
  arrange(desc(total_reads))

cat("\nCONTROLS WITH >1000 READS (HIGHLY CONTAMINATED):\n")
if(nrow(high_contamination) > 0) {
  print(high_contamination[, c("SAMPLE", "total_reads", "n_otus", "Type", "Run")])
} else {
  cat("None found.\n")
}

# Identify which OTUs are driving the contamination in the most contaminated control
if(nrow(most_contaminated) > 0) {
  most_contaminated_sample <- most_contaminated$SAMPLE[1]
  
  # Get the OTUs present in this control
  control_otus <- otutab.predecon %>%
    dplyr::select(ID, !!most_contaminated_sample) %>%
    filter(!!sym(most_contaminated_sample) > 0) %>%
    arrange(desc(!!sym(most_contaminated_sample)))
  
  cat("\nTOP OTUS IN THE MOST CONTAMINATED CONTROL (", most_contaminated_sample, "):\n", sep = "")
  print(head(control_otus, 20))
}
```


## 3. Community dissimilarity

This code calculates community dissimilarity using a chosen distance method (e.g., Bray-Curtis), then runs non-metric multidimensional scaling (NMDS) to visualize how samples cluster in 2D space, with ellipses showing 95% confidence intervals for each group. It then uses envfit to identify which environmental or metadata variables are significantly correlated with the community patterns (p < 0.05), adding them as arrows to the NMDS plot to show direction and strength of influence.

### MDS

```{r}
table.dist<- vegdist(community.table.t, method=distance.method, binary=F) # distance matrix from community OTU table
metatable.reps$group_factor <- as.factor(metatable.reps[[GROUPING]]) # grouping variable for ordination

set.seed(13) 

NMDSdat<-metaMDS(table.dist, k=2, weakties=TRUE, maxit=1000, pc=TRUE, zerodist="add", autotransform = F) # ordination function, with k dimensions 

stress_value <- round(NMDSdat$stress, 3) #extract 2D stress value (watch stress, should be < 0.15, lower the better)

PDat <- data.frame(
  NMDS1 = NMDSdat$points[, 1],
  NMDS2 = NMDSdat$points[, 2],
  row.names(table.t)) # creates dataframe of ordination axes results

names(PDat)[3] <- SAMPLE # add in sample information
PDat<-merge(PDat, metatable.reps, by=SAMPLE) # merge with metadata

nmds.plot <- qplot(x = NMDS1, y = NMDS2, 
                   colour = group_factor,
                   alpha = I(0.6), 
                   data = PDat) +  
  geom_vline(xintercept = 0, lty = 2, col = "grey66") + 
  geom_hline(yintercept = 0, lty = 2, col = "grey66") + 
  stat_ellipse(aes(fill = group_factor),
               level = 0.95, 
               geom = "polygon", 
               alpha = 0.15,
               colour = NA) +  # Remove outline
  theme_bw() +  
  labs(colour = NULL, 
       fill = NULL, 
       subtitle = paste("Stress =", stress_value)) + 
  scale_color_manual(values = cols.method) +
  scale_fill_manual(values = cols.method)
nmds.plot

ggsave(
  filename=file.path(paths$results_figs,paste0("NMDSplot_", GROUPING,"_teleo2.pdf")),
  plot    =nmds.plot,
  width   =6,
  height  =4,
  units   ="in",
  device  =cairo_pdf,
  family  ="Arial"
)
```


### Dispersion test

Tests for homogeneity of dispersion, The p-value is greater than 0.05, indicating that the homogeneity of dispersion across the groups is not significantly different. In other words, the variation within the groups (dispersion) is similar, and there is no strong evidence to suggest that one group is more or less dispersed than the others.The lack of significance in betadisper indicates that the spread of points (dispersion) within STEP groups is not significantly different, meaning that the differences in the dissimilarities found in adonis2 are more likely due to differences in group centroids rather than differences in variability within the groups.

```{r}
# Check for variability across SITE groups
dispersion <- betadisper(table.dist, group = metatable.reps[[GROUPING]])
anova(dispersion) # if significant -> ANOSIM suggested over PERMANOVA
```

### Centroid tests

This code runs two statistical tests to check if your groups (e.g., sites or treatments) have significantly different microbial communities. ANOSIM tests whether differences between groups are larger than differences within groups (R close to 1 = strong separation), while PERMANOVA tests if group membership explains a significant portion of the variation in community composition (p-value < 0.05 indicates significant differences).

#### ANOSIM

```{r}
anosim.mod=anosim(table.dist, group = metatable.reps$group_factor, permutations = numperm)
anosim.mod
```

#### PERMANOVA

```{r}
permanova.mod=adonis2(table.dist ~ group_factor, data = metatable.reps, permutations = numperm)
permanova.mod 
```

## 4. Hill diversity

This code calculates Hill diversity (richness at q=0, Shannon at q=1, and Simpson at q=2) for each sample using the iNEXT package, then standardizes all samples to the same coverage level (95% of the minimum coverage) for fair comparison. It produces two visualizations: group comparison plots showing diversity across your grouping variable (e.g., sites or treatments) using violin-boxplots, and a richness over time plot with a dashed line connecting medians across ordered dates.

#### Calculation

```{r}
# Get total number of columns
total_cols <- ncol(metatable.reps.reads)

# Define columns to remove: first and all metadata columns
cols_to_remove <- c(1, (total_cols - (ncol(metadata)-2)):total_cols)

# Subset the data frame to keep only the other columns
otu_df <- metatable.reps.reads[, -cols_to_remove]

# Force numeric and replace NA with 0
otu_df <- as.data.frame(lapply(otu_df, function(x) as.numeric(as.character(x))))
rownames(otu_df) <- metatable.reps.reads[[SAMPLE]]

otu_df[is.na(otu_df)] <- 0

# Remove samples with zero total reads
otu_df <- otu_df[rowSums(otu_df) > 0, ]

otu_list_samples <- apply(
  otu_df,
  1,                # apply over rows (samples)
  function(x) x[x > 0]   # keep only OTUs with >0 abundance
)

# Drop any remaining empty samples
otu_list_samples <- otu_list_samples[lengths(otu_list_samples) > 0]
otu_list_samples <- otu_list_samples[names(otu_list_samples) != ""] # remove any empty factors from the list

inext_out <- iNEXT(otu_list_samples, q = c(0, 1, 2), datatype = "abundance", se = FALSE)
```

#### Minimum coverage standardization

Not reliable if minimum coverage = 0.99-1.00! Do not remove singleton OTUs during bioinformatics! Use size-based (read depth) standardization instead.

```{r}
##Get coverage information
coverage_info <- DataInfo(otu_list_samples, datatype = "abundance")
target_coverage <- min(coverage_info$SC, na.rm = TRUE)
  
#coverage_info=inext_out$iNextEst$coverage_based # only want observed SC!!
#coverage_info=subset(coverage_info, Method == "Observed")

##Find the minimum coverage among samples (or choose a specific target)
#target_coverage=min(coverage_info$SC, na.rm=TRUE) # use minimum sample coverage across all samples
cat("Target coverage:", target_coverage, "\n")

####Coverage-Standardization ####################################
standardized.coverage=estimateD(
  otu_list_samples,
  q=c(0, 1, 2),
  datatype="abundance",
  base="coverage",  #Standardize by coverage
 level=target_coverage, conf=0.95)
#############################################################
```


#### Minimum read depth standardization

Standardizes DNA reads to minimum shared read depth, while also filtering out extremely low-read samples. Set the low-read threshold cutoff to remove samples with low sequencing depth. 

```{r}
# Calculate read depths for all samples
sample_reads <- sapply(otu_list_samples, sum)

# Set your threshold (e.g., 1000 reads)
read_threshold <- 100

# Filter samples below threshold
samples_to_keep <- names(sample_reads[sample_reads >= read_threshold])
otu_list_filtered <- otu_list_samples[samples_to_keep]

# Check how many samples were removed
removed_samples <- length(otu_list_samples) - length(otu_list_filtered)
cat("Removed", removed_samples, "samples with <", read_threshold, "reads\n")

# Now proceed with standardization on filtered data
sample_reads_filtered <- sapply(otu_list_filtered, sum)
min_read_depth <- min(sample_reads_filtered)

# Run iNEXT
standardized.size <- estimateD(
  otu_list_filtered,
  q = c(0, 1, 2),
  datatype = "abundance",
  base = "size", #standardize by reads
  level = min_read_depth
)
```

```{r}
hill_df_cov=standardized.coverage %>% 
  dplyr::select(Assemblage, Order.q, qD) %>%
  tidyr::pivot_wider(
    names_from=Order.q,
    values_from=qD,
    names_prefix="Hill_q"
  ) %>%
  dplyr::rename(!!SAMPLE := Assemblage) %>% 
  dplyr::rename(Richness_cov=Hill_q0) %>%
  dplyr::rename(Shannon_cov=Hill_q1) %>%
  dplyr::rename(Simpson_cov=Hill_q2)

hill_df_size=standardized.size%>%
  dplyr::select(Assemblage, Order.q, qD) %>%
  tidyr::pivot_wider(
    names_from=Order.q,
    values_from=qD,
    names_prefix="Hill_q"
  ) %>%
  dplyr::rename(!!SAMPLE := Assemblage) %>% 
  dplyr::rename(Richness_size=Hill_q0) %>%
  dplyr::rename(Shannon_size=Hill_q1) %>%
  dplyr::rename(Simpson_size=Hill_q2)

hill_df_ext=inext_out$AsyEst %>%
 dplyr::select(Assemblage, Diversity, Estimator) %>%
 tidyr::pivot_wider(
   names_from=Diversity,
   values_from=Estimator) %>%
 dplyr::rename(!!SAMPLE := Assemblage) %>% 
 dplyr::rename(Richness_ext=`Species richness`) %>%
 dplyr::rename(Shannon_ext=`Shannon diversity`) %>%
 dplyr::rename(Simpson_ext=`Simpson diversity`)

hill_df_obs = inext_out$AsyEst %>%
  dplyr::select(Assemblage, Diversity, Observed) %>%
  tidyr::pivot_wider(
    names_from = Diversity,
    values_from = Observed
  ) %>%
  dplyr::rename(!!SAMPLE := Assemblage) %>% 
  dplyr::rename(Richness = `Species richness`) %>%
  dplyr::rename(Shannon = `Shannon diversity`) %>%
  dplyr::rename(Simpson = `Simpson diversity`)

hill_df <- hill_df_obs %>%
  dplyr::full_join(hill_df_ext, by = SAMPLE) %>%
  dplyr::full_join(hill_df_size, by = SAMPLE) %>%
  dplyr::full_join(hill_df_cov, by = SAMPLE)

#Merge with metadata
alpha_div=hill_df %>%
  left_join(metadata, by=SAMPLE)

alpha_summary <- alpha_div %>%
  group_by(TREATMENT) %>%
  summarise(
    n = n(),
    Richness_mean = mean(Richness_size, na.rm = TRUE),
    Richness_sd = sd(Richness_size, na.rm = TRUE),
    Shannon_mean = mean(Shannon_size, na.rm = TRUE),
    Shannon_sd = sd(Shannon_size, na.rm = TRUE),
    Simpson_mean = mean(Simpson_size, na.rm = TRUE),
    Simpson_sd = sd(Simpson_size, na.rm = TRUE)
  )

alpha_summary=as.data.frame(alpha_summary)
head(alpha_summary)
```


#### Group diversity

```{r}
group.levels=as.factor(alpha_div[[GROUPING]])

p1 <- ggplot(alpha_div, aes(x=group.levels, y=Richness_size, fill=group.levels)) +
  geom_violin(alpha=0.5, scale = "width") + 
  geom_point(position=position_jitter(width=0.2), size=0.8) +
  geom_boxplot(width=0.1, outlier.shape=NA) +
  labs(title="Richness (q=0)", x="", y="") + theme_bw() +
  theme(legend.position="none", axis.text.x = element_text(angle=45, hjust=1)) + scale_color_manual(values = cols.method) + scale_fill_manual(values = cols.method)

p2 <- ggplot(alpha_div, aes(x=group.levels, y=Shannon_size, fill=group.levels)) +
  geom_violin(alpha=0.5, scale = "width") +
  geom_point(position=position_jitter(width=0.2), size=0.8) +
  geom_boxplot(width=0.1, outlier.shape=NA) +
  labs(title="Shannon (q=1)", x="", y="") +
  theme_bw() +  theme(legend.position="none", axis.text.x = element_text(angle=45, hjust=1)) + scale_color_manual(values = cols.method) + scale_fill_manual(values = cols.method)

p3 <- ggplot(alpha_div, aes(x=group.levels, y=Simpson_size, fill=group.levels)) +
  geom_violin(alpha=0.5, scale = "width") +
  geom_point(position=position_jitter(width=0.2), size=0.8) +
  geom_boxplot(width=0.1, outlier.shape=NA) +
  labs(title="Simpson (q=2)", x="", y="") +
theme_bw() +  theme(legend.position="none", axis.text.x = element_text(angle=45, hjust=1)) + scale_color_manual(values = cols.method) + scale_fill_manual(values = cols.method)

alphaplot <- ggarrange(p1, p2, p3, ncol=3)
alphaplot

ggsave(
  filename=file.path(paths$results_figs,paste0("alphaplot_", GROUPING, "teleo2.pdf")),
  plot    =alphaplot,
  width   =8,
  height  =5,
  units   ="in",
  device  =cairo_pdf,
  family  ="Arial"
)
```


## 5. Sampling effort

This code creates two types of accumulation curves: a sequence depth rarefaction curve showing how many distinct organisms (MOTUs) you detect as you sequence more reads per sample, and sample rarefaction curves showing how taxonomic coverage increases as you include more samples. The sample curves are generated both for all data combined and separately for different groups (defined by your GROUPING variable, like sites or treatments), using the iNEXT package to interpolate and extrapolate the curves with confidence intervals.

### Sequence depth rarefaction

```{r}
rarecurve(table.reads.t, step = 1000, col = rainbow(nrow(table.reads.t)), label = F, xlab = "Sequence Depth",  ylab = "MOTUs")
```

### Sample rarefaction

#### Overall

```{r}
# Combine data from all sites and steps into a single dataset
# Transpose and clean the data
smallset_combined <- t(metatable.reps[, c(2:ntax)])

# Remove samples with zero reads
smallset_combined <- smallset_combined[rowSums(smallset_combined) > 0, ]

# Remove MOTUs with zero reads across all samples
smallset_combined <- smallset_combined[, colSums(smallset_combined) > 0]

# Convert to presence-absence
smallset_combined <- decostand(smallset_combined, method = "pa")

# Create a list with the combined dataset
otu_list_combined <- list("Combined" = smallset_combined)

out_combined <- iNEXT(otu_list_combined, q = 0, datatype = "incidence_raw", se = TRUE)

info_samples <- DataInfo(otu_list_combined, datatype="incidence_raw")
min_coverage_samples <- min(info_samples$SC)

rarecurve_combined <- ggiNEXT(out_combined, type = 2, se = TRUE, facet.var = "None", color.var = "Order.q") 

rarecurve.overall=rarecurve_combined+ theme_bw() + xlab("Sample effort") + ylab("Coverage")

rarecurve.overall=rarecurve.overall + ylim(0,1) 

rarecurve.overall

ggsave(
  filename=file.path(paths$results_figs,paste0("rarecurve_teleo2.pdf")),
  plot    =rarecurve.overall,
  width   =8,
  height  =5,
  units   ="in",
  device  =cairo_pdf,
  family  ="Arial"
)
```


#### By grouping variable

##### Spatial

```{r}
# Get unique groups from your metadata
unique_groups <- unique(metadata[[GROUPING]])

# Create an empty list to store processed data for each group
otu_list <- list()

for(group in unique_groups) {
  # Get samples for this group
  samples_in_group <- metadata[metadata[[GROUPING]] == group, SAMPLE]
  
  # Subset and transpose (preserve matrix structure)
  temp_obj <- t(metatable.reps.reads[metatable.reps.reads[[SAMPLE]] %in% samples_in_group, c(2:ntax), drop = FALSE])
  
  # REMOVE INDIVIDUALS WITH ZERO SEQUENCE READS (rows with all zeros)
  temp_obj <- temp_obj[rowSums(temp_obj != 0) > 0, , drop = FALSE]
  
  # Skip if no rows remaining
  if(nrow(temp_obj) == 0) next
  
  # REMOVE MOTUS WITH ZERO SUMS (columns that sum to zero)
  temp_obj <- temp_obj[, colSums(temp_obj) != 0, drop = FALSE]
  
  # Skip if no columns remaining
  if(ncol(temp_obj) == 0) next
  
  # Convert to presence/absence
  temp_obj <- decostand(temp_obj, method = "pa")
  
  otu_list[[group]] <- temp_obj
}

otu_list <- otu_list[sapply(otu_list, length) > 0] 
otu_list <- otu_list[names(otu_list) != ""] # remove any empty factors from the list

# Now proceed with iNEXT analysis
out.site <- iNEXT(otu_list, q=0, datatype="incidence_raw", se=T, endpoint = 120)

# Get data info
info_sites <- DataInfo(otu_list, datatype="incidence_raw")
min_coverage_sites <- min(info_sites$SC)

# Create rarefaction curve
rarecurve.site <- ggiNEXT(out.site, type=2, se=T, facet.var="None", 
                          color.var="Assemblage", grey=FALSE)

rarecurve.site.plot=rarecurve.site + theme_bw() + xlab("Sample effort") + ylab("Coverage") + ylim(0,1)

rarecurve.site.plot + scale_color_manual(values = cols.method) + scale_fill_manual(values = cols.method) + ylim(0,1)

site.summary=out.site$AsyEst
site.summary

# Get the iNEXT output as a data frame
rare_data <- fortify(out.site, type=2)  # Extract data

# Ensure group names match your metadata
# Create color mapping
color_mapping <- cols.method[names(cols.method) %in% unique(rare_data$Assemblage)]

# Plot manually using ggplot
rarecurve.site.plot <- ggplot(rare_data, aes(x = x, y = y, color = Assemblage, fill = Assemblage)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = y.lwr, ymax = y.upr), alpha = 0.2, color = NA) + 
  theme_bw() +
  xlab("Sample effort") +
  ylab("Coverage") +
  ylim(0,1) +
  scale_color_manual(values = cols.method) +
  scale_fill_manual(values = cols.method) +
  labs(color = GROUPING, fill = GROUPING)

rarecurve.site.plot

ggsave(
  filename=file.path(paths$results_figs,paste0("rarecurve_", GROUPING, "_teleo2.pdf")),
  plot    =rarecurve.site.plot,
  width   =8,
  height  =5,
  units   ="in",
  device  =cairo_pdf,
  family  ="Arial"
)
```

## 6. Indicator taxa
This code aggregates OTUs by a higher taxonomic rank (e.g., genus or family), then runs indicator species analysis (IndVal) to identify which taxa are significantly associated with specific groups (e.g., sites or treatments). It filters significant indicators (adjusted p < 0.15), then creates a faceted time-series plot showing how the abundance of each indicator taxon changes over time, with separate panels for each group and each indicator taxon, using violin-boxplots and a dashed line connecting medians.

More info:
The indicator species analysis uses an index called Indicator Value (the square root of the
indicator value is shown on the generated table under column “stat”) which is the product of
two components also shown on the table: A and B.
A is the probability that a site is in the target group given that the species has been found. It
is the mean abundance of species in the target group divided by the sum of the mean
abundance values over all groups. Horsetail (EQUI.SPPS) has an A value of 1 because it
only occurs in the loamy-sand ecosite.
B is the probability that you will find the species in a site that is in the target site group. B is
the relative frequency of occurrence of the species in the target group. Western wheatgrass
(PASC.SMI) has a B value of 1 because it occurs in every loam site. Sandgrass
(CALA.LON) has a B value of 1 because it occurs in every loamy-sand site. A high B value
means that if you do not find the species at a given site, the probability that the site belongs
to the target group is low.
The indicator value is maximum (i.e. 1) when all individuals of a species are found in a
single group of sites and when the species occurs in all sites of that group. Western
wheatgrass, Sandgrass, and Horsetail seem to be strong indicators of their groups because
the square root of their indicator values are all nearly 1. They have high associations to their
site groups.

### Aggregate OTU table

```{r}
group=as.factor(metatable.reps[[GROUPING]])
taxonomic.rank=community.data[[TAXONOMICRANK]] 

com=community.table.t

colnames(com)=taxonomic.rank

prefixes = unique(sub("\\..*", "", colnames(com)))
prefixes = prefixes[prefixes != ""]

com=sapply(prefixes, function(x) rowSums(com[,startsWith(colnames(com), x), drop=F])) ##OTU table with combined taxonomies (no repeated columns)

com=as.data.frame(com)

ncom=ncol(com)
```

### IndVal analysis

```{r}
##BY SITE
# Run indicator species analysis
indval.site = multipatt(com, group, control = how(nperm = numperm), duleg = T, func="r.g")

# View summary
summary(indval.site)

# Extract results as a dataframe
summary_result = as.data.frame(indval.site$sign)

# Extract p-values from the results
pvals <- indval.site$sign$p.value

# Adjust p-values using Benjamini-Hochberg (FDR) correction
pvals_adj <- p.adjust(pvals, method = "BH")

# Add adjusted p-values back to results
summary_result$p.adjusted <- pvals_adj

# Filter significant species after correction (adjusted p-value < 0.05)
significant_species <- summary_result[pvals_adj < 0.05, ]

# Get the site/group that each indicator taxa is associated with
# For each row in significant_species, find which "s." column has value 1
indicator_groups <- apply(significant_species[, grep("^s\\.", names(significant_species))], 1, function(x) {
  # Find column names where value is 1
  sites <- names(x)[x == 1]
  # Remove the "s." prefix
  sites <- gsub("^s\\.", "", sites)
  # If multiple sites, combine with comma, otherwise return single site
  if(length(sites) > 1) {
    return(paste(sites, collapse = ", "))
  } else if(length(sites) == 1) {
    return(sites)
  } else {
    return(NA)
  }
})

# Add this information to significant_species
significant_species$indicator_group <- indicator_groups
significant_species$indicator_taxon <- rownames(significant_species)

significant_species
```


## 7. Heatmap

```{r}
# ============================================
# HEATMAP - LOG-TRANSFORMED SUMMED READS BY TREATMENT
# ============================================

library(pheatmap)

# Match samples
common_samples <- intersect(rownames(com), metadata[[SAMPLE]])
com_subset <- com[common_samples, ]
metadata_subset <- metadata[match(common_samples, metadata[[SAMPLE]]), ]

# Sum reads by treatment
sum_by_treatment <- aggregate(com_subset, 
                              by = list(Treatment = metadata_subset$TREATMENT), 
                              FUN = sum)

# Set treatment as rownames and remove the Treatment column
rownames(sum_by_treatment) <- sum_by_treatment$Treatment
sum_by_treatment <- sum_by_treatment[, -1]

# Transpose so taxa are rows, treatments are columns
heatmap_data <- t(sum_by_treatment)

# Optional: Remove taxa with zero reads across all treatments
#heatmap_data <- heatmap_data[rowSums(heatmap_data) > 0, ]

# Log transform (log1p = log(1+x) to handle zeros)
heatmap_data_log <- log1p(heatmap_data)

# Create heatmap (no scaling)
heatmap=pheatmap(heatmap_data_log,
         scale = "none", fontsize_row = 8,
         fontsize_col = 10,
         angle_col = 45,
         cellwidth = 30,
         cellheight = 10,
         show_rownames = TRUE,
         show_colnames = TRUE,
         cluster_rows = T,
         cluster_cols = F,
         color = colorRampPalette(c("white", "tan", "firebrick3"))(100))
heatmap

ggsave(
  filename=file.path(paths$results_figs,paste0("heatmap_", GROUPING, "_teleo2.pdf")),
  plot    =heatmap,
  width   =6,
  height  =12,
  units   ="in",
  device  =cairo_pdf,
  family  ="Arial")
```

