---
title: "eDNA Course - Diversity analysis"
author: "Patrick Nichols"
date: "2026-07-02"
output: html_document
---

## 1. Setup

Session -> Set working directory -> To source file location...

Carefully update the user-defined settings for each run, making sure column names and index positions match.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(iNEXT)
library(vegan)
library(ggpubr)
library(indicspecies)
library(microDecon)
library(purrr)
library(tibble)
library(readxl)

My_Theme = theme_set(theme_bw()+theme(
  panel.grid.minor = element_blank(),
  panel.background = element_blank(), 
  panel.border = element_rect(fill=NA, colour = "black"), 
  panel.spacing = unit(1, "lines"),
  axis.title.x = element_text(size = 12, color="black"),
  axis.text.x = element_text(size = 12, color="black"),
  axis.title.y = element_text(size = 12, color = "black"),
  axis.text.y = element_text(size=12, color = "black"),
  legend.text = element_text(size=12, color = "black"),
  strip.text.x = element_text(size=14, color="black"),
  strip.text.y = element_text(size=14, color="black"),
  strip.text = element_text(size = 14, color = "black"),
  strip.background = element_blank(), 
  strip.placement = "outside",
  title = element_text(size=14, color="black", face="bold"),
  axis.title = element_text(size=12, face="bold", colour = "black"),
  axis.text = element_blank()))

set.seed(13)
numperm=9999 #Permutation tests

paths=list(
  data_raw      ="data/raw",
  data_processed="data/processed",
  results_rds   ="results/rds",
  results_tables="results/tables",
  results_figs  ="results/figures"
)

walk(paths, dir.create, showWarnings = FALSE, recursive = TRUE)

cols.method=c("#031d44", "#1a659e", "#d5896f", "#dab785", "#70a288", "snow3", "black")

#### RUN SETTINGS #######################################################
num.blanks=27 # how many columns of blanks/negative control samples?
SAMPLE="Sample_ID" # name of metadata column with sample IDs
OTU_ID="OTU_ID" # name of OTU ID column in OTU table
GROUPING="TREATMENT" # name of grouping variable column in metadata
TAXONOMICRANK= "Genus" # choose taxonomic rank for indicator taxon analysis
distance.method="bray" #choose distance method for community dissimilarity ordination
#########################################################################

#### READ IN OTU TABLE ##################################################
otutab.raw=read_excel(file.path(paths$data_raw, "combined_data_read_table.xlsx"))  # raw OTU table, with Unique ID/OTU_ID as first column, apscale taxonomic information
otutab.predecon=otutab.raw[,c(1:257)] # select only OTU ID and numerical read count columns
otutab.predecon[, -1] <- lapply(otutab.predecon[, -1], function(x) {
  if(is.character(x)) as.numeric(trimws(x)) else x
})
# Remove zero-sum rows (keep ID column)
otutab.predecon <- otutab.predecon[rowSums(otutab.predecon[, -1]) != 0, ]
# Remove zero-sum columns (keep ID column)
otutab.predecon <- otutab.predecon[, c(1, which(colSums(otutab.predecon[, -1]) != 0) + 1)]
apscale.info=otutab.raw[,c(1,258:ncol(otutab.raw))] # select taxonomic info and other columns from apscale
#########################################################################

#### READ IN METADATA FILE #############################################
metadata=read_excel(file.path(paths$data_raw, "metadata.xlsx"))
metadata=subset(metadata, Run != "1")
metadata=subset(metadata, SITE=="Station2" | SITE=="Station3")
metadata=subset(metadata, SITE != "Control")
#metadata=subset(metadata, SAMPLE != "Teleo2_run4_barcode65" | SAMPLE != "Teleo2_run4_barcode68" | SAMPLE != "Teleo2_run2_barcode81") # failed samples?
metadata$QUBIT <- as.numeric(gsub(",", ".", trimws(metadata$QUBIT)))
#######################################################################
```

### Decontamination

#### MicroDecon

```{r}
otutab.predecon <- as.data.frame(otutab.predecon)

# Run decontamination
microdecon.decontaminated <- decon(otutab.predecon,numb.blanks = num.blanks, numb.ind = (ncol(otutab.predecon)-num.blanks-1),taxa = FALSE,thresh = 1,prop.thresh = 0, runs = 2)

reads.removed=microdecon.decontaminated$reads.removed
decontaminated.table <- microdecon.decontaminated$decon.table
decontaminated.table$Mean.blank <- NULL

reads.removed=microdecon.decontaminated$reads.removed
reads.removed=merge(apscale.info, reads.removed, by=OTU_ID)

decontaminated.table.metadata=merge(decontaminated.table, apscale.info, by=OTU_ID) # merge taxonomic info back into decontaminated file

write.csv(decontaminated.table.metadata, file=file.path(paths$data_processed, "decontaminated_data.csv"), row.names = F) ## save to PC
```


### Data cleaning and prepping

This block removes human-related taxa (genus Homo) from the dataset and renames the rows using organism IDs. It then isolates the taxonomic classification columns into a separate table called 'taxa'.

```{r}
otutab=decontaminated.table.metadata

# Filter unwanted taxa
data=subset(otutab, Order !="Primates") # if taxonomy information is present in OTU table
data=subset(data, Kingdom !="NoMatch") # if taxonomy information is present in OTU table
data=subset(data, Genus !="Cranioleuca") # if taxonomy information is present in OTU table


row.names(data)=data[[OTU_ID]]
final=data[,2:230] # Update me!!
rownames(final)=data[["OTU_ID"]]

final.t=as.data.frame(t(final))
final.t[[SAMPLE]]=rownames(final.t)
final.t.meta=merge(final.t, metadata, by=SAMPLE)

final.t=final.t.meta[,2:(nrow(data)+1)]
rownames(final.t)=final.t.meta[[SAMPLE]]

final=as.data.frame(t(final.t))
final <- final[, colSums(final) != 0] # remove zero sum columns
final <- final[rowSums(final) != 0, ] # remove zero sum rows
#final <- final[, colSums(final != 0) > 1] # remove samples with only 1 OTU
ntax=nrow(final)
nsample=ncol(final)
final[[OTU_ID]]=rownames(final)

final=merge(final, apscale.info, by=OTU_ID)
rownames(final)=final[[OTU_ID]]

# Taxonomic metadata only
taxa=final[, (nsample+2):ncol(final)]
rownames(taxa)=final[[OTU_ID]]
taxa[[OTU_ID]]=rownames(taxa)
```


### Transformations

This code prepares three versions of your OTU table (raw counts, Hellinger-transformed, and eDNA index), then lets you choose one for downstream analysis. The final steps transpose the table so taxa become columns and samples become rows, then merges it with site metadata for statistical testing.

#### Raw Reads (no transformation)

```{r}
# Raw reads OTU table only, no metadata
table.raw=final[, 2:(nsample+1)]
final_reads=final # make a copy for raw reads (will not undergo transformations)
table.reads=final_reads[,2:(nsample+1)]

sum(table.reads)

table.reads.t <- as.data.frame(t(table.reads))  # Flip OTU table so taxa as columns

table.reads.t[[SAMPLE]] <- row.names(table.reads.t) # add sample column to merge with metadata
metatable.reps.reads=merge(table.reads.t, metadata, by=SAMPLE) ##OTU table with site metadata correctly formatted for downstream analyses

table.reads.t[[SAMPLE]]=NULL #remove non-numeric columns for downstream analyses
```

#### Presence-Absence (Binary) 

```{r}
table_pa=decostand(table.reads, method = "pa")
table_pa[[OTU_ID]]=rownames(table_pa)
final_pa=merge(table_pa, taxa, by=OTU_ID)
rownames(final_pa)=final_pa[[OTU_ID]]
table_pa[[OTU_ID]]=NULL
table_pa.t=t(table_pa)
```

#### Hellinger 

```{r}
table_hel=decostand(table.reads, method = "hellinger")
table_hel[[OTU_ID]]=rownames(table_hel)
final_hel=merge(table_hel, taxa, by=OTU_ID)
rownames(final_hel)=final_hel[[OTU_ID]]
table_hel[[OTU_ID]]=NULL
table_hel.t=t(table_hel)
```

#### eDNA Index

```{r}
table.raw <- as.matrix(table.reads)  # Compute relative abundance per sample (column-wise)
relative_abundance <- sweep(table.raw, 2, colSums(table.raw), "/")  # Find the maximum observed abundance for each OTU across all samples (row-wise)
max_relative_abundance <- apply(relative_abundance, 1, max, na.rm = TRUE)  # Compute the eDNA index by dividing each OTU's abundance by its max
table_index <- as.data.frame(sweep(relative_abundance, 1, max_relative_abundance, "/"))  # Add OTU IDs back to the table

table_index[[OTU_ID]] <- rownames(table_index) # Merge with taxa metadata (assuming 'taxa' has OTU_ID as key)
final_index <- merge(table_index, taxa, by = OTU_ID) # Ensure row names match OTU_ID
rownames(final_index) <- final_index[[OTU_ID]] 
table_index[[OTU_ID]] <- NULL
table_index.t=t(table_index)
```

#### Prepare community data

```{r}
####CHOOSE MATRIX###################################################
community.table.t=table_index.t #transposed otu table version (OPTIONS: table_index.t, table_hel.t, table.reads.t, table_pa.t)
community.table=table_index #otu table only (OPTIONS: table_index, table_hel, table.reads, table_pa, etc.)
community.data=final_index #version with metadata (OPTIONS: final_index, final_hel, final_reads, final_pa, etc.)
###################################################################
```

#### Merge filtered metadata

```{r}
table.t <- as.data.frame(t(as.matrix(community.table)))

ntax=ncol(table.t)
nsample=nrow(table.t)

table.t[[SAMPLE]]=row.names(table.t) ##OTU table only

# Remove leading zeros from SAMPLE column
metatable.reps=merge(table.t, metadata, by=SAMPLE) ##OTU table with site metadata
table.t[[SAMPLE]]=NULL
```


####Melt data long format

```{r}
data_long <- as.data.frame(community.table.t) %>%
  rownames_to_column(var = SAMPLE) %>%
  pivot_longer(
    cols = -SAMPLE,  # Keep SampleID as identifier
    names_to = OTU_ID,
    values_to = "abundance"
  )

# Remove zero abundances (optional)
data_long <- data_long %>%
  filter(abundance > 0)

data_long=merge(data_long, metadata, by=SAMPLE)
data_long=merge(data_long, apscale.info, by=OTU_ID)
```

## 2. Treatment Summaries

### Total DNA Concentration

#### Plot

```{r}
# Create a distinct sample-level dataset for QUBIT values
qubit_by_sample <- data_long %>%
  distinct(!!sym(SAMPLE), .keep_all = TRUE) 

# OR, if you want to be more explicit:
qubit_by_sample <- data_long %>%
  dplyr::select(SAMPLE, Run, TREATMENT, FILTRATION, EXTRACTION, QUBIT, SITE) %>%
  distinct()

qubit_by_sample %>% 
  group_by(Run) %>% 
  summarise(total_qubit = sum(as.numeric(trimws(QUBIT)), na.rm = TRUE))

# By Treatment - QUBIT summary (corrected)
qubit_by_sample %>% 
  group_by(TREATMENT) %>% 
  summarise(total_qubit = sum(QUBIT, na.rm = TRUE))

# By Filtration - QUBIT summary (corrected)
qubit_by_sample %>% 
  group_by(FILTRATION) %>% 
  summarise(total_qubit = sum(QUBIT, na.rm = TRUE))

# By Extraction - QUBIT summary (corrected)
qubit_by_sample %>% 
  group_by(EXTRACTION) %>% 
  summarise(total_qubit = sum(QUBIT, na.rm = TRUE))

# By all four factors - QUBIT summary (corrected)
qubit_by_sample %>% 
  group_by(TREATMENT, FILTRATION, EXTRACTION, Run) %>% 
  summarise(
    total_qubit = sum(QUBIT, na.rm = TRUE),
    mean_qubit = mean(QUBIT, na.rm = TRUE),
    n = n()
  ) %>% 
  arrange(desc(total_qubit))

# Sum QUBIT across treatments and runs (corrected)
qubit_by_sample %>%
  group_by(TREATMENT, Run) %>%
  summarise(total_qubit = sum(QUBIT, na.rm = TRUE)) %>%
  pivot_wider(names_from = Run, values_from = total_qubit, values_fill = 0)

# Overall summary by Treatment - QUBIT (corrected)
qubit_by_sample %>%
  group_by(TREATMENT) %>%
  summarise(
    total = sum(QUBIT, na.rm = TRUE),
    mean = mean(QUBIT, na.rm = TRUE),
    median = median(QUBIT, na.rm = TRUE),
    min = min(QUBIT, na.rm = TRUE),
    max = max(QUBIT, na.rm = TRUE),
    sd = sd(QUBIT, na.rm = TRUE),
    n = n()
  )

# By TREATMENT - Boxplot with jitter (QUBIT) - CORRECTED
meanqubitplot=ggplot(qubit_by_sample, aes(x = TREATMENT, y = QUBIT, fill = TREATMENT)) +
  geom_violin(alpha = 0.5, scale = "width") +  
  geom_boxplot(width = 0.2, alpha = 0.7) +     
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.8) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 2, fill = "white") +
  labs(x = "",
    y = "DNA Concentration (ng/μL)"
  ) +
  My_Theme +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) + 
  scale_fill_manual(values = cols.method) +
  scale_y_log10(
    breaks = c(0.01, 0.1, 1, 10, 100, 1000),
    labels = scales::label_number(accuracy = 0.01)
  ) 

meanqubitplot
```

#### Model

```{r}
library(dplyr)
library(FSA)
library(rcompanion)
library(multcompView)

kruskal_qubit <- kruskal.test(QUBIT ~ TREATMENT, data = qubit_by_sample)
print(kruskal_qubit)

dunn_qubit <- dunnTest(QUBIT ~ TREATMENT, 
                         data = qubit_by_sample, 
                         method = "bh")
print(dunn_qubit)

# Create compact letter display
treatments <- unique(qubit_by_sample$TREATMENT)
p_matrix <- matrix(1, nrow = length(treatments), ncol = length(treatments),
                   dimnames = list(treatments, treatments))

for (i in 1:nrow(dunn_qubit$res)) {
  comp <- strsplit(dunn_qubit$res$Comparison[i], " - ")[[1]]
  p_val <- dunn_qubit$res$P.adj[i]
  p_matrix[comp[1], comp[2]] <- p_val
  p_matrix[comp[2], comp[1]] <- p_val
}
diag(p_matrix) <- 1

library(multcompView)
cld <- multcompLetters(p_matrix, threshold = 0.05)
letters_df <- data.frame(
  TREATMENT = names(cld$Letters),
  letters = cld$Letters
)
print(letters_df)
```


### Read abundance

#### Plot

```{r}
# By Run
aggregate(abundance ~ Run, data = data_long, FUN = sum)
# or with dplyr
library(dplyr)
data_long %>% group_by(Run) %>% summarise(total_abund = sum(abundance))

# By Treatment
data_long %>% group_by(TREATMENT) %>% summarise(total_abund = sum(abundance))

# By Filtration
data_long %>% group_by(FILTRATION) %>% summarise(total_abund = sum(abundance))

# By Extraction
data_long %>% group_by(EXTRACTION) %>% summarise(total_abund = sum(abundance))

# By all four factors
data_long %>% 
  group_by(TREATMENT, FILTRATION, EXTRACTION, Run) %>% 
  summarise(
    total_abund = sum(abundance),
    mean_abund = mean(abundance),
    n = n()
  ) %>% 
  arrange(desc(total_abund))

# Sum abundance across treatments and runs
data_long %>%
  group_by(TREATMENT, Run) %>%
  summarise(total_abund = sum(abundance)) %>%
  pivot_wider(names_from = Run, values_from = total_abund, values_fill = 0)

# Overall summary by Treatment
data_long %>%
  group_by(TREATMENT) %>%
  summarise(
    total = sum(abundance),
    mean = mean(abundance),
    median = median(abundance),
    min = min(abundance),
    max = max(abundance),
    sd = sd(abundance),
    n = n()
  )

data_long <- data_long %>%
  mutate(Run = as.factor(Run))

#  By TREATMENT - Boxplot with jitter
meanreadabund=ggplot(data_long, aes(x = TREATMENT, y = abundance, fill = TREATMENT)) +
  geom_violin(alpha = 0.5, scale = "width") +  # Added this line
  geom_boxplot(width = 0.2, alpha = 0.7) +     # Changed width from default to 0.2
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.8) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 2, fill = "white") +
  scale_y_log10() +
  labs(x = "",
    y = "Read Abundance"
  ) +
  My_Theme +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  scale_fill_manual(values = cols.method)  +
  scale_y_log10(
    breaks = c(0.01, 0.1, 1, 10, 100, 1000, 10000),
    labels = scales::label_number(accuracy = 1)
  )

meanreadabund
```


#### Model

```{r}
# Alternative method using multcomp
library(multcomp)
data_long$TREATMENT <- as.factor(data_long$TREATMENT)

# Fit the model
nb_model <- glm.nb(abundance ~ TREATMENT, data = data_long)

# Create contrast matrix for all pairwise comparisons
tukey_contrasts <- glht(nb_model, linfct = mcp(TREATMENT = "Tukey"))

# Get compact letter display
cld_result <- cld(tukey_contrasts, level = 0.05)
print(cld_result)

# Extract letters as data frame
letters_df_multcomp <- data.frame(
  TREATMENT = names(cld_result$mcletters$Letters),
  letters = cld_result$mcletters$Letters
)
print(letters_df_multcomp)
```


### Contamination

#### ASVs

```{r}
removed.reads.data=reads.removed

# Get sample column names
sample_cols <- names(removed.reads.data)[17:ncol(removed.reads.data)] 

# Reshape to long format
removed_long <- removed.reads.data %>%
  # Keep OTU_ID and taxonomic info for reference (optional)
  select(OTU_ID, Kingdom, Phylum, Class, Order, Family, Genus, Species, all_of(sample_cols)) %>%
  # Pivot to long format
  pivot_longer(
    cols = all_of(sample_cols),
    names_to = SAMPLE,
    values_to = "count"
  ) %>%
  # Filter out NA and zero counts (ASVs not present)
  filter(!is.na(count) & count > 0)

# Extract sample-treatment mapping from your existing data_long
sample_treatment_map <- data_long %>%
  select(Sample_ID, TREATMENT,SITE) %>%
  distinct()

# Join with removed_long
removed_long <- removed_long %>%
  left_join(sample_treatment_map, by = SAMPLE)

removed_long=removed_long%>%
  drop_na(TREATMENT)

# Count ASVs per sample (richness)
asv_richness <- removed_long %>%
  group_by(Sample_ID, TREATMENT) %>%
  summarise(
    n_ASVs = n_distinct(OTU_ID),  # Number of unique ASVs present
    total_reads = sum(count),      # Optional: total contaminant reads
    .groups = 'drop'
  )

# View
head(asv_richness)

# Boxplot by treatment
ggplot(asv_richness, aes(x = TREATMENT, y = n_ASVs, fill = TREATMENT)) +
  geom_violin(alpha = 0.5, scale = "width") +
  geom_boxplot(width = 0.2, alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.8) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 2, fill = "white") +
  labs(
    x = "",
    y = "Contnaminant ASVs"
  ) +
  My_Theme +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  scale_fill_manual(values = cols.method)

# Summary statistics by treatment
asv_richness %>%
  group_by(TREATMENT) %>%
  summarise(
    n_samples = n(),
    mean_ASVs = round(mean(n_ASVs), 1),
    sd_ASVs = round(sd(n_ASVs), 1),
    median_ASVs = median(n_ASVs),
    min_ASVs = min(n_ASVs),
    max_ASVs = max(n_ASVs)
  )
```

#### Reads

```{r}
# 1. Calculate total contaminant reads per sample
contaminant_reads <- removed_long %>%
  group_by(Sample_ID, TREATMENT) %>%
  summarise(
    total_contaminant_reads = sum(count, na.rm = TRUE),
    n_contaminant_ASVs = n_distinct(OTU_ID),
    .groups = 'drop'
  )

# View the data
head(contaminant_reads)

# 2. Summary statistics by treatment
contaminant_reads %>%
  group_by(TREATMENT) %>%
  summarise(
    n_samples = n(),
    mean_reads = round(mean(total_contaminant_reads), 1),
    sd_reads = round(sd(total_contaminant_reads), 1),
    median_reads = median(total_contaminant_reads),
    min_reads = min(total_contaminant_reads),
    max_reads = max(total_contaminant_reads),
    mean_ASVs = round(mean(n_contaminant_ASVs), 1),
    median_ASVs = median(n_contaminant_ASVs)
  )

readsremoved=ggplot(contaminant_reads, aes(x = TREATMENT, y = total_contaminant_reads, fill = TREATMENT)) +
  geom_violin(alpha = 0.5, scale = "width") +
  geom_boxplot(width = 0.2, alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.8) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 2, fill = "white") +
  scale_y_log10(
    breaks = c(1, 10, 100, 1000, 10000),
    labels = scales::label_number(accuracy = 1)
  ) +
  labs(
    x = "",
    y = "Reads Removed"
  ) +
  My_Theme +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  scale_fill_manual(values = cols.method)

readsremoved
```

#### Model

```{r}
library(glmmTMB)
contamload.mod=glmmTMB(count ~ TREATMENT + (1|SITE),
        family=nbinom2,
        data=removed_long)
contamload.mod
summary(contamload.mod)

# Install if needed
# install.packages("emmeans")

library(emmeans)

# Get estimated marginal means for treatments
emm <- emmeans(contamload.mod, ~ TREATMENT)

# Get pairwise comparisons with CLD
cld_results <- cld(emm, 
                   Letters = letters,        # Use lowercase letters
                   adjust = "tukey",         # Multiple comparison adjustment
                   alpha = 0.05)             # Significance level

# View results
print(cld_results)

# Alternative with more options
cld(emm, 
    Letters = letters,
    adjust = "tukey",
    alpha = 0.05,
    reversed = TRUE,        # Reverse letter order if desired
    sort = TRUE)            # Sort by estimate
```

### Plotted together

```{r}
dnaabundplot=ggarrange(meanqubitplot, meanreadabund, readsremoved, nrow = 3, labels = "AUTO")

ggsave(
  filename=file.path(paths$results_figs,paste0("DNA_abundance_", GROUPING,".pdf")),
  plot    =dnaabundplot,
  width   =7,
  height  =14,
  units   ="in",
  device  =cairo_pdf,
  family  ="Arial"
)
```



## 3. Community dissimilarity

This code calculates community dissimilarity using a chosen distance method (e.g., Bray-Curtis), then runs non-metric multidimensional scaling (NMDS) to visualize how samples cluster in 2D space, with ellipses showing 95% confidence intervals for each group. It then uses envfit to identify which environmental or metadata variables are significantly correlated with the community patterns (p < 0.05), adding them as arrows to the NMDS plot to show direction and strength of influence.


```{r}
table.dist<- vegdist(community.table.t, method=distance.method, binary=F) # distance matrix from community OTU table

metatable.reps$group_factor <- as.factor(metatable.reps[[GROUPING]]) # grouping variable for ordination

PDAT_MiBird=read.csv(file = file.path(paths$data_processed, "NMDS_data_MiBird.csv"), header=T)
PDAT_MiBird$Marker="MiBird"
PDAT_Teleo2=read.csv(file = file.path(paths$data_processed, "NMDS_data_Teleo2.csv"), header=T)
PDAT_Teleo2$Marker="Teleo2"

PDAT=rbind(PDAT_MiBird, PDAT_Teleo2)

facet_labels <- c(
  "MiBird" = "MiBird (Stress = 0.08)",
  "Teleo2" = "Teleo2 (Stress = 0.15)"
)

# Rename the Marker factor
PDAT$Marker_label <- factor(PDAT$Marker, 
                            levels = c("MiBird", "Teleo2"),
                            labels = c("MiBird (Stress = 0.08)", "Teleo2 (Stress = 0.15)"))

nmds.plot <- ggplot(PDAT, aes(x = NMDS1, y = NMDS2, colour = group_factor)) +
  geom_vline(xintercept = 0, lty = 2, col = "grey66") + 
  geom_hline(yintercept = 0, lty = 2, col = "grey66") + 
  geom_point(alpha = 0.6, size = 2) +
  stat_ellipse(aes(fill = group_factor),
               level = 0.95, 
               geom = "polygon", 
               alpha = 0.15,
               colour = NA) +
  facet_wrap(~ Marker_label, scales = "free", ncol = 2) +
  My_Theme +  
  labs(colour = NULL, 
       fill = NULL) + 
  scale_color_manual(values = cols.method) +
  scale_fill_manual(values = cols.method)
nmds.plot

ggsave(
  filename=file.path(paths$results_figs,paste0("NMDSplot_", GROUPING,".pdf")),
  plot    =nmds.plot,
  width   =9,
  height  =4,
  units   ="in",
  device  =cairo_pdf,
  family  ="Arial"
)
```


## 4. Hill diversity

This code calculates Hill diversity (richness at q=0, Shannon at q=1, and Simpson at q=2) for each sample using the iNEXT package, then standardizes all samples to the same coverage level (95% of the minimum coverage) for fair comparison. It produces two visualizations: group comparison plots showing diversity across your grouping variable (e.g., sites or treatments) using violin-boxplots, and a richness over time plot with a dashed line connecting medians across ordered dates.

#### Calculation

```{r}
# Remove any species (rows) that were never detected
table_pa_clean <- table_pa[rowSums(table_pa) > 0, ]

# Remove any sites (columns) with no species (optional but recommended)
table_pa_clean <- table_pa_clean[, colSums(table_pa_clean) > 0]

# Convert to a matrix to strip any problematic attributes
table_pa_matrix <- as.matrix(table_pa_clean)

# Remove any lingering attributes
attributes(table_pa_matrix) <- list(
  dim = dim(table_pa_matrix),
  dimnames = dimnames(table_pa_matrix)
)

# Get the number of sites (samples)
n_sites <- ncol(table_pa_clean)

# Calculate incidence frequencies (number of sites where each species occurs)
incidence_freq <- rowSums(table_pa)

# Remove species with zero incidence (not detected anywhere)
incidence_freq <- incidence_freq[incidence_freq > 0]

# Create the correct data structure for iNEXT
# First element = number of sites, then all incidence frequencies
inext_data <- c(n_sites, incidence_freq)

# Now run iNEXT with datatype = "incidence_freq"
out <- iNEXT::iNEXT(
  x = inext_data,
  datatype = "incidence_freq",
  q = 0)

# Check results
ggiNEXT(out, type = 3) + My_Theme

# Your data has one assemblage, so the coverage is just the single value
coverage_info <- DataInfo(table_pa_clean, datatype = "incidence_raw")
target_coverage <- coverage_info$SC 
print(paste("Target coverage:", target_coverage))
```


### By treatment

#### ASV level

```{r}
# Get treatment information
metadata_subset <- metadata[metadata$Sample_ID %in% colnames(table_pa_clean), ]
metadata_subset <- metadata_subset[match(colnames(table_pa_clean), metadata_subset$Sample_ID), ]

# Get treatments in alphabetical order (same as ggiNEXT)
treatments_sorted <- sort(treatments)

# Name the colors with the alphabetical treatment names
names(cols.method) <- treatments_sorted

# Prepare data for each treatment as a list
treatment_data <- list()

for (treatment in treatments) {
  # Get sample IDs for this treatment
  sample_ids <- metadata_subset$Sample_ID[metadata_subset$TREATMENT == treatment]
  sample_ids <- sample_ids[sample_ids %in% colnames(table_pa_clean)]
  
  if (length(sample_ids) < 2) {
    warning(paste("Treatment", treatment, "has too few samples, skipping"))
    next
  }
  
  # Subset the incidence matrix for this treatment
  table_pa_treatment <- table_pa_clean[, sample_ids, drop = FALSE]
  table_pa_treatment <- table_pa_treatment[rowSums(table_pa_treatment) > 0, , drop = FALSE]
  table_pa_treatment <- table_pa_treatment[, colSums(table_pa_treatment) > 0, drop = FALSE]
  
  if (ncol(table_pa_treatment) < 2 || nrow(table_pa_treatment) < 2) {
    warning(paste("Treatment", treatment, "has too few sites or species, skipping"))
    next
  }
  
  # Convert to matrix
  table_pa_matrix <- as.matrix(table_pa_treatment)
  
  # Calculate incidence frequencies
  incidence_freq <- rowSums(table_pa_matrix)
  n_sites <- ncol(table_pa_matrix)
  
  # Create iNEXT data structure - first value is number of sites
  treatment_data[[treatment]] <- c(n_sites, incidence_freq)
}

# Run iNEXT on all treatments simultaneously
if (length(treatment_data) > 0) {
  out_combined <- iNEXT::iNEXT(
    x = treatment_data,
    datatype = "incidence_freq",
    q = 0,
    se = TRUE)
  
  # Get treatment names in the order they appear in the plot (alphabetical)
  treatment_names_in_plot <- sort(names(treatment_data))
  
  # Subset colors to only include treatments that exist
  cols_method_subset <- cols.method[names(cols.method) %in% treatment_names_in_plot]
  
  # Reorder colors to match the alphabetical order
  cols_method_ordered <- cols_method_subset[treatment_names_in_plot]
  
  # Print color assignment to verify
  print("Color assignment:")
  print(cols_method_ordered)
  
  # Plot 1: Sample-size-based R/E curves
  p1 <- ggiNEXT(out_combined, type = 1, color.var = "Assemblage") +
    theme_bw(base_size = 14) +
    labs(x = "Sample effort", 
         y = "ASV richness (q = 0)") +
    theme(legend.position = "right",
          legend.title = element_blank()) +
    scale_color_manual(values = cols_method_ordered) +
    scale_fill_manual(values = cols_method_ordered) +ylim(0,250)
  
  # Plot 2: Sample completeness curves
  p2 <- ggiNEXT(out_combined, type = 2, color.var = "Assemblage") +
    theme_bw(base_size = 14) +
    labs(x = "Sample effort", 
         y = "Sample coverage") +
    theme(legend.position = "right",
          legend.title = element_blank()) +
    scale_color_manual(values = cols_method_ordered) +
    scale_fill_manual(values = cols_method_ordered)
  
  # Plot 3: Coverage-based R/E curves
  p3 <- ggiNEXT(out_combined, type = 3, color.var = "Assemblage") +
    theme_bw(base_size = 14) +
    labs(x = "Sample coverage", 
         y = "ASV richness (q = 0)") +
    theme(legend.position = "right",
          legend.title = element_blank()) +
    scale_color_manual(values = cols_method_ordered) +
    scale_fill_manual(values = cols_method_ordered) + xlim(0,1)
  
  # Print plots
  print(p1)
  print(p2)
  print(p3)
 } else {
  print("No treatments were successfully processed.")
}

```

#### Genus level

```{r}
# Extract the Genus column from final_pa
# Identify the Genus column (adjust column name if needed)
genus_col <- grep("Genus|genus|g__", colnames(final_pa), value = TRUE)
if (length(genus_col) == 0) {
  stop("No Genus column found in final_pa. Please check column names.")
}
print(paste("Using genus column:", genus_col))

# Extract genera as character vector
genera <- as.character(final_pa[[genus_col]])

# Handle NA or empty values
genera[is.na(genera) | genera == ""] <- "Unclassified"

# Check length matches
if (length(genera) != nrow(table_pa)) {
  stop("Length of genera does not match number of rows in table_pa")
}

# Convert table_pa to matrix
table_pa_matrix <- as.matrix(table_pa)

# Aggregate by genus using rowsum (sum across ASVs)
genus_aggregated <- rowsum(table_pa_matrix, group = genera)

# Convert to presence/absence (1 if genus detected in sample)
genus_matrix <- genus_aggregated
genus_matrix[genus_matrix > 0] <- 1

# Remove genera with no occurrences across all samples
genus_matrix <- genus_matrix[rowSums(genus_matrix) > 0, ]

# Store as table_pa_genus
table_pa_genus <- genus_matrix

# Print summary
print(paste("ASV-level: 213 ASVs × 105 samples"))
print(paste("Genus-level:", nrow(table_pa_genus), "genera ×", ncol(table_pa_genus), "samples"))


# 2. PREPARE METADATA FOR GENUS-LEVEL DATA
# ----------------------------------------------------------------------------
# Get treatment information
metadata_subset <- metadata[metadata$Sample_ID %in% colnames(table_pa_genus), ]
metadata_subset <- metadata_subset[match(colnames(table_pa_genus), metadata_subset$Sample_ID), ]

# Get treatments
treatments <- unique(metadata_subset$TREATMENT)
treatments_sorted <- sort(treatments)

names(cols.method) <- treatments_sorted


# 3. PREPARE iNEXT DATA FOR EACH TREATMENT
# ----------------------------------------------------------------------------
treatment_data_genus <- list()

for (treatment in treatments) {
  # Get sample IDs for this treatment
  sample_ids <- metadata_subset$Sample_ID[metadata_subset$TREATMENT == treatment]
  sample_ids <- sample_ids[sample_ids %in% colnames(table_pa_genus)]
  
  if (length(sample_ids) < 2) {
    warning(paste("Treatment", treatment, "has too few samples, skipping"))
    next
  }
  
  # Subset the incidence matrix for this treatment
  table_pa_treatment <- table_pa_genus[, sample_ids, drop = FALSE]
  table_pa_treatment <- table_pa_treatment[rowSums(table_pa_treatment) > 0, , drop = FALSE]
  table_pa_treatment <- table_pa_treatment[, colSums(table_pa_treatment) > 0, drop = FALSE]
  
  if (ncol(table_pa_treatment) < 2 || nrow(table_pa_treatment) < 2) {
    warning(paste("Treatment", treatment, "has too few sites or species, skipping"))
    next
  }
  
  # Convert to matrix
  table_pa_matrix <- as.matrix(table_pa_treatment)
  
  # Calculate incidence frequencies
  incidence_freq <- rowSums(table_pa_matrix)
  n_sites <- ncol(table_pa_matrix)
  
  # Create iNEXT data structure - first value is number of sites
  treatment_data_genus[[treatment]] <- c(n_sites, incidence_freq)
}

# Print summary
print(paste("Number of treatments with data:", length(treatment_data_genus)))
print(names(treatment_data_genus))


# 4. RUN iNEXT ON GENUS-LEVEL DATA
# ----------------------------------------------------------------------------
if (length(treatment_data_genus) > 0) {
  out_combined_genus <- iNEXT::iNEXT(
    x = treatment_data_genus,
    datatype = "incidence_freq",
    q = 0,
    se = TRUE,
    nboot = 50
  )
  
  # Get treatment names in alphabetical order
  treatment_names_in_plot <- sort(names(treatment_data_genus))
  
  # Subset and order colors
  cols_method_subset <- cols.method[names(cols.method) %in% treatment_names_in_plot]
  cols_method_ordered <- cols_method_subset[treatment_names_in_plot]
  
  print("Color assignment for genus-level plots:")
  print(cols_method_ordered)
  
  
  # 5. CREATE GENUS-LEVEL PLOTS
  # --------------------------------------------------------------------------
  # Plot 1: Sample-size-based R/E curves
  p1_genus <- ggiNEXT(out_combined_genus, type = 1, color.var = "Assemblage") +
    theme_bw(base_size = 14) +
    labs(x = "Sample effort", 
         y = "Genus richness (q = 0)") +
    theme(legend.position = "right",
          legend.title = element_blank()) +
    scale_color_manual(values = cols_method_ordered) +
    scale_fill_manual(values = cols_method_ordered)
  
  # Plot 2: Sample completeness curves
  p2_genus <- ggiNEXT(out_combined_genus, type = 2, color.var = "Assemblage") +
    theme_bw(base_size = 14) +
    labs(x = "Sample effort", 
         y = "Sample coverage") +
    theme(legend.position = "right",
          legend.title = element_blank()) +
    scale_color_manual(values = cols_method_ordered) +
    scale_fill_manual(values = cols_method_ordered)
  
  # Plot 3: Coverage-based R/E curves
  p3_genus <- ggiNEXT(out_combined_genus, type = 3, color.var = "Assemblage") +
    theme_bw(base_size = 14) +
    labs(x = "Sample coverage", 
         y = "Genus richness (q = 0)") +
    theme(legend.position = "right",
          legend.title = element_blank()) +
    scale_color_manual(values = cols_method_ordered) +
    scale_fill_manual(values = cols_method_ordered)
  
  
  # 6. DISPLAY PLOTS
  # -----------------------------------------------------------------------
  print(p1_genus)
  print(p2_genus)
  print(p3_genus)
  }

```

####Plotted together

```{r}
coveragerarefaction=ggarrange(p3, p1, p3_genus, p1_genus, nrow=2, ncol=2, labels = "AUTO", common.legend = T, legend = "right")
coveragerarefaction

ggsave(
  filename=file.path(paths$results_figs,paste0("Rarefaction_", GROUPING,".pdf")),
  plot    =coveragerarefaction,
  width   =11,
  height  =9,
  units   ="in",
  device  =cairo_pdf,
  family  ="Arial"
)
```


### Observed vs Estimated

```{r}
info_data=DataInfo(treatment_data, datatype="incidence_freq") # method coverages
info_data

coverage_data=estimateD(
    treatment_data,
    datatype = "incidence_freq",
    base = "coverage",
    level = 0.5,
    q = 0
)

coverage_data=merge(coverage_data, info_data, by="Assemblage")

hill_df_cov=coverage_data %>% 
  dplyr::select(Assemblage, Order.q, qD, qD.UCL, qD.LCL, S.obs, SC.y) %>%
  tidyr::pivot_wider(
    names_from=Order.q,
    values_from=qD,
    names_prefix="Hill_q"
  )
hill_df_cov


#hill_df_cov$Assemblage <- factor(hill_df_cov$Assemblage, levels = names(cols.method))

ggplot(hill_df_cov,
       aes(x = Hill_q0,
           y = Assemblage)) +
  geom_errorbarh(
    aes(xmin = qD.LCL,
        xmax = qD.UCL,
        color = Assemblage),
    height = 0.2,
    linewidth = 0.8
  ) +
  geom_point(
    aes(color = Assemblage),
    size = 3
  ) +
  scale_color_manual(values = cols.method) +
  theme_bw(base_size = 14) +theme(legend.position = "none") +
  labs(
    x = "Estimated ASV richness (q = 0)",
    y = NULL,
    color = NULL
  )

plot_df <- hill_df_cov %>%
  select(Assemblage, S.obs, Hill_q0, qD.LCL, qD.UCL) %>%
  pivot_longer(
    cols = c(S.obs, Hill_q0),
    names_to = "Estimate",
    values_to = "Richness"
  ) %>%
  mutate(
    Estimate = factor(
      Estimate,
      levels = c("S.obs", "Hill_q0"),
      labels = c("Observed richness", 
                 "Coverage-standardized richness")
    )
  )

plot_df$Assemblage <- factor(
  plot_df$Assemblage,
  levels = hill_df_cov$Assemblage[order(hill_df_cov$Hill_q0)]
)

coveragevsobserved=ggplot(plot_df,
       aes(x = Richness,
           y = Assemblage,
           color = Assemblage,
           shape = Estimate)) +
  geom_errorbarh(
    data = hill_df_cov,
    aes(
      y = reorder(Assemblage, Hill_q0),
      xmin = qD.LCL,
      xmax = qD.UCL
    ),
    height = 0.15,
    linewidth = 0.5,
    inherit.aes = FALSE
  ) +
  geom_point(
    size = 3
  ) +
  scale_color_manual(values = cols.method, guide = "none") +
  scale_shape_manual(
    values = c(
      "Observed richness" = 16,
      "Coverage-standardized richness" = 17
    )
  ) +
  theme_bw(base_size = 14) +
  labs(
    x = "ASV richness (q = 0)",
    y = NULL,
    color = NULL,
    shape = NULL
  ) + theme(legend.position = "top")


ggsave(
  filename=file.path(paths$results_figs,paste0("richness_obs_vs_est_", GROUPING, ".pdf")),
  plot    =coveragevsobserved,
  width   =8,
  height  =5,
  units   ="in",
  device  =cairo_pdf,
  family  ="Arial"
)
```


## 5. Sampling effort

This code creates two types of accumulation curves: a sequence depth rarefaction curve showing how many distinct organisms (MOTUs) you detect as you sequence more reads per sample, and sample rarefaction curves showing how taxonomic coverage increases as you include more samples. The sample curves are generated both for all data combined and separately for different groups (defined by your GROUPING variable, like sites or treatments), using the iNEXT package to interpolate and extrapolate the curves with confidence intervals.

### Sequence depth rarefaction

```{r}
rarecurve(table.reads.t, step = 1000, col = rainbow(nrow(table.reads.t)), label = F, xlab = "Sequence Depth",  ylab = "MOTUs")
```

## 6. Indicator taxa
This code aggregates OTUs by a higher taxonomic rank (e.g., genus or family), then runs indicator species analysis (IndVal) to identify which taxa are significantly associated with specific groups (e.g., sites or treatments). It filters significant indicators (adjusted p < 0.15), then creates a faceted time-series plot showing how the abundance of each indicator taxon changes over time, with separate panels for each group and each indicator taxon, using violin-boxplots and a dashed line connecting medians.

### Aggregate OTU table

```{r}
group=as.factor(metatable.reps[[GROUPING]])
taxonomic.rank=community.data[[TAXONOMICRANK]] 

com=community.table.t

colnames(com)=taxonomic.rank

prefixes = unique(sub("\\..*", "", colnames(com)))
prefixes = prefixes[prefixes != ""]

com=sapply(prefixes, function(x) rowSums(com[,startsWith(colnames(com), x), drop=F])) ##OTU table with combined taxonomies (no repeated columns)

com=as.data.frame(com)

ncom=ncol(com)
```


## 7. Heatmap

### Pooled

```{r}
# ============================================
# HEATMAP - LOG-TRANSFORMED SUMMED READS BY TREATMENT
# ============================================

library(pheatmap)

# Match samples
common_samples <- intersect(rownames(com), metadata[[SAMPLE]])
com_subset <- com[common_samples, ]
metadata_subset <- metadata[match(common_samples, metadata[[SAMPLE]]), ]

# Sum reads by treatment
sum_by_treatment <- aggregate(com_subset, 
                              by = list(Treatment = metadata_subset$TREATMENT), 
                              FUN = sum)

# Set treatment as rownames and remove the Treatment column
rownames(sum_by_treatment) <- sum_by_treatment$Treatment
sum_by_treatment <- sum_by_treatment[, -1]

# Transpose so taxa are rows, treatments are columns
heatmap_data <- t(sum_by_treatment)

# Log transform (log1p = log(1+x) to handle zeros)
heatmap_data_log <- log1p(heatmap_data)


# Create heatmap (no scaling)
heatmap=pheatmap(heatmap_data_log,
         scale = "none", fontsize_row = 8,
         fontsize_col = 10,
         angle_col = 45,
         cellwidth = 30,
         cellheight = 10,
         show_rownames = TRUE,
         show_colnames = TRUE,
         cluster_rows = T,
         cluster_cols = F,
         color = colorRampPalette(c("white", "tan", "firebrick3"))(100))
heatmap

ggsave(
  filename=file.path(paths$results_figs,paste0("heatmap_", GROUPING, ".pdf")),
  plot    =heatmap,
  width   =6,
  height  =12,
  units   ="in",
  device  =cairo_pdf,
  family  ="Arial")
```
