---
title: "MM eDNA (12S)"
author: "Patrick Nichols"
date: "Spring 2026"
output:
  html_document: 
    fig_width: 8 
    fig_height: 6
---

```{r setup, include=FALSE}
library(knitr)
#if (!requireNamespace("BiocManager", quietly = TRUE)) 
#install.packages("BiocManager")
#BiocManager::install("phyloseq", dependencies=T)
#BiocManager::install("rblast", dependencies=T)
#BiocManager::install("decontam")

#devtools::install_github("tidyverse/ggplot2")
library("ggplot2")

#install.packages("remotes")
#remotes::install_github("MadsAlbertsen/ampvis2")
#remotes::install_github("GuillemSalazar/EcolUtils")

#library("phyloseq")
library(Matrix)
library(lattice)
library(xtable)
library(effects)
library(car)
library(MASS)
library(pscl)
library(lmtest)
library(boot)
library(MuMIn)
library(mgcv)
library(lme4)
library(lmerTest)
library(pbkrtest)
library(sp)
library(vegan)
library(MCMCglmm)
library(RColorBrewer)
library(plotrix)
library(gridExtra)
library(gam)
library(ade4)
library(rsq)
library(R2admb)
library(plyr)
library(tidyr)
library(reshape2)
library(xts)
library(VennDiagram)
library(venn)
library(gplots)
library(vegan3d)
library(forcats)
library(ecodist)
library(funrar)
library(GUniFrac)
#library(unionplot)
library(dplyr)
library(magrittr)
library(multcomp)
#library(RAM)
#library(EcolUtils)
library(ggsci)
library(maps)
library(mapdata)
library(coefplot)
library(tidyverse)
library(ggfortify)
library(data.table)
library(microDecon)
library("ggspatial")
library(ggrepel)
#install.packages("indicspecies")
library(indicspecies)
library(zetadiv)
library(ggpubr)
library(tibble)
library(purrr)
library(ggstats)
library(ape)
library(microDecon)
library(iNEXT)
library(glmmTMB)
library(TITAN2)
library(beepr)
library(scales)
library(lunar)
library(climate)
library(readxl)

#if (!requireNamespace("BiocManager", quietly = TRUE))
#install.packages("BiocManager")
#BiocManager::install("ggtree")
library(ggtree)
#install.packages("devtools") #Installs devtools (if not already installed)
#devtools::install_github("donaldtmcknight/microDecon") #Installs microDecon

# install bioconductor dependencies
#install.packages("BiocManager")
#BiocManager::install("biomformat")

# install metabaR package
#install.packages("remotes")
#remotes::install_github("metabaRfactory/metabaR")


st.err <- function(x) {
    sd(x)/sqrt(length(x))
}

My_Theme = theme_set(theme_bw()+theme(
  panel.grid.major = element_blank(), 
  panel.grid.minor = element_blank(),
  panel.background = element_blank(), 
  panel.border = element_rect(fill=NA, colour = "black"), 
  panel.spacing = unit(1, "lines"),
  axis.title.x = element_text(size = 12, color="black"),
  axis.text.x = element_text(size = 12, color="black"),
  axis.title.y = element_text(size = 12, color = "black"),
  axis.text.y = element_text(size=12, color = "black"),
  legend.text = element_text(size=12, color = "black"),
  strip.text.x = element_text(size=14, color="black"),
  strip.text.y = element_text(size=14, color="black"),
  strip.text = element_text(size = 14, color = "black"),
  strip.background = element_blank(), 
  strip.placement = "outside",
  title = element_text(size=14, color="black", face="bold"),
  axis.title = element_text(size=12, face="bold", colour = "black"), 
  axis.ticks = element_blank(), 
  axis.text = element_blank()))

My_Theme = theme_set(theme_bw()+theme(
  panel.grid.major = element_blank(), 
  panel.grid.minor = element_blank(),
  panel.background = element_blank(), 
  panel.border = element_rect(fill=NA, colour = "black"), 
  panel.spacing = unit(1, "lines"),
  axis.title.x = element_text(size = 12, color="black"),
  axis.text.x = element_text(size = 12, color="black"),
  axis.title.y = element_text(size = 12, color = "black"),
  axis.text.y = element_text(size=12, color = "black"),
  legend.text = element_text(size=12, color = "black"),
  strip.text.x = element_text(size=14, color="black"),
  strip.text.y = element_text(size=14, color="black"),
  strip.text = element_text(size = 14, color = "black"),
  strip.background = element_blank(), 
  strip.placement = "outside",
  title = element_text(size=14, color="black", face="bold"),
  axis.title = element_text(size=12, face="bold", colour = "black"), 
  axis.ticks = element_blank(), 
  axis.text = element_blank()))

cols.method=c("#031d44", "#1a659e", "#d5896f", "#dab785", "#70a288", "snow3", "black")

heatmap.cols=colorRampPalette(c("white", "#be95c4", "#231942"))(30)

numperm=9999 #zeta.mc
numsam=9999 #permanova, anosim

#### SPECIFY FOR EACH ANALYSIS ############################
marker.name="Vertebrates (12S)"
taxonomic.rank="genus" 
failed.samples="" 
# Set export directory relative to current working directory
export_dir <- file.path(getwd(), "vertebrates")
core_threshold <- 0.8 # define proportion of sites must be present in
############################################################

# Create folder if it doesn't exist
if (!dir.exists(export_dir)) {
  dir.create(export_dir, recursive = TRUE)
}
```


# 1. Data Setup

## A. Import Data

```{r}
####IMPORT METADATA#####################################
metadata=read_excel("data/raw/metadata.xlsx")
metadata <- unique(metadata)
metadata$TREATMENT=as.factor(metadata$TREATMENT)

####METADATA BY SITE-LEVEL
metadata_site <- metadata %>%
  group_by(SITE_ID) %>%
  summarise(
    REGION  = first(REGION),
    LAT     = first(LAT),
    LON     = first(LON),
    .groups = "drop"
  )

#### Add BLAST information
blast.12S=read.delim("12S_blast.tsv", header = F, sep = "\t")
colnames(blast.12S) <- c(
  "zotu",      # V1  - query ID (Zotu)
  "sseqid",      # V2  - subject sequence ID
  "staxids",     # V3
  "sScientific", # V4  - scientific name
  "sCommon",     # V5  - common name
  "sKingdom",    # V6
  "pident",      # V7
  "length",      # V8
  "mismatch",    # V9
  "bitscore",     # V10
  "qstart",      # V11
  "qend",        # V12
  "sstart",      # V13
  "send",        # V14
  "sstrand",      # V15
  "gapopen",    # V16
  "qcovs",       # V17  - query coverage
  "stitle",      # V18
  "evalue",     # V19
  "scov",        # V20
  "slen",
  "unk"# V21
)
blast.12S$evalue <- as.numeric(blast.12S$evalue)

best_hits <- blast.12S %>%
  group_by(zotu) %>%
  arrange(evalue, desc(bitscore)) %>%   # lowest evalue, then highest bitscore
  slice(1)

####Keep only species names in BLAST result
best_hits <- best_hits %>%
  mutate(
    # 0. Remove brackets around genus names
    sScientific = str_replace_all(sScientific, "\\[|\\]", ""),

    # 0b. Set purely environmental / higher taxonomic placeholders to NA
    sScientific = ifelse(
      str_detect(sScientific, "(?i)environmental|Foraminifera|Metazoa|uncultured"),
      NA_character_,
      sScientific
    ),

    # 1. Remove "uncultured..." completely (extra safety)
    sScientific = str_replace(sScientific, "^uncultured.*", NA_character_),

    # 2. Remove strain / voucher indicators
    sScientific = str_replace(sScientific, "\\bstr\\.?\\s*[A-Za-z0-9._-]+", ""),
    sScientific = str_replace(sScientific, "\\bstrain\\s*[A-Za-z0-9._-]+", ""),
    sScientific = str_replace(sScientific, "\\bDSM\\s*[0-9]+", ""),
    sScientific = str_replace(sScientific, "\\bMIT\\s*[0-9]+", ""),
    sScientific = str_replace(sScientific, "=.*", ""),

    # 3. Remove subtype/serovar/genomosp
    sScientific = str_replace(sScientific, "\\bsubsp\\.\\s*\\S+", ""),
    sScientific = str_replace(sScientific, "\\bserovar\\s*\\S+", ""),
    sScientific = str_replace(sScientific, "\\bgenomosp\\.\\s*\\S+", ""),

    # 4. Remove taxonomic noise terms
    sScientific = str_replace(sScientific, "\\b(bacterium|archaeon|microorganism|eukaryote)\\b", ""),

    # 5. Remove cf., aff., nr. before species epithet
    sScientific = str_replace(sScientific, "\\b(cf|aff|nr)\\.\\s*", ""),

    # 6. Clean whitespace
    sScientific = str_squish(sScientific),

    # 7. Remove trailing codes ONLY if 3+ words OR last token is not lowercase (species)
    sScientific = ifelse(
      str_count(sScientific, "\\S+") >= 3 |
        !str_detect(sScientific, "\\b[a-z]+$"),
      str_replace(sScientific, "\\s+\\S+$", ""),
      sScientific
    ),

    # 8. Collapse any "sp" / "sp." after genus to just genus
    sScientific = ifelse(
      str_detect(sScientific, "^[A-Z][a-z]+\\s+sp\\b\\.?(\\s|$)"),
      word(sScientific, 1),
      sScientific
    ),

    # 9. If name is Genus + species + extra → keep binomial
    sScientific = ifelse(
      str_detect(sScientific, "^[A-Z][a-z]+\\s+[a-z]+\\s+"),
      str_extract(sScientific, "^[A-Z][a-z]+\\s+[a-z]+"),
      sScientific
    ),

    # 10. Final whitespace cleanup
    sScientific = str_squish(sScientific)
  )

####CALCULATE DISTANCE FROM BLACK POINT####
library(geosphere)
ref_point <- c(-157.923681, 21.312027) #HNL
metadata$DISTANCE_m <- distGeo(p1 = ref_point, p2 = metadata[, c("LON", "LAT")])
metadata$DISTANCE_km=metadata$DISTANCE_m/1000

####IMPORT OTU TABLE####
library(stringr)  # for str_squish and str_extract

otutab <- read.delim("12S_data.tsv", header = TRUE, sep = "\t")
old_names <- names(otutab)

library(stringr)

old_names <- names(otutab)

new_names <- sapply(old_names, function(nm) {
  # 1) clean whitespace
  nm <- str_squish(nm)

  # 2) look for retain tokens
  # NTC_#
  ntc_match <- str_extract(nm, "NTC_[0-9]+")
  if (!is.na(ntc_match)) return(sub("_$", "", ntc_match))

  # EXT_Neg_XXX
  extneg_match <- str_extract(nm, "EXT_Neg_[A-Za-z0-9]{3}")
  if (!is.na(extneg_match)) return(sub("_$", "", extneg_match))

  # PCRPos or PCRPos_id
  pcrpos_match <- str_extract(nm, "PCRPos(?:_[A-Za-z0-9]+)?")
  if (!is.na(pcrpos_match)) return(sub("_$", "", pcrpos_match))

  # CBA, CBB, CBC, CB1, CB2, CB3 or any CB
  cb_match <- str_extract(nm, "CB[A-Za-z0-9]*")
  if (!is.na(cb_match)) return(sub("_$", "", cb_match))

  # 3) MM rule
  if (grepl("MM", nm)) {
    out <- sub("^[^_]*_([^_]*)_.*", "\\1", nm)
    return(sub("_$", "", out))
  }

  # 4) MK rule
  if (grepl("^MK", nm)) {
    out <- substr(nm, 1, 6)
    return(sub("_$", "", out))
  }

  # 5) fallback
  sub("_$", "", sub("\\s+.*$", "", nm))
})

names(otutab) <- new_names

data.unfiltered=subset(otutab, phylum != "dropped")
```

## B. Data Cleaning

### i.Decontamination

Data must be in format of:
  -col 1: OTU ID list
  -col 2-x: blanks in consecutive columns
  -col x-y: biological samples in consecutive columns
  -col z: taxonomic rank in single column (optional)


```{r}
df <- data.unfiltered    # rename for convenience

df <- df[ , !grepl("PCRPos", names(df)) ] # remove PCR positive sample

# original first 11 metadata columns
meta_cols <- names(df)[1:11]

taxonomic.zotus=df[,meta_cols]
taxonomic.info <- taxonomic.zotus %>%
  left_join(best_hits, by = "zotu")

# columns matching collection blanks "CBA, CBB, CBC"
cbc_cols <- grep("CB", names(df), value = TRUE)
# NTC columns
ntc_cols <- grep("NTC", names(df), value = TRUE)
# EXT_Neg columns
extneg_cols <- grep("EXT_Neg", names(df), value = TRUE)
# all columns that match any of the "control" patterns
control_cols <- unique(c(cbc_cols, ntc_cols, extneg_cols))
# remaining sample columns after removing meta & priority
sample_cols <- setdiff(names(df), c(meta_cols, control_cols))

microdecon_order <- c(control_cols, sample_cols, meta_cols)
df.microdecon.reordered <- df[ , microdecon_order ]
# Move Zotu to the very first column
df.microdecon.reordered <- df.microdecon.reordered[, c("zotu", setdiff(names(df.microdecon.reordered), "zotu"))]
df.microdecon <- df.microdecon.reordered[ , -( (ncol(df.microdecon.reordered)-9) : ncol(df.microdecon.reordered) ) ] # Remove taxonomic data columns for microdecon

controls=as.numeric(length(control_cols))
samples=as.numeric(length(sample_cols))

microdecon.decontaminated=decon(df.microdecon, numb.blanks = controls, numb.ind = samples, taxa = F, thresh = 1, prop.thresh = 0, runs = 1)
#Setting prop.thresh = 0 and thresh = 1 will avoid whole-OTU deletion, but small real contaminants may remain in the final table. You trade stringency for sensitivity (preserve rare true positives).

microdecon.removed=microdecon.decontaminated$OTUs.removed
decontaminated.table=microdecon.decontaminated$decon.table
decontaminated.table$Mean.blank=NULL

decontaminated.table=merge(decontaminated.table, taxonomic.info, by="zotu")
```

### ii. Filter metazoans

```{r}
####FIX ALGAE ################################
decontaminated.table$phylum <- ifelse(
  decontaminated.table$class == "Phaeophyceae",
  "Ochrophyta",
  decontaminated.table$phylum
)

####FIX CORAL #########################################
decontaminated.table <- decontaminated.table %>%
  mutate(
    genus = case_when(
      family == "Poritidae"             ~ "Porites",
      family == "Acroporidae"           ~ "Montipora",
      family == "Pocilloporidae"        ~ "Pocillopora",
      family == "Xeniidae"              ~ "Sarcothelia",
      genus == "Sahlingia"              ~ "Erythrotrichia",
      family == "Agaricia"              ~ "Pavona",
      genus == "Podabacia"              ~ "Leptastrea",
      grepl("Agaricia", sScientific)    ~ "Pavona",
      grepl("Porites", sScientific)     ~ "Porites",
      grepl("Podabacia", sScientific)  ~ "Leptastrea",
      TRUE                               ~ genus
    ),
    family = case_when(
      grepl("Podabacia", sScientific)   ~ "Leptastreidae",
      grepl("Agaricia", sScientific)    ~ "Agariciidae",
      grepl("Porites", sScientific)     ~ "Poritidae",
      TRUE                               ~ family
    )
  )

######################################################### 
#decontaminated.table <- decontaminated.table %>%
#  mutate(
#    genus = case_when(
#      family == "Poritidae"             ~ "Porites",
#      family == "Acroporidae"           ~ "Montipora",
#      family == "Pocilloporidae"        ~ "Pocillopora",
#      family == "Xeniidae"              ~ "Sarcothelia",
#      genus == "Sahlingia"              ~ "Erythrotrichia",
#      family=="Agaricia"~"Pavona",
#      genus=="Podabacia"~"Leptastrea",
#      TRUE                               ~ genus
#    )
#  )
####FILTER MICRO TAXA##############################
# Define taxa to exclude
remove_domain = c("Bacteria")

remove_kingdoms <- c("Fungi", "Bacillati", "Pseudomonadati", "Thermotogati", 
                     "Fusobacteriati", "Methanobacteriati")

remove_phyla <- c("Apicomplexa", "Streptophyta", "Bacillariophyta", "Bolidophyceae", "Euglenida", "Verrucomicrobiota", "Euglenozoa", "Haptophyta", "Prasinodermophyta", "Cercozoa", "Foraminifera", "Candidatus Peregrinibacteriota", "Oomycota", "Nebulidia", "Endomyxa", "Chromerida","Telonemia", "Perkinsozoa", "Picozoa", "")

remove_classes <- c("Mamiellophyceae", "Trebouxiophyceae", "Acantharea", "Chrysophyceae", 
                    "Cryptophyta", "Chlorodendrophyceae", "Chloropicophyceae", 
                    "Collembola", "Dictyochophyceae", "Dinophyceae", "Haptophyceae", 
                    "Heterotrichea", "Insecta", "Aves", "Labyrinthulomycetes", 
                    "Litostomatea", "Nephroselmidophyceae", "Oligohymenophorea", 
                    "Oomycetes", "Pelagophyceae", "Phyllopharyngea", "Polycystinea", 
                    "Prostomatea", "Raphidophyceae", "Spirotrichea", "Pedinophyceae", "Bolidophyceae", "Porphyridiophyceae", "Cryptophyceae", "Bigyra", "Pinguiophyceae", "Ichthyosporea", "Choanoflagellata", "Chrysomerophyceae", "Pseudoscourfieldiophyceae")

remove_orders <- c("Pyramimonadales", "Bicosoecida", "Choanoflagellida", "Chlamydomonadales", "Isochrysidales", "Pavlovales", "Perkinsida", "Phaeocystales", 
                   "Prymnesiales", "Himatismenida", "Kinetoplastida", "Chlamydomonadales", "Synechococcales", "Prasinococcales", "Primates")

remove_families <- c("Paramoebidae", "Pycnococcaceae", "Canidae")

remove_genera = c("Lotharella")


#### Apply all filters ##################################################
data <- subset(decontaminated.table, (is.na(domain) | !(domain %in% remove_domain)) & (is.na(kingdom) | !(kingdom %in% remove_kingdoms)) & (is.na(phylum) | !(phylum %in% remove_phyla)) & (is.na(class) | !(class %in% remove_classes)) & (is.na(order) | !(order %in% remove_orders)) & (is.na(genus) | !(genus %in% remove_genera)) & (is.na(family) | !(family %in% remove_families)))
#########################################################################
#### Unhide to keep all taxa ###########################################
#data=decontaminated.table
########################################################################


rownames(data)=data$zotu
data$zotu=NULL

# Remove zero sum rows/cols
data.narm <- data[, colSums(data[, 1:samples]) != 0]
data.narm <- data[rowSums(data[, 1:samples]) != 0, ]

final <- data.narm
final$OTU=rownames(final)
taxa=final[, ((samples+1):ncol(final))]

write.csv(final, file=file.path(export_dir, "2data_12S.csv"), row.names = F)
```

### iii. Top Phyla

#### a. Reads

```{r}
top.phyla.df = data
top.phyla.df$total <- rowSums(top.phyla.df[1:samples], na.rm = TRUE)

# Summarize total reads per phylum
top_5_phyla <- ddply(top.phyla.df, .(phylum), summarise, total = sum(total))

# Calculate grand total and percentage
top_5_phyla$grandtotal = sum(top_5_phyla$total)
top_5_phyla$percent = top_5_phyla$total / top_5_phyla$grandtotal * 100

# Remove missing or empty phylum values
top_5_phyla <- top_5_phyla %>% filter(!is.na(phylum) & phylum != "")

# Rank phyla by total count and assign color labels
top_5_phyla <- top_5_phyla %>%
  arrange(desc(total)) %>%
  mutate(color_group = ifelse(phylum %in% c("Chlorohyta", "Rhodohyta"), "Other", 
                              ifelse(row_number() <= 5, "Top 5", "Other")))

# Ensure phylum is a factor, and levels are in order of total
top_5_phyla$phylum <- factor(top_5_phyla$phylum, levels = top_5_phyla$phylum)

library(scales)
topphyla=ggplot(top_5_phyla, aes(x = phylum, y = total, fill = color_group)) +
  geom_col() +  
  scale_fill_manual(values = c("Top 5" = "black", "Other" = "grey")) +
  labs(x = "", y = "Total Reads") +
  My_Theme +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_sqrt(labels = scales::comma) + coord_flip()

topphyla

#png(filename="3top5phyla_12S.png", type="cairo",units="in", width=5, height=7, pointsize=12, res=300)
#topphyla
#dev.off()
```

#### b. OTUs

```{r}
# Create a column that counts OTUs as 1 per row (presence of a taxon)
top.phyla.df$OTU_count <- 1

# Summarize OTU counts per phylum
top_5_otus <- ddply(top.phyla.df, .(phylum), summarise, total_otus = sum(OTU_count, na.rm = TRUE))

# Remove missing or empty phylum values
top_5_otus <- top_5_otus %>% filter(!is.na(phylum) & phylum != "")

# Rank phyla by OTU counts and categorize Top 5 vs Other
top_5_otus <- top_5_otus %>%
  arrange(desc(total_otus)) %>%
  mutate(color_group = ifelse(row_number() <= 5, "Top 5", "Other"))

# Make phylum a factor in order of total OTU counts
top_5_otus$phylum <- factor(top_5_otus$phylum, levels = top_5_otus$phylum)


otu_plot <- ggplot(top_5_otus, aes(x = phylum, y = total_otus, fill = color_group)) +
  geom_col() +
  scale_fill_manual(values = c("Top 5" = "black", "Other" = "grey")) +
  labs(x = "", y = "Number of OTUs") +
  My_Theme +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_flip()

otu_plot
```

#### c. Combined

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Summarize read counts per phylum
reads_df <- top.phyla.df %>%
  mutate(total_reads = rowSums(select(., starts_with("MM"), starts_with("MK")), na.rm = TRUE)) %>%
  group_by(phylum) %>%
  summarise(value = sum(total_reads, na.rm = TRUE)) %>%
  mutate(metric = "Reads")

# Summarize OTU counts per phylum
otus_df <- top.phyla.df %>%
  mutate(OTU_count = 1) %>%
  group_by(phylum) %>%
  summarise(value = sum(OTU_count, na.rm = TRUE)) %>%
  mutate(metric = "zOTUs")

# Combine
combined_df <- bind_rows(reads_df, otus_df)

# Remove missing/empty phylum
combined_df <- combined_df %>% filter(!is.na(phylum) & phylum != "")

# Optionally, keep only top 5 by reads or OTUs for coloring
top_phyla <- combined_df %>%
  group_by(phylum) %>%
  summarise(max_value = max(value)) %>%
  arrange(desc(max_value)) %>%
  slice_head(n = 5) %>%
  pull(phylum)

combined_df <- combined_df %>%
  mutate(color_group = ifelse(phylum %in% top_phyla, "Top 5", "Other"))

# Make phylum a factor in order of max value for consistent plotting
combined_df$phylum <- factor(combined_df$phylum, levels = combined_df %>% 
                               group_by(phylum) %>% summarise(max_value = max(value)) %>%
                               arrange(max_value) %>% pull(phylum))

combined_df$Marker="12S"

ggplot(combined_df, aes(x = phylum, y = value, fill = color_group)) +
  geom_col() +
  scale_fill_manual(values = c("Top 5" = "black", "Other" = "grey")) +
  labs(x = "", y = "", fill = "") +
  My_Theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") + scale_y_sqrt() +
  coord_flip() +
  facet_wrap(~metric, scales = "free_x")  # free_x allows different scales for reads vs OTUs

write.csv(combined_df, file=file.path(export_dir, "2OTUreadcounts_12S.csv"), row.names = F)
```

#### d. Taxonomic Diagnostic

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Define taxonomic ranks
rank_levels <- c("phylum", "class", "order", "family", "genus", "species")

# Replace "dropped" with "" in taxonomic columns
top.phyla.df <- top.phyla.df %>%
  mutate(across(all_of(rank_levels), ~ ifelse(. == "dropped", "", .)))

# Compute total reads per OTU (sum across all sample columns)
sample_cols <- grep("^(MM|MK)", colnames(top.phyla.df), value = TRUE)
otu_reads <- top.phyla.df %>%
  rowwise() %>%
  mutate(total_reads = sum(c_across(all_of(sample_cols)), na.rm = TRUE)) %>%
  ungroup()

# Summarize OTUs and reads at each rank
rank_summary <- lapply(rank_levels, function(rank) {
  subset_otu <- otu_reads %>% filter(!is.na(.data[[rank]]) & .data[[rank]] != "")
  data.frame(
    rank = rank,
    total_OTUs = nrow(subset_otu),
    total_reads = sum(subset_otu$total_reads)
  )
}) %>% bind_rows()

# Optionally, fractions for mirrored plot
rank_summary <- rank_summary %>%
  mutate(
    frac_OTUs = total_OTUs / max(total_OTUs),
    frac_reads = total_reads / max(total_reads)
  ) %>%
  pivot_longer(cols = c(frac_OTUs, frac_reads), 
               names_to = "metric", values_to = "fraction") %>%
  mutate(
    fraction = ifelse(metric == "frac_OTUs", -fraction, fraction),
    metric = recode(metric, "frac_OTUs" = "OTUs", "frac_reads" = "Reads"),
    rank = factor(rank, levels = rank_levels)
  )

# Plot mirrored OTUs vs Reads
ggplot(rank_summary, aes(x = rank, y = fraction, fill = metric)) +
  geom_col(width = 0.6, position = "identity") +
  coord_flip() +
  scale_y_continuous(labels = abs, limits = c(-1,1)) +
  scale_fill_manual(values = c("OTUs" = "skyblue", "Reads" = "orange")) +
  labs(x = "Taxonomic Rank", y = "Fraction of total", fill = "Metric") +
  theme_minimal() +
  theme(axis.text.y = element_text(face = "bold"),
        legend.position = "top")

```


## C. Normalizing Reads

### i. Rarefaction to even depth

```{r}
# Raw reads data
table.raw=final[, 1:samples]
final_reads=final
#write.csv(final_reads, file=file.path(export_dir, "2OTU_reads_12S.csv"), row.names = F)
table.reads=final_reads[,1:samples]

# Remove zero sum samples
#table.reads <- table.reads[, colSums(table.reads) > 0]
# Create histogram without x-axis
# Calculate quantiles
qs <- quantile(colSums(table.reads), probs = c(0.25, 0.5, 0.75))  # 25%, 50%, 75% quartiles

# Create histogram without x-axis
hist(colSums(table.reads), breaks=50, xaxt="n", main="Read counts per sample", 
     xlab="Reads per sample", col="lightblue")

# Add x-axis with more frequent labels
x_vals <- seq(0, max(colSums(table.reads)), by=1000)  # adjust interval as needed
axis(1, at=x_vals, labels=x_vals)

# Add vertical lines for quantiles
abline(v=qs, col=c("red","green","blue"), lty=2, lwd=2)

# Optional: add legend
legend("topright", legend=c("25%","50%","75%"), col=c("red","green","blue"), lty=2, lwd=2)

# Calculate the first quartile of read counts
quartile <- quantile(colSums(table.reads), probs = 0.5)

# Set threshold to the first quartile
threshold <- quartile

table.reads.filtered <- table.reads[, colSums(table.reads) >= threshold]
min_depth <- min(colSums(table.reads.filtered))
#table.rarefied <- t(rrarefy(t(table.reads.filtered), sample = min_depth))

table.rarefied=table.raw # unrarefied
```

### ii. Transformations

```{r}
# Hellinger transformation
table_hel=as.data.frame(decostand(table.reads, method = "hellinger"))
table_hel$OTU=rownames(table_hel)
final_hel=merge(table_hel, taxa, by="OTU")
rownames(final_hel)=final_hel$OTU
#write.csv(final_hel, file=file.path(export_dir, "2OTU_hellinger_12S.csv"), row.names = F)
table_hel$OTU=NULL

# Binary table
table_pa=as.data.frame(decostand(table.reads, method = "pa"))
table_pa$OTU=rownames(table_pa)
final_pa=merge(table_pa, taxa, by="OTU")
rownames(final_pa)=final_pa$OTU
table_pa$OTU=NULL

# eDNA index
# Convert to matrix if not already
table.raw <- as.matrix(table.reads)  
# Step 1: Compute relative abundance per sample (column-wise)
relative_abundance <- sweep(table.raw, 2, colSums(table.raw), "/")  
# Step 2: Find the maximum observed abundance for each OTU across all samples (row-wise)
max_relative_abundance <- apply(relative_abundance, 1, max, na.rm = TRUE)  
# Step 3: Compute the eDNA index by dividing each OTU's abundance by its max
table_index <- as.data.frame(sweep(relative_abundance, 1, max_relative_abundance, "/"))  
# Add OTU IDs back to the table
table_index$OTU <- rownames(table_index)
# Merge with taxa metadata (assuming 'taxa' has OTU_ID as key)
final_index <- merge(table_index, taxa, by = "OTU")
# Ensure row names match OTU_ID
rownames(final_index) <- final_index$OTU
# Remove OTU_ID column from table_index if needed
table_index$OTU <- NULL
```

## D. Choose & Merge Metadata

```{r}
####CHOOSE MATRIX#############
community.table=table_index
community.data=final_index
distance.method="bray"
##############################

table.t <- as.data.frame(t(as.matrix(community.table)))

# Replace NaN (and NA) with 0
# Save row names
rn <- rownames(table.t)

# Replace NA/NaN safely
table.t <- as.data.frame(lapply(table.t, function(x) {
  x[is.na(x) | is.nan(x)] <- 0
  x
}))

# Restore row names
rownames(table.t) <- rn

# Remove zero sum columns
table.t <- table.t[, colSums(table.t) != 0, drop = FALSE]
# Remove rows where the sum is 0
table.t <- table.t[rowSums(table.t) != 0, , drop = FALSE]

table.t$SAMPLE=row.names(table.t) ##OTU table only
metatable.reps=merge(table.t, metadata, by="SAMPLE") ##OTU table with site metadata
table.t$SAMPLE=NULL

ntax=as.numeric(ncol(table.t))
nsample=as.numeric(nrow(table.t))
```

### i. Reads by Site

```{r}
table.reads.t <- as.data.frame(t(table.rarefied[rowSums(table.rarefied) > 0, ]))  # Remove samples with zero reads

sum(table.reads.t) #412,476 total post-processed rarefied reads

table.reads.t$SAMPLE=row.names(table.reads.t) ##OTU table only
# Remove leading zeros from SAMPLE column
metatable.reps.reads=merge(table.reads.t, metadata, by="SAMPLE") ##OTU table with site metadata
metatable.reps.reads$MARKER="12S"
table.reads.t$SAMPLE=NULL
```


## E. Accumulation curves

### i. Sequencing

```{r}
#### SAMPLE RAREFACTION ############################################################
#png(filename=file.path(export_dir, "3rarecurve_read_coverage_12S.png"), type="cairo",units="in", width=7, height=5, pointsize=12, res=300)
#rarecurve(table.reads.t, step = 100, col = rainbow(nrow(table.reads.t)), label = F, xlab = "Sequence Depth",  ylab = "MOTUs", main=marker.name)
#dev.off()
####################################################################################
```




### ii. Overall

```{r}
library(iNEXT)
library(ggplot2)
library(reshape2)

# Step 1: Combine all samples across methods
metatable.all <- metatable.reps  # use full dataset

# Step 2: Create incidence matrix (presence/absence)
otu_matrix <- as.matrix(t(metatable.all[, 2:ntax]))
otu_matrix[is.nan(otu_matrix)] <- 0
otu_matrix <- otu_matrix[rowSums(!otu_matrix) < ncol(otu_matrix), ]
otu_matrix <- otu_matrix[, colSums(otu_matrix) != 0]
otu_matrix <- decostand(otu_matrix, method = "pa")

# Step 3: Wrap into a list for iNEXT (even for one sample)
otu_list_all <- list("Overall" = otu_matrix)

######## Step 4: Run iNEXT #####################################################
out.all <- iNEXT(otu_list_all, q = 0, datatype = "incidence_raw", se = TRUE)
saveRDS(out.all, file=file.path(export_dir, "2inext_overall_12S.rds"))
###############################################################################
out.all=readRDS(file.path(export_dir, "2inext_overall_12S.rds"))

asymp_richness <- out.all$AsyEst %>%
  dplyr::filter(Diversity == "Species richness")
asymp_richness


# Step 5: Plot total richness accumulation curve
rarecurve.all <- ggiNEXT(out.all, type = 1, se = TRUE, facet.var = "None", color.var = "Assemblage", grey = FALSE)
rarecurve.all

png(filename=file.path(export_dir, "3rarecurve_totalrichness_12S.png"), type="cairo", units="in", width=4, height=6, pointsize=12, res=300)
#cairo_pdf(file = "2rarecurve_totalrichness_12S.pdf", width = 5, height = 4)
rarecurve.all +
  scale_colour_manual(values = c("Overall" = "black")) +
  scale_fill_manual(values = c("Overall" = "black")) +
  My_Theme + theme(legend.position = "none")+
  ggtitle(marker.name) + xlab("Samples") + ylab("MOTU Richness")
dev.off()
```

### iii. Method

```{r}
metatable.active=subset(metatable.reps.reads, FILTRATION=="Active")
rownames(metatable.active)=metatable.active$SAMPLE
metatable.passive=subset(metatable.reps.reads, FILTRATION=="Passive")
rownames(metatable.passive)=metatable.passive$SAMPLE

smallset.active=as.matrix(t(metatable.active[, c(2:ntax)]))
smallset.active[is.nan(smallset.active)] <- 0
smallset.active=smallset.active[rowSums(!smallset.active) < ncol(smallset.active), ] #remove individuals with zero sequence reads
smallset.active <- smallset.active[, colSums(smallset.active) != 0] # remove MOTUs with zero sums
smallset.active=decostand(smallset.active, method = "pa")

smallset.passive=as.matrix(t(metatable.passive[, c(2:ntax)]))
smallset.passive[is.nan(smallset.passive)] <- 0
smallset.passive=smallset.passive[rowSums(!smallset.passive) < ncol(smallset.passive), ] #remove individuals with zero sequence reads
smallset.passive <- smallset.passive[, colSums(smallset.passive) != 0] # remove MOTUs with zero sums
smallset.passive=decostand(smallset.passive, method = "pa")

otu_active <- colSums(smallset.active)
otu_passive <- colSums(smallset.passive)

otu_list_method <- list(
  "Active" = smallset.active,
  "Passive" = smallset.passive
)

#### Rarefaction SITE ##############################################
out.method=iNEXT(otu_list_method, q=0, datatype="incidence_raw", se=T) 
saveRDS(out.method, file=file.path(export_dir, "2inext_method_12S.rds"))
####################################################################
out.method=readRDS(file=file.path(export_dir, "2inext_method_12S.rds"))

asymp_richness <- out.method$AsyEst %>%
  dplyr::filter(Diversity == "Species richness")
asymp_richness

rarecurve.method=ggiNEXT(out.method, type=1, se=TRUE, facet.var="None", color.var="Assemblage", grey=FALSE) 

rarecurve.method

#cairo_pdf(file = "3rarecurve_method_12S.pdf", width = 5, height = 4)
png(filename=file.path(export_dir, "3rarecurve_method_12S.png"), type="cairo", units="in", width=5, height=4, pointsize=12, res=300)
rarecurve.method + scale_colour_manual(values=cols.method) + scale_fill_manual(values=cols.method) + My_Theme + ggtitle(marker.name) 
dev.off()
```

####a. Coverage rarefaction

```{r}
###Do methods differ in richness, independent of region?
coverage.method <- out.method$DataInfo$SC
coverage.method


richness_cov_method <- estimateD(
  otu_list_method,
  q = c(0,1,2),
  datatype = "incidence_raw",
  base = "coverage",
  level = min(coverage.method)
)

richness_cov_method

method_hill_df <- richness_cov_method |>
  dplyr::mutate(
    Order.q = factor(
      Order.q,
      levels = c(0, 1, 2),
      labels = c("q = 0\n(Richness)",
                 "q = 1\n(Shannon)",
                 "q = 2\n(Simpson)")
    )
  )


ggplot(method_hill_df,
       aes(x = Order.q,
           y = qD,
           color = Assemblage)) +

  geom_point(
    position = position_dodge(width = 0.4),
    size = 3
  ) +

  geom_errorbar(
    aes(ymin = qD.LCL, ymax = qD.UCL),
    position = position_dodge(width = 0.4),
    width = 0.15,
    linewidth = 0.8
  ) +

  scale_color_manual(
    values = cols.method,   # your existing colors
    name = ""
  ) +

  labs(
    x = NULL,
    y = "Hill diversity (effective number of taxa)"
  ) +

  My_Theme +
  theme(
    legend.position = "top" )

```


### iv. Region

```{r}
metatable.keehi=subset(metatable.reps.reads, REGION=="Keehi")
metatable.maunalua=subset(metatable.reps.reads, REGION=="Maunalua")


smallset.keehi=as.matrix(t(metatable.keehi[, c(2:ntax)]))
smallset.keehi[is.nan(smallset.keehi)] <- 0
smallset.keehi=smallset.keehi[rowSums(!smallset.keehi) < ncol(smallset.keehi), ] #remove individuals with zero sequence reads
smallset.keehi <- smallset.keehi[, colSums(smallset.keehi) != 0] # remove MOTUs with zero sums
smallset.keehi=decostand(smallset.keehi, method = "pa")

smallset.maunalua=as.matrix(t(metatable.maunalua[, c(2:ntax)]))
smallset.maunalua[is.nan(smallset.maunalua)] <- 0
smallset.maunalua=smallset.maunalua[rowSums(!smallset.maunalua) < ncol(smallset.maunalua), ] #remove individuals with zero sequence reads
smallset.maunalua <- smallset.maunalua[, colSums(smallset.maunalua) != 0] # remove MOTUs with zero sums
smallset.maunalua=decostand(smallset.maunalua, method = "pa")

otu_keehi <- colSums(smallset.keehi)
otu_maunalua <- colSums(smallset.maunalua)

otu_list_region <- list(
  "Keehi" = smallset.keehi,
  "Maunalua" = smallset.maunalua
)

#### Rarefaction SITE ##############################################
out.region=iNEXT(otu_list_region, q=0, datatype="incidence_raw", se=T) 
saveRDS(out.region, file=file.path(export_dir, "2inext_region_12S.rds"))
####################################################################
out.region=readRDS(file=file.path(export_dir, "2inext_region_12S.rds"))

asymp_richness <- out.region$AsyEst %>%
  dplyr::filter(Diversity == "Species richness")
asymp_richness

rarecurve.region=ggiNEXT(out.region, type=1, se=TRUE, facet.var="None", color.var="Assemblage", grey=FALSE) 

rarecurve.region

#cairo_pdf(file = "3rarecurve_region_12S.pdf", width = 5, height = 4)
png(filename=file.path(export_dir, "3rarecurve_region_12S.png"), type="cairo", units="in", width=5, height=4, pointsize=12, res=300)
rarecurve.region + scale_colour_manual(values=cols.region2) + scale_fill_manual(values=cols.region) + My_Theme + ggtitle(marker.name)
dev.off()
```


####a. Coverage rarefaction

```{r}
###Do regions differ in richness, independent of method?
coverage.region <- out.region$DataInfo$SC
coverage.region

#summary, singletons (Q1), doubletons (Q2), etc.
out.region$DataInfo

richness_cov_region <- estimateD(
  otu_list_region,
  q = c(0,1,2),
  datatype = "incidence_raw",
  base = "coverage",
  level = min(coverage.region)
)

richness_cov_region

region_hill_df <- richness_cov_region |>
  dplyr::mutate(
    Order.q = factor(
      Order.q,
      levels = c(0, 1, 2),
      labels = c("q = 0\n(Richness)",
                 "q = 1\n(Shannon)",
                 "q = 2\n(Simpson)")
    )
  )


ggplot(region_hill_df,
       aes(x = Order.q,
           y = qD,
           color = Assemblage)) +

  geom_point(
    position = position_dodge(width = 0.4),
    size = 3
  ) +

  geom_errorbar(
    aes(ymin = qD.LCL, ymax = qD.UCL),
    position = position_dodge(width = 0.4),
    width = 0.15,
    linewidth = 0.8
  ) +

  scale_color_manual(
    values = cols.region,   # your existing colors
    name = ""
  ) +

  labs(
    x = NULL,
    y = "Hill diversity (effective number of taxa)"
  ) +

  My_Theme +
  theme(
    legend.position = "top" )

```
### v. Interaction

```{r}
# Keehi
keehi_active   <- subset(metatable.reps.reads, REGION=="Keehi" & FILTRATION=="Active")
keehi_passive  <- subset(metatable.reps.reads, REGION=="Keehi" & FILTRATION=="Passive")

# Maunalua
maun_active    <- subset(metatable.reps.reads, REGION=="Maunalua" & FILTRATION=="Active")
maun_passive   <- subset(metatable.reps.reads, REGION=="Maunalua" & FILTRATION=="Passive")

make_incidence <- function(df, ntax) {
  m <- as.matrix(t(df[, 2:ntax]))
  m[is.nan(m)] <- 0
  m <- m[rowSums(!m) < ncol(m), ]
  m <- m[, colSums(m) != 0]
  decostand(m, method = "pa")
}

keehi_active_m  <- make_incidence(keehi_active, ntax)
keehi_passive_m<- make_incidence(keehi_passive, ntax)

maun_active_m  <- make_incidence(maun_active, ntax)
maun_passive_m<- make_incidence(maun_passive, ntax)

## Keehi
otu_keehi_methods <- list(
  Active  = keehi_active_m,
  Passive = keehi_passive_m
)

out.keehi <- iNEXT(
  otu_keehi_methods,
  q = c(0,1,2),
  datatype = "incidence_raw",
  se = TRUE
)

cov_keehi <- min(out.keehi$DataInfo$SC)

rich_keehi <- estimateD(
  otu_keehi_methods,
  q = c(0,1,2),
  datatype = "incidence_raw",
  base = "coverage",
  level = cov_keehi
)


## Maunalua 
otu_maun_methods <- list(
  Active  = maun_active_m,
  Passive = maun_passive_m
)

out.maun <- iNEXT(
  otu_maun_methods,
  q = c(0,1,2),
  datatype = "incidence_raw",
  se = TRUE
)

cov_maun <- min(out.maun$DataInfo$SC)

rich_maun <- estimateD(
  otu_maun_methods,
  q = c(0,1,2),
  datatype = "incidence_raw",
  base = "coverage",
  level = cov_maun
)
```

####a. Coverage rarefaction

```{r}
###Does method performance differ between regions?
interaction_hill <- dplyr::bind_rows(
  rich_keehi  |> dplyr::mutate(Region = "Keehi"),
  rich_maun   |> dplyr::mutate(Region = "Maunalua")
) |>
  dplyr::mutate(
    Order.q = factor(
      Order.q,
      levels = c(0,1,2),
      labels = c("q = 0\n(Richness)",
                 "q = 1\n(Shannon)",
                 "q = 2\n(Simpson)")
    )
  )
interaction_hill$Marker="12S"
write.csv(interaction_hill, file=file.path(export_dir, "2iNext_hill_12S.csv"), fileEncoding = "UTF-8", row.names = F)

ggplot(interaction_hill,
       aes(x = Order.q,
           y = qD,
           color = Assemblage)) +

  geom_point(
    position = position_dodge(width = 0.4),
    size = 3
  ) +

  geom_errorbar(
    aes(ymin = qD.LCL, ymax = qD.UCL),
    position = position_dodge(width = 0.4),
    width = 0.15,
    linewidth = 0.8
  ) +

  facet_wrap(~ Region) +

  scale_color_manual(
    values = cols.method,
    name = "Method"
  ) +

  labs(
    x = NULL,
    y = "Hill diversity (effective number of taxa)"
  ) +

  My_Theme +
  theme(
    legend.position = "top"
  )

## Export iNext sample coverage
keehi_cov <- out.keehi$iNextEst$coverage_based |>
  dplyr::mutate(Region = "Keehi")

maun_cov <- out.maun$iNextEst$coverage_based |>
  dplyr::mutate(Region = "Maunalua")
interaction_coverage <- dplyr::bind_rows(keehi_cov, maun_cov)
interaction_coverage <- interaction_coverage |>
  dplyr::mutate(
    Order.q = factor(
      Order.q,
      levels = c(0, 1, 2),
      labels = c(
        "q = 0\n(Richness)",
        "q = 1\n(Shannon)",
        "q = 2\n(Simpson)"
      )
    )
  )

interaction_coverage$Marker="12S"
write.csv(interaction_coverage, file=file.path(export_dir, "2iNext_coverage_12S.csv"), fileEncoding = "UTF-8", row.names = F)
```

###xi. Site Richness

```{r}
otu_cols <- grep("^Zotu", colnames(metatable.reps.reads), value = TRUE)

# -----------------------------
# 2. Aggregate abundances by SITE_ID
# -----------------------------
site_abundance <- metatable.reps.reads %>%
  group_by(SITE_ID) %>%
  summarise(across(all_of(otu_cols), sum, na.rm = TRUE)) %>%
  ungroup()

# -----------------------------
# 3. Remove sites with <=1 species (optional but prevents errors)
# -----------------------------
num_species_site <- rowSums(site_abundance[, otu_cols] > 0)
site_abundance <- site_abundance[num_species_site > 1, ]

# -----------------------------
# 4. Estimate coverage-based richness per site
# -----------------------------
site_coverage_df <- bind_rows(
  lapply(1:nrow(site_abundance), function(i) {
    result <- tryCatch({
      df <- estimateD(
        as.numeric(site_abundance[i, otu_cols]),
        datatype = "abundance",
        base = "coverage"        # coverage-standardized
        # ,level = 0.9            # optional: specify target coverage (e.g., 90%)
      )
      df$SITE_ID <- site_abundance$SITE_ID[i]
      df
    }, error = function(e) {
      # Return NA for sites that fail
      data.frame(Order = NA, qD = NA, SC = NA, SITE_ID = site_abundance$SITE_ID[i])
    })
    result
  })
)

richness_df <- site_coverage_df %>% filter(Order.q == 0)
metadata_site <- metadata %>%
  group_by(SITE_ID) %>%
  summarise(
    LAT = first(LAT),
    LON = first(LON),
    REGION = first(REGION),
    SITE = first(SITE),
    .groups = "drop"
  )

richness_df <- richness_df %>%
  left_join(metadata_site, by = "SITE_ID")
richness_df$Marker="12S"

write.csv(richness_df, file=file.path(export_dir, "2iNext_coverage_site_12S.csv"), fileEncoding = "UTF-8", row.names = F)
```

# 2. Exploration

## A. Read Abundance

```{r}
data.raw <- final_reads %>% 
  pivot_longer(cols = 1:samples, 
               names_to = "SAMPLE", 
               values_to = "value") %>%
  dplyr::select(OTU, phylum, SAMPLE, value)  # Explicitly use dplyr's select

data.raw=merge(data.raw, metadata, by="SAMPLE")
data.raw.narm=subset(data.raw, value>0)

# Summary of raw reads
data_summary <- data.raw.narm %>%
  group_by(SITE) %>%
  summarise(
    n          = sum(!is.na(value)),
    mean_value = mean(value, na.rm = TRUE),
    sd_value   = sd(value, na.rm = TRUE),
    se_value   = sd_value / sqrt(n),
    .groups = "drop"
  )
data_summary
```

### i. Data Melt

```{r}
data.long <- community.data %>% 
  pivot_longer(cols = c(2:nsample+1), 
               names_to = "SAMPLE", 
               values_to = "value") %>%
  dplyr::select(OTU, phylum, class, order, family, genus, species, 
                sScientific, pident, evalue, bitscore, 
                SAMPLE, value)

data.long.narm=filter(data.long, value>0)

data.long.both=merge(data.long.narm, metadata, by="SAMPLE")
data.long.both$value=as.numeric(data.long.both$value)
data.long.both$SITE=as.factor(data.long.both$SITE)

write.csv(data.long.both, file=file.path(export_dir, "2eDNA_data_long_12S.csv"), fileEncoding = "UTF-8", row.names = F)
```

### ii. Reads by rank and site

```{r}
# Summarize the data
Modelplot.site <- data.long.both %>%
  dplyr::group_by(phylum, SITE) %>%
  dplyr::summarize(
    mean_reads = mean(value, na.rm = TRUE),
    se = sd(value, na.rm = TRUE) / (length(value)),
    sd=sd(value),
    N = length(value),  # Count of rows in each group
    total_reads = sum(value)
  )

write.csv(Modelplot.site, file=file.path(export_dir, "2reads_by_site_12S.csv"))
```

### iii. Species list

```{r}
#species.list <- data.long.both %>%
#  distinct(sScientific) %>%
#  filter(sScientific != "")
#library(rinat)
#
#get_inat_common_throttle <- function(sp, max_obs = 10, pause_sec = 0.9) {
#  
#  Sys.sleep(pause_sec)  # throttle
#  
#  tryCatch({
#    obs <- get_inat_obs(taxon_name = sp, maxresults = max_obs)
#    
#    # extract common names that aren't NA
#    cn <- obs$common_name[!is.na(obs$common_name)]
#    
#    if (length(cn) == 0) return(NA_character_)
#    
#    # pick the most frequent name
#    most_common <- names(sort(table(cn), decreasing = TRUE))[1]
#    most_common
#    
#  }, error = function(e) {
#    NA_character_
#  })
#}

#species.list$common_inat <- sapply(species.list$sScientific, get_inat_common_throttle)
#saveRDS(species.list, file="2species_list_12S.rds")
#species.list=readRDS(file="2species_list_12S.rds")

##merge with reads data
#species.list.table=merge(species.list, final_reads, by="sScientific")

## summarize by site
## Step 1: summarize site/region info using all rows
#site_region_summary <- species.list.table %>%
#  pivot_longer(
#    cols = 3:(2 + samples),
#    names_to = "SAMPLE",
#    values_to = "READS"
#  ) %>%
#  filter(READS > 0) %>%
#  left_join(metadata %>% select(SAMPLE, SITE_ID, REGION), by = "SAMPLE") %>%
#  group_by(sScientific) %>%
#  summarise(
#    sites_present = paste(unique(SITE_ID), collapse = ", "),
#    regions_present = paste(unique(REGION), collapse = ", "),
#    n_sites = n_distinct(SITE_ID),
#    .groups = "drop"
#  )

## Step 2: pick the row with the largest bitscore per species
#best_hit <- species.list.table %>%
#  group_by(sScientific) %>%
#  slice_max(bitscore, n = 1, with_ties = FALSE) %>%
#  ungroup() %>%
#  select(
#    sScientific, common_inat, phylum, class, order, family, genus, pident, bitscore, evalue
#  )

## Step 3: join the site/region summary
#species_summary <- best_hit %>%
#  left_join(site_region_summary, by = "sScientific") %>%
#  arrange(desc(n_sites)) %>%
#  mutate(MARKER = "12S")



## Flatten the 'common' list-column to character
#species_summary <- species_summary %>%
#  mutate(common_inat = sapply(common_inat, function(x) {
#    if(length(x) == 0) {
#      NA_character_       # replace empty list with NA
#    } else {
#      paste(x, collapse = "; ")  # join all entries with "; "
#    }
#  }))

#species_summary$common_inat=ifelse(species_summary$common_inat=="NA", "", species_summary$common_inat)
#readr::write_excel_csv(species_summary, "2species_list_12S.csv")
```


# 3. Community Data

## A. Dissimilarity matrices

### i. Binary

```{r}
##12S##
table.bin.dist<-dist(decostand(table.t, method = "pa"), method='binary')
```

### ii. Bray-Curtis

```{r}
##12S##
table.dist<- vegdist(table.t, method=distance.method, binary=F)
```

### iii. Taxonomic-level

```{r}
site=as.factor(metatable.reps$SITE)
method=as.factor(metatable.reps$FILTRATION)
location=as.factor(metatable.reps$LOCATION)

com <- as.matrix(t(community.table))
colnames(com) <- community.data[[taxonomic.rank]]

mode(com) <- "numeric"

prefixes <- unique(sub("\\..*", "", colnames(com)))
prefixes <- prefixes[prefixes != ""]

com_collapsed <- sapply(prefixes, function(x) {
  cols <- which(startsWith(colnames(com), x))
  
  if (length(cols) == 0) {
    # If no columns match the prefix, return a vector of zeros or NAs
    return(rep(0, nrow(com)))
  } else {
    return(rowSums(com[, cols, drop = FALSE], na.rm = TRUE))
  }
})

com=as.data.frame(com_collapsed)
ncom=ncol(com)

com$SAMPLE=row.names(com)
com.meta=merge(com, metadata, by="SAMPLE")
com$SAMPLE=NULL
colnames(com)[is.na(colnames(com)) | colnames(com) == "dropped"] <- "Unknown"
```

####a. Reads community table

```{r}
com.reads=as.matrix(t(table.rarefied))

colnames(com.reads)=community.data[[taxonomic.rank]]


prefixes = unique(sub("\\..*", "", colnames(com.reads)))
prefixes = prefixes[prefixes != ""]

com_collapsed.reads <- sapply(prefixes, function(x) {
  cols <- which(startsWith(colnames(com.reads), x))
  
  if (length(cols) == 0) {
    # If no columns match the prefix, return a vector of zeros or NAs
    return(rep(0, nrow(com.reads)))
  } else {
    return(rowSums(com.reads[, cols, drop = FALSE], na.rm = TRUE))
  }
})


com.reads=as.data.frame(com_collapsed.reads)
com.reads$SAMPLE=row.names(com.reads)
com.reads.meta=merge(com.reads, metadata, by="SAMPLE")
rownames(com.reads.meta)=com.reads.meta$SAMPLE
com.reads$SAMPLE=NULL
com.reads.meta$SAMPLE=NULL
colnames(com.reads)[is.na(colnames(com.reads)) | colnames(com.reads) == "dropped"] <- "Unknown"

#write.csv(com.reads.meta, file=file.path(export_dir, "2OTU_com_12S.csv"), row.names = F)
```


### iv. Dissimilarity by method

```{r}
dist_df <- as.matrix(table.dist) %>%
  as.data.frame() %>%
  rownames_to_column("Sample1") %>%
  pivot_longer(-Sample1, names_to = "Sample2", values_to = "distance")

dist_df <- dist_df %>%
  filter(Sample1 < Sample2)

filtration_lookup <- metadata %>%
  select(SAMPLE, FILTRATION) %>%
  distinct()


dist_df <- dist_df %>%
  dplyr::left_join(filtration_lookup,
            by = c("Sample1" = "SAMPLE")) %>%
  dplyr::rename(Filtration1 = FILTRATION) %>%
  dplyr::left_join(filtration_lookup,
            by = c("Sample2" = "SAMPLE")) %>%
  dplyr::rename(Filtration2 = FILTRATION)

dist_df <- dist_df %>%
  mutate(
    Filtration_comp = case_when(
      Filtration1 == "Active" & Filtration2 == "Active" ~ "Active",
      Filtration1 == "Passive" & Filtration2 == "Passive" ~ "Passive",
      TRUE ~ "Active-Passive"
    )
  )

dist_summary <- dist_df %>%
  group_by(Filtration_comp) %>%
  summarise(
    mean_dist = mean(distance, na.rm = TRUE),
    sd_dist = sd(distance, na.rm = TRUE),
    n = n()
  )

dist_summary=subset(dist_summary, Filtration_comp != "Active-Passive")
dist_summary

dist_df=subset(dist_df, Filtration_comp != "Active-Passive")
dist_df$Marker="12S"

write.csv(dist_df, file=file.path(export_dir, "2dissimilarity_method_12S.csv"), row.names = F)


ggplot(dist_df, aes(Filtration_comp, distance, fill=Filtration_comp)) +
  geom_violin(trim = FALSE, alpha=0.25, color=NA, scale="width") +
  geom_jitter(width = 0.1, height = 0, alpha = 0.05) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.2, color = "black") +
  #geom_boxplot(width = 0.05) +
  scale_fill_manual(values = cols.method) +  
  My_Theme + ylim(0,1)+ theme(legend.position = "none")+ 
  labs(x = "", y = "Community dissimilarity") 
```




## B. Ordination

### i. PERMANOVA


```{r}
# Prepare data
metatable.reps$REGION <- as.factor(metatable.reps$REGION)
metatable.reps$FILTRATION <- as.factor(metatable.reps$FILTRATION)
metatable.reps$SITE_ID <- as.factor(metatable.reps$SITE_ID)

# Distance matrix
table.dist <- vegdist(table.t, method = "bray")

# ============================================================================
# 1. DISPERSION TESTS
# ============================================================================

# Dispersion by REGION
disp_region <- betadisper(table.dist, group = metatable.reps$REGION)
disp_region_test <- permutest(disp_region, permutations = 9999)
disp_region_test #p 0.001

# Dispersion by FILTRATION
disp_filt <- betadisper(table.dist, group = metatable.reps$FILTRATION)
disp_filt_test <- permutest(disp_filt, permutations = 9999)
disp_filt_test #p 0.89
# ============================================================================
# 2. ANOSIM (Use as primary if dispersions differ)
# ============================================================================

# Testing REGION (blocked by site)
anosim_region <- anosim(table.dist, 
                        grouping = metatable.reps$REGION,
                        strata = metatable.reps$SITE_ID,
                        permutations = 9999)

# Testing FILTRATION (blocked by site)
anosim_filt <- anosim(table.dist, 
                      grouping = metatable.reps$FILTRATION,
                      strata = metatable.reps$SITE_ID,
                      permutations = 9999)

# ============================================================================
# 3. PERMANOVA (Use for variance partitioning and interactions)
# ============================================================================

# Main effects (each after accounting for the other)
permanova_main <- adonis2(table.dist ~ REGION + FILTRATION,
                          data = metatable.reps,
                          permutations = 9999,
                          strata = metatable.reps$SITE_ID,
                          by = "margin")
permanova_main
```

### ii. MDS

#### a. Spatial (region)

```{r}
set.seed(13)
table.t <- table.t[rownames(table.t) != failed.samples, ]

#############################################################
NMDSdat<-metaMDS(vegdist(table.t, distance.method), k=3, distance = distance.method, trymax = 1000, weakties=TRUE, maxit=1000, scaling=TRUE, pc=TRUE, zerodist="add")
saveRDS(NMDSdat, file=file.path(export_dir, "2metaMDS_12S.rds"))
############################################################
NMDSdat=readRDS(file=file.path(export_dir, "2metaMDS_12S.rds"))

PDat<-data.frame(NMDS1=NMDSdat$points[,1],
                 NMDS2=NMDSdat$points[,2],
                 SAMPLE=row.names(table.t))
PDat<-merge(PDat, metatable.reps, by="SAMPLE",all.x=T,all.y=F)
write.csv(PDat, file=file.path(export_dir, "2NMDS_data_12S.csv"), row.names = F)

nmds.plot=qplot(x=NMDS1,y=NMDS2,colour=REGION,alpha=I(0.6),data=PDat)+  geom_vline(xintercept = 0, lty=2, col="grey66")+ geom_hline(yintercept = 0, lty=2, col="grey66") + stat_ellipse(aes(fill=REGION), level = 0.95, geom="polygon", alpha=0.3) +
  scale_fill_manual(values = cols.region2) +
  scale_color_manual(values = cols.region2) + My_Theme + theme(legend.position = "bottom") + ggtitle(marker.name)

nmds.plot

NMDSdat$stress
#cairo_pdf(file = "2nmds_location_12S.pdf", width = 6, height = 6)
#nmds.plot
#dev.off()
```


#### b. Method

```{r}
nmds.plot.filtration=qplot(x=NMDS1,y=NMDS2,colour=FILTRATION,alpha=I(0.6),data=PDat)+  geom_vline(xintercept = 0, lty=2, col="grey66")+ geom_hline(yintercept = 0, lty=2, col="grey66") + stat_ellipse(aes(fill=FILTRATION), level = 0.95, geom="polygon", alpha=0.3) +
  scale_fill_manual(values = cols.method) +
  scale_color_manual(values = cols.method) + My_Theme + theme(legend.position = "bottom")

nmds.plot.filtration

#cairo_pdf(file = "3nmds_method_12S.pdf", width = 6, height = 6)
#png(filename="3nmds_method_12S.png", type="cairo", units="in", width=6, height=6, pointsize=12, res=300)
#nmds.plot.filtration + ggtitle(marker.name)
#dev.off()
```

## C. Nestedness

```{r}
# Install (if needed) and load maxnodf
# install.packages("remotes")
#remotes::install_github("christophhoeppke/maxnodf")
library(maxnodf)

# Suppose you have your presence–absence matrix (e.g., 2 methods: Active, Passive)
# Convert reads to presence/absence
otu_pa <- (table.reads > 0) * 1

# Aggregate by method
active_cols  <- metatable.reps$SAMPLE[metatable.reps$FILTRATION == "Active"]
passive_cols <- metatable.reps$SAMPLE[metatable.reps$FILTRATION == "Passive"]

otu_method <- cbind(
  Active  = rowSums(otu_pa[, active_cols, drop=FALSE]) > 0,
  Passive = rowSums(otu_pa[, passive_cols, drop=FALSE]) > 0
)

otu_method <- 1*otu_method  # convert logical to numeric
colSums(otu_method) #fewest otu method is nested within larger

# Compute observed NODF (from maxnodf or nodf_cpp)
obs_nodf <- nodf_cpp(otu_method)  
obs_nodf
```

## D. Heatmap

### i. eDNA Index

#### a. Pooled

```{r}
library(pheatmap)
library(RColorBrewer)
library(tibble)

# Prepare species data: convert to dataframe and remove unwanted columns
com.dropped <- com.reads %>%
  as.data.frame() %>%
  dplyr::select(-any_of("dropped"))

com.dropped$Unknown <- NULL  # remove "Unknown" column if present

# Grouping by SITE_ID
com.reads.meta$Group <- com.reads.meta$SITE_ID

# Sum species abundances per SITE_ID
com_grouped_sum <- com.dropped %>%
  as.data.frame() %>%
  tibble::rownames_to_column("SAMPLE") %>%
  left_join(com.reads.meta %>% 
              tibble::rownames_to_column("SAMPLE") %>%
              select(SAMPLE, SITE_ID),
            by = "SAMPLE") %>%
  group_by(SITE_ID) %>%
  summarise(across(where(is.numeric), sum)) %>%
  column_to_rownames("SITE_ID")


# Add  reads to any value > 0
com_grouped_sum_adj <- com_grouped_sum
com_grouped_sum_adj[com_grouped_sum_adj > 0] <- com_grouped_sum_adj[com_grouped_sum_adj > 0] + 1

# ----------- eDNA index transformation -----------

# Step 1: Compute relative abundance per site (row-wise)
relative_abundance <- sweep(com_grouped_sum_adj, 1, rowSums(com_grouped_sum_adj), "/")

# Step 2: Find max relative abundance for each taxon across sites (column-wise)
max_relative_abundance <- apply(relative_abundance, 2, max, na.rm = TRUE)

# Step 3: Compute the eDNA index: divide each value by its taxon max
edna_index <- sweep(relative_abundance, 2, max_relative_abundance, "/")

# Replace any NaN (from dividing by zero max) with zero
edna_index[is.na(edna_index)] <- 0

# ----------- Prepare annotation metadata -----------

group_meta <- com.reads.meta %>%
  select(SITE_ID, REGION) %>%
  distinct() %>%
  filter(SITE_ID %in% rownames(edna_index)) %>%
  tibble::remove_rownames() %>%       # <— remove existing rownames
  tibble::column_to_rownames("SITE_ID")


site_names <- unique(group_meta$REGION)
site_colors <- cols.region
names(site_colors) <- site_names
ann_colors <- list(REGION = site_colors)

# ----------- Plot heatmap -----------

# Compute Bray-Curtis distances
row_dist <- vegdist(edna_index, method = "bray")  # rows = taxa/samples
row_dist[is.nan(row_dist)] <- 1
col_dist <- vegdist(t(edna_index), method = "bray")  # columns = samples/taxa
col_dist[is.nan(col_dist)] <- 1

# Heatmap with Bray-Curtis
heatmap_edna_index=pheatmap(edna_index,
         scale = "none",
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         clustering_distance_rows = row_dist,
         clustering_distance_cols = col_dist,
         annotation_row = group_meta,
         annotation_colors = ann_colors,
         show_rownames = TRUE,
         show_colnames = TRUE,
         fontsize_row = 10,
         fontsize_col = 10,
         angle_col = 45,
         border_color = "black",
         main = marker.name,
         color = heatmap.cols
)

heatmap_edna_index

# Save to PDF
#cairo_pdf(file = "3heatmap_edna_index_12S.pdf", width = 18, height = 8)
png(filename=file.path(export_dir, "3heatmap_edna_index_12S.png"), type="cairo", units="in", width=18, height=8, pointsize=12, res=300)
heatmap_edna_index
dev.off()
```



# 4. Taxonomic Info

## A. Site Indicator Taxa

### i. Unique to each site

```{r}
#com.meta=subset(com.meta, REGION=="Keʻehi") ##hide for main analysis
# Run indicator species analysis
indval.site = multipatt(com.meta[2:ncom+1], com.meta$SITE_ID, control = how(nperm = numsam), duleg = T, func="r.g")

# View summary
summary(indval.site)

# Extract results as a dataframe
summary_result = as.data.frame(indval.site$sign)

# Extract p-values from the results
pvals <- indval.site$sign$p.value

# Adjust p-values using Benjamini-Hochberg (FDR) correction
pvals_adj <- p.adjust(pvals, method = "BH")

# Add adjusted p-values back to results
summary_result$p.adjusted <- pvals_adj

summary_result=subset(summary_result, p.adjusted <0.05)

# Save full results and significant species to CSV files
write.csv(summary_result, file=file.path(export_dir, "2indvalunique_taxa_12S.csv"), row.names = T)
```

### iii. Region

```{r}
indval.region = multipatt(com.meta[2:ncom+1], com.meta$REGION, control = how(nperm = numsam), duleg = F, func = "r.g")
summary(indval.region, indvalcomp = T)

# Extract results as a dataframe
indval.region_result = as.data.frame(indval.region$sign)

# Extract p-values from the results
indval.region_pvals <- indval.region$sign$p.value

# Adjust p-values using Benjamini-Hochberg (FDR) correction
indval.region_pvals_adj <- p.adjust(indval.region_pvals, method = "BH")

# Add adjusted p-values back to results
indval.region_result$p.adjusted <- indval.region_pvals_adj

indval.region_result=subset(indval.region_result, p.adjusted <0.05)

# Save full results and significant species to CSV files
write.csv(indval.region_result, file=file.path(export_dir, "2indvalunique_region_12S.csv"), row.names = T)
```

### iv. Method

```{r}
##BY METHOD
# Run indicator species analysis
indval.method = multipatt(com.meta[2:ncom+1], com.meta$FILTRATION, control = how(nperm = numsam), duleg = T, func="r.g")

# View summary
summary(indval.method)

# Extract results as a dataframe
indval.method_result = as.data.frame(indval.method$sign)

# Extract p-values from the results
indval.method_pvals <- indval.method$sign$p.value

# Adjust p-values using Benjamini-Hochberg (FDR) correction
indval.method_pvals_adj <- p.adjust(indval.method_pvals, method = "BH")

# Add adjusted p-values back to results
indval.method_result$p.adjusted <- indval.method_pvals_adj
indval.method_result=subset(indval.method_result, p.adjusted <0.05)

# Save full results and significant species to CSV files
write.csv(indval.method_result, file=file.path(export_dir, "2indvalunique_method_12S.csv"), row.names = T)
```

## B. Core taxa

### Within regions

```{r}
metatable.core=com.reads.meta

metatable.core <- metatable.core %>%
  dplyr::group_by(SITE) %>%
  dplyr::summarize(
    across(1:(ncom), sum, na.rm = TRUE), 
    across((ncom+1):ncol(com.reads.meta)-1, first)
  )

taxa_only <- metatable.core[, 2:(ncom+1)]

core_taxa <- names(taxa_only)[
  apply(taxa_only, 2, function(x) all(x > 0))
]

core_taxa #global core taxa

core_taxa_by_region <- metatable.core %>%
  group_by(REGION) %>%
  summarize(
    across(2:(ncom+1), ~ mean(.x > 0, na.rm = TRUE) >= core_threshold),
    .groups = "drop"
  )

# Extract only taxon column names (same as the global taxa_only dataframe)
taxa_cols <- colnames(taxa_only)

# Extract core taxa per region
core_taxa_list <- core_taxa_by_region %>%
  rowwise() %>%
  mutate(core_taxa = list(
    taxa_cols[unlist(c_across(all_of(taxa_cols)))]
  )) %>%
  select(REGION, core_taxa)

core_taxa_long <- core_taxa_list %>%
  rowwise() %>%   # treat each row separately
  mutate(core_taxa = list(core_taxa[core_taxa != "dropped"])) %>% # remove "dropped"
  unnest(core_taxa) %>%  # convert list column to long format
  ungroup() %>%           # ungroup for normal tibble operations
  mutate(Marker = "12S")

# Sum reads per taxon per region
core_taxa_long_reads <- core_taxa_long %>%
  left_join(
    # extract only the site, region, and taxa counts
    metatable.core %>%
      select(SITE, REGION, all_of(taxa_cols)) %>%
      pivot_longer(
        cols = all_of(taxa_cols),
        names_to = "core_taxa",
        values_to = "Reads"
      ),
    by = c("REGION", "core_taxa")
  ) %>%
  group_by(REGION, core_taxa, Marker) %>%
  summarize(
    Total_Reads = sum(Reads, na.rm = TRUE),
    .groups = "drop"
  )

write.csv(core_taxa_long_reads, file=file.path(export_dir, "2core_taxa_region_12S.csv"), row.names = FALSE)
```

### Within Methods

```{r}
metatable.core=com.reads.meta

metatable.core <- metatable.core %>%
  dplyr::group_by(SITE) %>%
  dplyr::summarize(
    across(1:(ncom), sum, na.rm = TRUE), 
    across((ncom+1):ncol(com.reads.meta)-1, first)
  )

taxa_only <- metatable.core[, 2:(ncom+1)]

core_taxa <- names(taxa_only)[
  apply(taxa_only, 2, function(x) all(x > 0))
]

core_taxa #global core taxa

core_taxa_by_method <- metatable.core %>%
  group_by(FILTRATION) %>%
  summarize(
    across(2:(ncom+1), ~ all(.x > 0)),
    .groups = "drop"
  )

# Extract only taxon column names (same as the global taxa_only dataframe)
taxa_cols <- colnames(taxa_only)

# Extract core taxa per region
core_taxa_list_method <- core_taxa_by_method %>%
  rowwise() %>%
  mutate(core_taxa = list(
    taxa_cols[unlist(c_across(all_of(taxa_cols)))]
  )) %>%
  select(FILTRATION, core_taxa)

core_taxa_long_method <- core_taxa_list_method %>%
  rowwise() %>%   # treat each row separately
  mutate(core_taxa = list(core_taxa[core_taxa != "dropped"])) %>% # remove "dropped"
  unnest(core_taxa) %>%  # convert list column to long format
  ungroup() %>%           # ungroup for normal tibble operations
  mutate(Marker = "12S")

# Sum reads per taxon per region
core_taxa_long_reads_method <- core_taxa_long_method %>%
  left_join(
    # extract only the site, region, and taxa counts
    metatable.core %>%
      select(SITE, FILTRATION, all_of(taxa_cols)) %>%
      pivot_longer(
        cols = all_of(taxa_cols),
        names_to = "core_taxa",
        values_to = "Reads"
      ),
    by = c("FILTRATION", "core_taxa")
  ) %>%
  group_by(FILTRATION, core_taxa, Marker) %>%
  summarize(
    Total_Reads = sum(Reads, na.rm = TRUE),
    .groups = "drop"
  )

write.csv(core_taxa_long_reads_method, file=file.path(export_dir, "2core_taxa_method_12S.csv"), row.names = FALSE)
```

# 5. Zeta Diversity

## A. Overall

### i. Zeta Decline

```{r}
library(zetadiv)

####ALL SITES####
metatable.z=metatable.reps
rownames(metatable.z)=metatable.z$SAMPLE
metatable.z <- metatable.z %>%
  dplyr::group_by(SITE) %>%
  dplyr::summarize(
    across(2:(ntax+1), sum, na.rm = TRUE), 
    across((ntax+2):ncol(metatable.reps)-1, first)
  )

table.z <- metatable.z[, -c(1, (ncol(metatable.z) - 12):ncol(metatable.z))] ##remove first two rows, last X rows of metadata

table.bin.z<-as.data.frame(table.z)
for (i in 1:nrow(table.bin.z)){
  for (j in 1:ncol(table.bin.z)){
    if(table.bin.z[i,j]>0){
      table.bin.z[i,j]<-1
    }
  }
}

xy=data.frame(X=metatable.z$LON)
xy$Y=metatable.z$LAT

#### Run model####################################
zeta.ex=Zeta.decline.ex(table.bin.z, orders = 1:nrow(table.bin.z), rescale=T, plot=F, sd.correct = T)
#zeta.ex=Zeta.decline.mc(table.bin.z, orders = 1:nrow(table.bin.z), sam=numperm, rescale=T, plot=F, sd.correct = T, silent = F, xy=xy, NON = T, DIR = F)
saveRDS(zeta.ex, file=file.path(export_dir, "2zetadecline_expected_12S.Rdata"))
#################################################
```

##### a. Expected Plots

```{r}
zeta.ex=readRDS(file=file.path(export_dir, "2zetadecline_expected_12S.Rdata"))

zeta.ex[1:4] %>% as.tibble() %>% mutate(Method = "12S")-> tib.ex

write.csv(tib.ex, file=file.path(export_dir, "2zeta_expected_decay_12S.csv"), row.names = F)

decay_all_ex <- tib.ex %>%
  ggplot(aes(x = zeta.order, y = zeta.val)) +
  geom_point(color = cols.marker[1]) + # Set point color
  geom_line(color = cols.marker[1]) + # Set line color
  geom_ribbon(
    aes(ymin = zeta.val - zeta.val.sd, ymax = zeta.val + zeta.val.sd), 
    fill = scales::alpha(cols.marker[1], 0.1), # Use less opaque version of cols.marker[1]
    color = NA # Remove border color
  ) +
  ylab("Zeta Diversity") +
  xlab("Zeta Order") +
  My_Theme
```


## B. By Region

### i. Zeta Decline

```{r}
####MAUNALUA####
metatable.mauna.z=subset(metatable.reps, REGION=="Maunalua")
rownames(metatable.mauna.z)=metatable.mauna.z$SAMPLE
metatable.mauna.z <- metatable.mauna.z %>%
  dplyr::group_by(SITE) %>%
  dplyr::summarize(
    across(2:(ntax+1), sum, na.rm = TRUE), 
    across((ntax+2):ncol(metatable.mauna.z)-1, first)
  )

table.mauna <- metatable.mauna.z[, -c(1, (ncol(metatable.mauna.z) - 12):ncol(metatable.mauna.z))]

table.bin.mauna<-as.data.frame(table.mauna)
for (i in 1:nrow(table.bin.mauna)){
  for (j in 1:ncol(table.bin.mauna)){
    if(table.bin.mauna[i,j]>0){
      table.bin.mauna[i,j]<-1
    }
  }
}

xy.mauna=data.frame(X=metatable.mauna.z$LON)
xy.mauna$Y=metatable.mauna.z$LAT

#### Run model#####################################
zeta.mauna.ex=Zeta.decline.ex(table.bin.mauna, orders = 1:nrow(table.bin.mauna), rescale=T, plot=F, sd.correct = T)
#zeta.mauna.ex=Zeta.decline.mc(table.bin.mauna, orders = 1:nrow(table.bin.mauna), sam=numperm, rescale=T, plot=F, sd.correct = T, silent = F, xy=xy.mauna, NON = T, DIR = F)
saveRDS(zeta.mauna.ex, file=file.path(export_dir, "2zetadecline_expected_mauna_12S.Rdata"))
###################################################


####KEEHI####
metatable.keehi.z=subset(metatable.reps, REGION=="Keehi")
rownames(metatable.keehi.z)=metatable.keehi.z$SAMPLE
metatable.keehi.z <- metatable.keehi.z %>%
  dplyr::group_by(SITE) %>%
  dplyr::summarize(
    across(2:(ntax+1), sum, na.rm = TRUE), 
    across((ntax+2):ncol(metatable.keehi.z)-1, first)
  )

table.keehi <- metatable.keehi.z[, -c(1, (ncol(metatable.keehi.z) - 12):ncol(metatable.keehi.z))] ##remove first two rows, last X rows of metadata

table.bin.keehi<-as.data.frame(table.keehi)
for (i in 1:nrow(table.bin.keehi)){
  for (j in 1:ncol(table.bin.keehi)){
    if(table.bin.keehi[i,j]>0){
      table.bin.keehi[i,j]<-1
    }
  }
}

xy.keehi=data.frame(X=metatable.keehi.z$LON)
xy.keehi$Y=metatable.keehi.z$LAT

#### Run model############################################
zeta.keehi.ex=Zeta.decline.ex(table.bin.keehi, orders = 1:nrow(table.bin.keehi), rescale=T, plot=F, sd.correct = T)
#zeta.keehi.ex=Zeta.decline.mc(table.bin.keehi, orders = 1:nrow(table.bin.keehi), sam=numperm, rescale=T, plot=F, sd.correct = T, silent = F, xy=xy.keehi, NON = T, DIR = F)
saveRDS(zeta.keehi.ex, file=file.path(export_dir, "2zetadecline_expected_keehi_12S.Rdata"))
#########################################################

zeta.mauna.ex$aic
zeta.keehi.ex$aic
```

##### a. Expected Plots

```{r}
zeta.mauna.ex=readRDS(file=file.path(export_dir, "2zetadecline_expected_mauna_12S.Rdata"))
zeta.keehi.ex=readRDS(file=file.path(export_dir, "2zetadecline_expected_keehi_12S.Rdata"))

zeta.mauna.ex[1:4] %>% as.tibble() %>% mutate(Region = "Maunalua")-> mauna_tib.ex
zeta.keehi.ex[1:4] %>% as.tibble() %>% mutate(Region = "Keehi")-> keehi_tib.ex

bind_rows(mauna_tib.ex, keehi_tib.ex) -> zeta_sites.ex
write.csv(zeta_sites.ex, file=file.path(export_dir, "2zeta_expected_region_12S.csv"), row.names = F)

decay_samples.ex <- zeta_sites.ex %>% 
ggplot(aes(x= zeta.order, y= zeta.val, group=Region,color=Region, fill=Region)) +geom_point() +geom_line() + geom_ribbon(aes(ymin = zeta.val - zeta.val.sd, ymax = zeta.val + zeta.val.sd),alpha =0.1, colour = NA)  + scale_fill_manual(values=cols.region2) +  scale_colour_manual(values=cols.region2) +ylab("Zeta Diversity") + xlab("Zeta Order") + My_Theme + scale_x_continuous(breaks = function(zeta.order) seq(floor(min(zeta.order)), ceiling(max(zeta.order)), by = 1))


zeta.mauna.ex[5] %>% as.tibble() %>% mutate(Region = "Maunalua", zeta.order = 1:n())-> mauna_tib.ratio.ex
zeta.keehi.ex[5] %>% as.tibble() %>% mutate(Region = "Keehi", zeta.order = 1:n())-> keehi_tib.ratio.ex

bind_rows(mauna_tib.ratio.ex, keehi_tib.ratio.ex) -> zeta_sites.ratio.ex
write.csv(zeta_sites.ratio.ex, file=file.path(export_dir, "2zeta_ratio_region_12S.csv"), row.names = F)

retention_samples.ratio.ex <- zeta_sites.ratio.ex %>% 
ggplot(aes(x= zeta.order, y= ratio, group=Region,color=Region, fill=Region)) +geom_point() +geom_line()  + scale_fill_manual(values=cols.region2) +  scale_colour_manual(values=cols.region2) +ylab("Zeta Ratio") + xlab("Zeta Order") + ylim(0,1) + My_Theme+ scale_x_continuous(breaks = function(zeta.order) seq(floor(min(zeta.order)), ceiling(max(zeta.order)), by = 1))
```


```{r}
zetadeclineplot=ggarrange(
  decay_samples.ex, retention_samples.ratio.ex, 
  ncol = 2,
  labels="AUTO",  common.legend = T, 
  legend = "top", 
  align = "hv"
)

zetadeclineplot
```


## B. By Method

### i. Zeta Decline

```{r}
####ACTIVE####
metatable.active.z=subset(metatable.reps, FILTRATION=="Active")
rownames(metatable.active.z)=metatable.active.z$SAMPLE
metatable.active.z <- metatable.active.z %>%
  dplyr::group_by(SITE) %>%
  dplyr::summarize(
    across(2:(ntax+1), sum, na.rm = TRUE), 
    across((ntax+2):ncol(metatable.active.z)-1, first)
  )

table.active <- metatable.active.z[, -c(1, (ncol(metatable.active.z) - 12):ncol(metatable.active.z))]

table.bin.active<-as.data.frame(table.active)
for (i in 1:nrow(table.bin.active)){
  for (j in 1:ncol(table.bin.active)){
    if(table.bin.active[i,j]>0){
      table.bin.active[i,j]<-1
    }
  }
}

xy.active=data.frame(X=metatable.active.z$LON)
xy.active$Y=metatable.active.z$LAT

#### Run model#####################################
zeta.active.ex=Zeta.decline.ex(table.bin.active, orders = 1:nrow(table.bin.active), rescale=T, plot=F, sd.correct = T)
#zeta.active.ex=Zeta.decline.mc(table.bin.active, orders = 1:nrow(table.bin.active), sam=numperm, rescale=T, plot=F, sd.correct = T, silent = F, xy=xy.active, NON = T, DIR = F)
saveRDS(zeta.active.ex, file=file.path(export_dir, "2zetadecline_expected_active_12S.Rdata"))
###################################################


####passive####
metatable.passive.z=subset(metatable.reps, FILTRATION=="Passive")
rownames(metatable.passive.z)=metatable.passive.z$SAMPLE
metatable.passive.z <- metatable.passive.z %>%
  dplyr::group_by(SITE) %>%
  dplyr::summarize(
    across(2:(ntax+1), sum, na.rm = TRUE), 
    across((ntax+2):ncol(metatable.passive.z)-1, first)
  )

table.passive <- metatable.passive.z[, -c(1, (ncol(metatable.passive.z) - 12):ncol(metatable.passive.z))] ##remove first two rows, last X rows of metadata

table.bin.passive<-as.data.frame(table.passive)
for (i in 1:nrow(table.bin.passive)){
  for (j in 1:ncol(table.bin.passive)){
    if(table.bin.passive[i,j]>0){
      table.bin.passive[i,j]<-1
    }
  }
}

xy.passive=data.frame(X=metatable.passive.z$LON)
xy.passive$Y=metatable.passive.z$LAT

#### Run model############################################
zeta.passive.ex=Zeta.decline.ex(table.bin.passive, orders = 1:nrow(table.bin.passive),  rescale=T, plot=F, sd.correct = T)
#zeta.passive.ex=Zeta.decline.mc(table.bin.passive, orders = 1:nrow(table.bin.passive), sam=numperm, rescale=T, plot=F, sd.correct = T, silent = F, xy=xy.passive, NON = T, DIR = F)
saveRDS(zeta.passive.ex, file=file.path(export_dir, "2zetadecline_expected_passive_12S.Rdata"))
#########################################################

zeta.active.ex$aic
zeta.passive.ex$aic
```

##### a. Expected Plots

```{r}
zeta.active.ex=readRDS(file=file.path(export_dir, "2zetadecline_expected_active_12S.Rdata"))
zeta.passive.ex=readRDS(file=file.path(export_dir, "2zetadecline_expected_passive_12S.Rdata"))

zeta.active.ex[1:4] %>% as.tibble() %>% mutate(Method = "Active")-> active_tib.ex
zeta.passive.ex[1:4] %>% as.tibble() %>% mutate(Method = "Passive")-> passive_tib.ex

bind_rows(active_tib.ex, passive_tib.ex) -> zeta_sites.ex
write.csv(zeta_sites.ex, file=file.path(export_dir, "2zeta_expected_method_12S.csv"), row.names = F)

decay_samples.ex <- zeta_sites.ex %>% 
ggplot(aes(x= zeta.order, y= zeta.val, group=Method,color=Method, fill=Method)) +geom_point() +geom_line() + geom_ribbon(aes(ymin = zeta.val - zeta.val.sd, ymax = zeta.val + zeta.val.sd),alpha =0.1, colour = NA)  + scale_fill_manual(values=cols.method) +  scale_colour_manual(values=cols.method) +ylab("Zeta Diversity") + xlab("")  + xlim(0,16) + My_Theme + scale_x_continuous(breaks = function(zeta.order) seq(floor(min(zeta.order)), ceiling(max(zeta.order)), by = 1))


zeta.active.ex[5] %>% as.tibble() %>% mutate(Method = "Active", zeta.order = 1:n())-> active_tib.ratio.ex
zeta.passive.ex[5] %>% as.tibble() %>% mutate(Method = "Passive", zeta.order = 1:n())-> passive_tib.ratio.ex

bind_rows(active_tib.ratio.ex, passive_tib.ratio.ex) -> zeta_sites.ratio.ex
write.csv(zeta_sites.ratio.ex, file=file.path(export_dir, "2zeta_ratio_method_12S.csv"), row.names = F)

retention_samples.ratio.ex <- zeta_sites.ratio.ex %>% 
ggplot(aes(x= zeta.order, y= ratio, group=Method,color=Method, fill=Method)) +geom_point() +geom_line()  + scale_fill_manual(values=cols.method) +  scale_colour_manual(values=cols.method) +ylab("Zeta Ratio") + xlab("") + xlim(0,19) + ylim(0,1) + My_Theme+ scale_x_continuous(breaks = function(zeta.order) seq(floor(min(zeta.order)), ceiling(max(zeta.order)), by = 1))
```


```{r}
zetadeclineplot=ggarrange(
  decay_samples.ex, retention_samples.ratio.ex, 
  ncol = 2, 
  labels = "AUTO", 
  common.legend = T, 
  legend = "top", 
  align = "hv"
)

zetadeclineplot
```