---
title: "eDNA Course - Diversity analysis"
author: "Patrick Nichols"
date: "2026-07-02"
output: html_document
---

## 1. Setup

Session -> Set working directory -> To source file location...

Carefully update the user-defined settings for each run, making sure column names and index positions match.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(iNEXT)
library(vegan)
library(ggpubr)
library(indicspecies)
library(microDecon)
library(purrr)
library(tibble)
library(readxl)

My_Theme = theme_set(theme_bw()+theme(
  panel.grid.minor = element_blank(),
  panel.background = element_blank(), 
  panel.border = element_rect(fill=NA, colour = "black"), 
  panel.spacing = unit(1, "lines"),
  axis.title.x = element_text(size = 12, color="black"),
  axis.text.x = element_text(size = 12, color="black"),
  axis.title.y = element_text(size = 12, color = "black"),
  axis.text.y = element_text(size=12, color = "black"),
  legend.text = element_text(size=12, color = "black"),
  strip.text.x = element_text(size=14, color="black"),
  strip.text.y = element_text(size=14, color="black"),
  strip.text = element_text(size = 14, color = "black"),
  strip.background = element_blank(), 
  strip.placement = "outside",
  title = element_text(size=14, color="black", face="bold"),
  axis.title = element_text(size=12, face="bold", colour = "black"),
  axis.text = element_blank()))

set.seed(13)
numperm=9999 #Permutation tests

paths=list(
  data_raw      ="data/raw",
  data_processed="data/processed",
  results_rds   ="results/rds",
  results_tables="results/tables",
  results_figs  ="results/figures"
)

walk(paths, dir.create, showWarnings = FALSE, recursive = TRUE)

cols.method=c("#031d44", "#1a659e", "#d5896f", "#dab785", "#70a288", "snow3", "black")

#### RUN SETTINGS #######################################################
num.blanks=27 # how many columns of blanks/negative control samples?
SAMPLE="Sample_ID" # name of metadata column with sample IDs
OTU_ID="OTU_ID" # name of OTU ID column in OTU table
GROUPING="TREATMENT" # name of grouping variable column in metadata
TAXONOMICRANK= "Genus" # choose taxonomic rank for indicator taxon analysis
distance.method="bray" #choose distance method for community dissimilarity ordination
#########################################################################

#### READ IN OTU TABLE ##################################################
otutab.raw=read_excel(file.path(paths$data_raw, "combined_data_read_table.xlsx"))  # raw OTU table, with Unique ID/OTU_ID as first column, apscale taxonomic information
otutab.predecon=otutab.raw[,c(1:257)] # select only OTU ID and numerical read count columns
otutab.predecon[, -1] <- lapply(otutab.predecon[, -1], function(x) {
  if(is.character(x)) as.numeric(trimws(x)) else x
})
# Remove zero-sum rows (keep ID column)
otutab.predecon <- otutab.predecon[rowSums(otutab.predecon[, -1]) != 0, ]
# Remove zero-sum columns (keep ID column)
otutab.predecon <- otutab.predecon[, c(1, which(colSums(otutab.predecon[, -1]) != 0) + 1)]
apscale.info=otutab.raw[,c(1,258:ncol(otutab.raw))] # select taxonomic info and other columns from apscale
#########################################################################

#### READ IN METADATA FILE #############################################
metadata=read_excel(file.path(paths$data_raw, "metadata.xlsx"))
metadata=subset(metadata, Run != "1")
metadata=subset(metadata, SITE=="Station2" | SITE=="Station3")
metadata=subset(metadata, SITE != "Control")
metadata=subset(metadata, Marker=="Teleo2")
metadata <- subset(metadata, !(Sample_ID %in% c("Teleo2_run2_barcode81", 
                                                "Teleo2_run4_barcode65", 
                                                "Teleo2_run4_barcode68")))
metadata$QUBIT <- as.numeric(gsub(",", ".", trimws(metadata$QUBIT)))
#######################################################################
```

### Decontamination

#### MicroDecon

```{r}
otutab.predecon <- as.data.frame(otutab.predecon)

# Run decontamination
microdecon.decontaminated <- decon(otutab.predecon,numb.blanks = num.blanks, numb.ind = (ncol(otutab.predecon)-num.blanks-1),taxa = FALSE,thresh = 1,prop.thresh = 0, runs = 2)

reads.removed=microdecon.decontaminated$reads.removed
decontaminated.table <- microdecon.decontaminated$decon.table
decontaminated.table$Mean.blank <- NULL

reads.removed=microdecon.decontaminated$reads.removed
reads.removed=merge(apscale.info, reads.removed, by=OTU_ID)

decontaminated.table.metadata=merge(decontaminated.table, apscale.info, by=OTU_ID) # merge taxonomic info back into decontaminated file

write.csv(decontaminated.table.metadata, file=file.path(paths$data_processed, "decontaminated_data_Teleo2.csv"), row.names = F) ## save to PC
```


### Data cleaning and prepping

This block removes human-related taxa (genus Homo) from the dataset and renames the rows using organism IDs. It then isolates the taxonomic classification columns into a separate table called 'taxa'.

```{r}
otutab=decontaminated.table.metadata

# Filter unwanted taxa
data=subset(otutab, Order !="Primates") # if taxonomy information is present in OTU table
data=subset(data, Kingdom !="NoMatch") # if taxonomy information is present in OTU table
data=subset(data, Genus !="Cranioleuca") # if taxonomy information is present in OTU table


row.names(data)=data[[OTU_ID]]
final=data[,2:230] # Update me!!
rownames(final)=data[["OTU_ID"]]

final.t=as.data.frame(t(final))
final.t[[SAMPLE]]=rownames(final.t)
final.t.meta=merge(final.t, metadata, by=SAMPLE)

final.t=final.t.meta[,2:(nrow(data)+1)]
rownames(final.t)=final.t.meta[[SAMPLE]]

final=as.data.frame(t(final.t))
final <- final[, colSums(final) != 0] # remove zero sum columns
final <- final[rowSums(final) != 0, ] # remove zero sum rows
#final <- final[, colSums(final != 0) > 1] # remove samples with only 1 OTU
ntax=nrow(final)
nsample=ncol(final)
final[[OTU_ID]]=rownames(final)

final=merge(final, apscale.info, by=OTU_ID)
rownames(final)=final[[OTU_ID]]

# Taxonomic metadata only
taxa=final[, (nsample+2):ncol(final)]
rownames(taxa)=final[[OTU_ID]]
taxa[[OTU_ID]]=rownames(taxa)
```


### Transformations

This code prepares three versions of your OTU table (raw counts, Hellinger-transformed, and eDNA index), then lets you choose one for downstream analysis. The final steps transpose the table so taxa become columns and samples become rows, then merges it with site metadata for statistical testing.

#### Raw Reads (no transformation)

```{r}
# Raw reads OTU table only, no metadata
table.raw=final[, 2:(nsample+1)]
final_reads=final # make a copy for raw reads (will not undergo transformations)
table.reads=final_reads[,2:(nsample+1)]

sum(table.reads)

table.reads.t <- as.data.frame(t(table.reads))  # Flip OTU table so taxa as columns

table.reads.t[[SAMPLE]] <- row.names(table.reads.t) # add sample column to merge with metadata
metatable.reps.reads=merge(table.reads.t, metadata, by=SAMPLE) ##OTU table with site metadata correctly formatted for downstream analyses

table.reads.t[[SAMPLE]]=NULL #remove non-numeric columns for downstream analyses
```

#### Presence-Absence (Binary) 

```{r}
table_pa=decostand(table.reads, method = "pa")
table_pa[[OTU_ID]]=rownames(table_pa)
final_pa=merge(table_pa, taxa, by=OTU_ID)
rownames(final_pa)=final_pa[[OTU_ID]]
table_pa[[OTU_ID]]=NULL
table_pa.t=t(table_pa)
```

#### Hellinger 

```{r}
table_hel=decostand(table.reads, method = "hellinger")
table_hel[[OTU_ID]]=rownames(table_hel)
final_hel=merge(table_hel, taxa, by=OTU_ID)
rownames(final_hel)=final_hel[[OTU_ID]]
table_hel[[OTU_ID]]=NULL
table_hel.t=t(table_hel)
```

#### eDNA Index

```{r}
table.raw <- as.matrix(table.reads)  # Compute relative abundance per sample (column-wise)
relative_abundance <- sweep(table.raw, 2, colSums(table.raw), "/")  # Find the maximum observed abundance for each OTU across all samples (row-wise)
max_relative_abundance <- apply(relative_abundance, 1, max, na.rm = TRUE)  # Compute the eDNA index by dividing each OTU's abundance by its max
table_index <- as.data.frame(sweep(relative_abundance, 1, max_relative_abundance, "/"))  # Add OTU IDs back to the table

table_index[[OTU_ID]] <- rownames(table_index) # Merge with taxa metadata (assuming 'taxa' has OTU_ID as key)
final_index <- merge(table_index, taxa, by = OTU_ID) # Ensure row names match OTU_ID
rownames(final_index) <- final_index[[OTU_ID]] 
table_index[[OTU_ID]] <- NULL
table_index.t=t(table_index)
```

#### Prepare community data

```{r}
####CHOOSE MATRIX###################################################
community.table.t=table_index.t #transposed otu table version (OPTIONS: table_index.t, table_hel.t, table.reads.t, table_pa.t)
community.table=table_index #otu table only (OPTIONS: table_index, table_hel, table.reads, table_pa, etc.)
community.data=final_index #version with metadata (OPTIONS: final_index, final_hel, final_reads, final_pa, etc.)
###################################################################
```

#### Merge filtered metadata

```{r}
table.t <- as.data.frame(t(as.matrix(community.table)))

ntax=ncol(table.t)
nsample=nrow(table.t)

table.t[[SAMPLE]]=row.names(table.t) ##OTU table only

# Remove leading zeros from SAMPLE column
metatable.reps=merge(table.t, metadata, by=SAMPLE) ##OTU table with site metadata
table.t[[SAMPLE]]=NULL
```


####Melt data long format

```{r}
data_long <- as.data.frame(community.table.t) %>%
  rownames_to_column(var = SAMPLE) %>%
  pivot_longer(
    cols = -SAMPLE,  # Keep SampleID as identifier
    names_to = OTU_ID,
    values_to = "abundance"
  )

# Remove zero abundances (optional)
data_long <- data_long %>%
  filter(abundance > 0)

data_long=merge(data_long, metadata, by=SAMPLE)
data_long=merge(data_long, apscale.info, by=OTU_ID)
```

## 3. Community dissimilarity

This code calculates community dissimilarity using a chosen distance method (e.g., Bray-Curtis), then runs non-metric multidimensional scaling (NMDS) to visualize how samples cluster in 2D space, with ellipses showing 95% confidence intervals for each group. It then uses envfit to identify which environmental or metadata variables are significantly correlated with the community patterns (p < 0.05), adding them as arrows to the NMDS plot to show direction and strength of influence.

### MDS

```{r}
table.dist<- vegdist(community.table.t, method=distance.method, binary=F) # distance matrix from community OTU table

metadata$group_factor <- as.factor(metadata[[GROUPING]]) # grouping variable for ordination

set.seed(13) 

NMDSdat<-metaMDS(table.dist, k=2, weakties=TRUE, maxit=1000, pc=TRUE, zerodist="add", autotransform = F) # ordination function, with k dimensions 

stress_value <- round(NMDSdat$stress, 3) #extract 2D stress value (watch stress, should be < 0.15, lower the better)

PDat <- data.frame(
  NMDS1 = NMDSdat$points[, 1],
  NMDS2 = NMDSdat$points[, 2],
  row.names(table.t)) # creates dataframe of ordination axes results

names(PDat)[3] <- SAMPLE # add in sample information
PDat<-merge(PDat, metadata, by=SAMPLE) # merge with metadata

write.csv(PDat, file=file.path(paths$data_processed, "NMDS_data_Teleo2.csv"), row.names = F)

nmds.plot <- qplot(x = NMDS1, y = NMDS2, 
                   colour = group_factor,
                   alpha = I(0.6), 
                   data = PDat) +  
  geom_vline(xintercept = 0, lty = 2, col = "grey66") + 
  geom_hline(yintercept = 0, lty = 2, col = "grey66") + 
  stat_ellipse(aes(fill = group_factor),
               level = 0.95, 
               geom = "polygon", 
               alpha = 0.15,
               colour = NA) +  # Remove outline
  My_Theme +  
  labs(colour = NULL, 
       fill = NULL, 
       subtitle = paste("Teleo2; Stress =", stress_value)) + 
  scale_color_manual(values = cols.method) +
  scale_fill_manual(values = cols.method)
nmds.plot

ggsave(
  filename=file.path(paths$results_figs,paste0("NMDSplot_Teleo2", GROUPING,".pdf")),
  plot    =nmds.plot,
  width   =6,
  height  =4,
  units   ="in",
  device  =cairo_pdf,
  family  ="Arial"
)
```


### Dispersion test

Tests for homogeneity of dispersion, The p-value is greater than 0.05, indicating that the homogeneity of dispersion across the groups is not significantly different. In other words, the variation within the groups (dispersion) is similar, and there is no strong evidence to suggest that one group is more or less dispersed than the others.The lack of significance in betadisper indicates that the spread of points (dispersion) within STEP groups is not significantly different, meaning that the differences in the dissimilarities found in adonis2 are more likely due to differences in group centroids rather than differences in variability within the groups.

```{r}
# Check for variability across SITE groups
dispersion <- betadisper(table.dist, group = metatable.reps[[GROUPING]])
anova(dispersion) # if significant -> ANOSIM suggested over PERMANOVA
```

### Centroid tests

This code runs two statistical tests to check if your groups (e.g., sites or treatments) have significantly different microbial communities. ANOSIM tests whether differences between groups are larger than differences within groups (R close to 1 = strong separation), while PERMANOVA tests if group membership explains a significant portion of the variation in community composition (p-value < 0.05 indicates significant differences).

#### ANOSIM

```{r}
anosim.mod=anosim(table.dist, group = metatable.reps[[GROUPING]], permutations = numperm)
anosim.mod
```

#### PERMANOVA

```{r}
permanova.mod=adonis2(table.dist ~ TREATMENT, data = metatable.reps, permutations = numperm)
permanova.mod 
```

## 6. Indicator taxa
This code aggregates OTUs by a higher taxonomic rank (e.g., genus or family), then runs indicator species analysis (IndVal) to identify which taxa are significantly associated with specific groups (e.g., sites or treatments). It filters significant indicators (adjusted p < 0.15), then creates a faceted time-series plot showing how the abundance of each indicator taxon changes over time, with separate panels for each group and each indicator taxon, using violin-boxplots and a dashed line connecting medians.

### Aggregate OTU table

```{r}
group=as.factor(metatable.reps[[GROUPING]])
taxonomic.rank=community.data[[TAXONOMICRANK]] 

com=community.table.t

colnames(com)=taxonomic.rank

prefixes = unique(sub("\\..*", "", colnames(com)))
prefixes = prefixes[prefixes != ""]

com=sapply(prefixes, function(x) rowSums(com[,startsWith(colnames(com), x), drop=F])) ##OTU table with combined taxonomies (no repeated columns)

com=as.data.frame(com)

ncom=ncol(com)
```

### IndVal analysis

```{r}
##BY SITE
# Run indicator species analysis
indval.site = multipatt(com, group, control = how(nperm = numperm), duleg = T, func="IndVal.g")

# View summary
summary(indval.site)

# Extract results as a dataframe
summary_result = as.data.frame(indval.site$sign)

# Extract p-values from the results
pvals <- indval.site$sign$p.value

# Adjust p-values using Benjamini-Hochberg (FDR) correction
pvals_adj <- p.adjust(pvals, method = "BH")

# Add adjusted p-values back to results
summary_result$p.adjusted <- pvals_adj

# Filter significant species after correction (adjusted p-value < 0.05)
significant_species <- summary_result[pvals < 0.05, ]

# Get the site/group that each indicator taxa is associated with
# For each row in significant_species, find which "s." column has value 1
indicator_groups <- apply(significant_species[, grep("^s\\.", names(significant_species))], 1, function(x) {
  # Find column names where value is 1
  sites <- names(x)[x == 1]
  # Remove the "s." prefix
  sites <- gsub("^s\\.", "", sites)
  # If multiple sites, combine with comma, otherwise return single site
  if(length(sites) > 1) {
    return(paste(sites, collapse = ", "))
  } else if(length(sites) == 1) {
    return(sites)
  } else {
    return(NA)
  }
})

# Add this information to significant_species
significant_species$indicator_group <- indicator_groups
significant_species$indicator_taxon <- rownames(significant_species)

significant_species
```

## 7. Heatmap

```{r}
# ============================================
# HEATMAP - LOG-TRANSFORMED SUMMED READS BY TREATMENT
# ============================================

library(pheatmap)

# Match samples
common_samples <- intersect(rownames(com), metadata[[SAMPLE]])
com_subset <- com[common_samples, ]
metadata_subset <- metadata[match(common_samples, metadata[[SAMPLE]]), ]

# Sum reads by treatment
sum_by_treatment <- aggregate(com_subset, 
                              by = list(Treatment = metadata_subset$TREATMENT), 
                              FUN = sum)

# Set treatment as rownames and remove the Treatment column
rownames(sum_by_treatment) <- sum_by_treatment$Treatment
sum_by_treatment <- sum_by_treatment[, -1]

# Transpose so taxa are rows, treatments are columns
heatmap_data <- t(sum_by_treatment)

# Optional: Remove taxa with zero reads across all treatments
#heatmap_data <- heatmap_data[rowSums(heatmap_data) > 0, ]

# Log transform (log1p = log(1+x) to handle zeros)
heatmap_data_log <- log1p(heatmap_data)

write.csv(heatmap_data_log, file=file.path(paths$data_processed, "heatmap_data_Teleo2.csv"), row.names = F)

# Create heatmap (no scaling)
heatmap=pheatmap(heatmap_data_log,
         scale = "none", fontsize_row = 8,
         fontsize_col = 10,
         angle_col = 45,
         cellwidth = 30,
         cellheight = 10,
         show_rownames = TRUE,
         show_colnames = TRUE,
         cluster_rows = T,
         cluster_cols = F,
         color = colorRampPalette(c("white", "tan", "firebrick3"))(100))
heatmap
```